---
title: "Student_Data"
author: "Lorraine Gaudio"
date: "2025-08-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this document, I wrangle student data. 


# Roster


I need the emailing roster so that I can email students the pre- and post- survey

```{r load-data, message=FALSE, warning=FALSE}
# ---- Step 1: Load Excel (sheets "900" and "1030") --------------------------
library(readxl)
library(dplyr)
library(janitor)

# Path to the roster file
roster_path <- "data/Class_Roster.xlsx"

# Sanity check
if (!file.exists(roster_path)) {
  stop("File not found: data/Class_Roster.xlsx (check working directory).")
}

# List sheets
sheets_available <- excel_sheets(roster_path)
required_sheets  <- c("900", "1030")
missing_sheets   <- setdiff(required_sheets, sheets_available)
if (length(missing_sheets)) {
  stop("Missing sheet(s): ", paste(missing_sheets, collapse = ", "),
       ". Found sheets: ", paste(sheets_available, collapse = ", "))
}

# Helper function: read, clean, drop NA course_id
read_roster_sheet <- function(path, sheet_name) {
  read_excel(path, sheet = sheet_name, col_types = "text") |>
    clean_names() |>
    filter(!is.na(course_id)) |>   # remove NA course_id
    mutate(section = sheet_name, .before = 1)
}

# Keep each section separate
roster_900  <- read_roster_sheet(roster_path, "900")
roster_1030 <- read_roster_sheet(roster_path, "1030")

# Quick peek
glimpse(roster_900)
glimpse(roster_1030)
```


I need to create several new columns in both datasets. 1. We need to create a random nine-digit identification number for each row called "ExternalDataReference" 2. We need to create a column "FirstName" and "LastName" from the column name. The column name, in both datasets, is formatted like this: "LastName,FirstName". It is comma-seperated with no space, and the last name is first.


```{r add-qualtrics-columns, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(stringr)

# Helper to generate unique 9-digit IDs (as zero-padded text)
make_ids <- function(n) {
  formatC(sample(0:999999999, n, replace = FALSE), width = 9, format = "d", flag = "0")
}

transform_roster <- function(df) {
  df %>%
    mutate(
      # 1) Unique 9-digit ExternalDataReference (text, preserves leading zeros)
      ExternalDataReference = make_ids(n())
    ) %>%
    mutate(name = as.character(name)) %>%
    # 2) Split "LastName,FirstName" -> LastName, FirstName
    separate(name, into = c("LastName", "FirstName"), sep = ",", fill = "right", remove = FALSE) %>%
    mutate(
      FirstName = str_trim(coalesce(FirstName, "")),
      LastName  = str_trim(coalesce(LastName, ""))
    )
}

# Apply to both datasets
roster_900  <- transform_roster(roster_900)
roster_1030 <- transform_roster(roster_1030)

# Quick check
roster_900  %>% select(section, name, FirstName, LastName, ExternalDataReference) %>% head()
roster_1030 %>% select(section, name, FirstName, LastName, ExternalDataReference) %>% head()
```
I rename email and "Email"and select columns c(FirstName, LastName, Email, ExternalDataReference)
Then I store both datasets in the data folder so I can upload them into Qualtrics.

```{r prep-for-qualtrics, message=FALSE, warning=FALSE}
library(readr)

# Helper to finalize Qualtrics format
finalize_roster <- function(df) {
  df %>%
    rename(Email = email) %>%
    select(FirstName, LastName, Email, ExternalDataReference)
}

# Apply to each section
roster_900_final  <- finalize_roster(roster_900)
roster_1030_final <- finalize_roster(roster_1030)

# Save to data folder as CSV
write_csv(roster_900_final,  "data/roster_900.csv")
write_csv(roster_1030_final, "data/roster_1030.csv")

# Quick peek
head(roster_900_final)
head(roster_1030_final)
```



# Pre-survey

I downloaded and manually cleaned the pre-survey data. I changed the column names to be consistent with the previous survey I did in the summer. 

```{r}
library(readxl)
pre_survey <- read_excel("data/pre_survey_manual_edit.xlsx")
available_fields <- read_excel("data/pre_survey_manual_edit.xlsx", 
    sheet = "available_fields")
```

```{r}
colnames(pre_survey)
```

## Instructor Tips

The first priority is to determine what tips do students have for me. I will read these tips, add important notes to the Class_Roster online and schedule check ins. 

```{r}
# Base R
# 1) Keep only the needed columns; rename `First _Name` -> First_Name
tips <- pre_survey[, c("Start_Date", "Last_Name", "First _Name", "instructor_tips")]
names(tips)[names(tips) == "First _Name"] <- "First_Name"

# 2) Trim whitespace in character columns (so "   " becomes "")
tips$instructor_tips <- trimws(tips$instructor_tips)

# 3) Keep rows where instructor_tips is not NA and not empty
tips <- tips[!is.na(tips$instructor_tips) & tips$instructor_tips != "", ]

# 4) Print all rows
print(tips)
```


```{r}
# dplyr
library(dplyr)

# 1) Keep only the needed columns; rename `First _Name` -> First_Name
tips <- pre_survey %>%
  select(Start_Date, Last_Name, First_Name = `First _Name`, instructor_tips)

# 2) Trim whitespace in character columns (so "   " becomes "")
tips <- tips %>%
  mutate(across(where(is.character), trimws))

# 3) Keep rows where instructor_tips is not NA and not empty
tips <- tips %>%
  filter(!is.na(instructor_tips) & instructor_tips != "")

# 4) Print all rows
print(tips, n = Inf)
```


## Low Skill Managing Files

The next thing we need to do is determine who scored two or below on familiarity_managing_files?

```{r}
str(pre_survey$familiarity_managing_files)
```


```{r}
# Base R

# 1) Keep only the columns we need; fix the space in First _Name
low_files <- pre_survey[, c("Last_Name", "First _Name", "Email_address", "familiarity_managing_files")]
names(low_files)[names(low_files) == "First _Name"] <- "First_Name"

# 2) Tidy text fields (trim spaces) and standardize email to lowercase
low_files$Last_Name <- trimws(low_files$Last_Name)
low_files$First_Name <- trimws(low_files$First_Name)
low_files$Email_address <- tolower(trimws(low_files$Email_address))

# 3) Recode NA values to 0
low_files$familiarity_managing_files[is.na(low_files$familiarity_managing_files)] <- 0

# 4) Filter to scores of 2 or below
low_files <- low_files[low_files$familiarity_managing_files <= 2, ]

# 5) Arrange for easy reading
low_files <- low_files[order(low_files$Last_Name, low_files$First_Name), ]

# 6) Print the table
print(low_files)
```


```{r}
library(dplyr)

low_files <- pre_survey %>%
  # 1) Keep only the columns we need; fix the space in First _Name
  select(Last_Name, First_Name = `First _Name`, Email_address, familiarity_managing_files) %>%
  
  # 2) Tidy text fields (trim spaces) and standardize email to lowercase
  mutate(
    across(c(Last_Name, First_Name, Email_address), trimws),
    Email_address = tolower(Email_address),
    
    # 3) Recode NA values to 0
    familiarity_managing_files = ifelse(is.na(familiarity_managing_files), 0, familiarity_managing_files)
  ) %>%
  
  # 4) Filter to scores of 2 or below
  filter(familiarity_managing_files <= 2) %>%
  
  # 5) Arrange for easy reading
  arrange(Last_Name, First_Name)

# 6) Print the table
print(low_files, n = Inf)
```



