---
title: "Chapter 13: Descriptive Statistics"
subtitle: "Group by, Summarize, and Count"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Chapter 13: Descriptive Statistics"
subtitle: "Group by, Summarize, and Count"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
# ---
title: "Chapter 13: Descriptive Statistics"
subtitle: "Group by, Summarize, and Count"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```
# Overview

In Chapter 11 and 12, we explored several functions from `dplyr` (`select()`, `filter()`, `mutate()`, and `case_when()`). This week, we will look at other functions in the `dplyr` package (`group_by()`, `summarize()` and `count()`) that will speed your ability to compare sub-groups and calculate descriptive statistics.

By the end of Chapter 13 you will be able to:

1. ⚡ Remember – List the roles of `group_by()` `count()` and `summarize()`.

2. 🔍 Understand – Explain how grouping changes the behavior of summary functions.

3. 🛠 Apply –  Compute mean, standard deviation, and count for each group in one call.

4. 📊 Analyze – Combine multiple grouping variables.

5. 🧠️ Evaluate – Decide when to overwrite vs. keep original variables.

Keep these goals in mind as you move through each section.

# Setting Up Your Notes

In Chapter 11, you learned how to create a Quarto file, keep the YAML header, use Markdown headings, write memos outside code chunks, and control chunks with #| options. In Chapter 12, you add a table of contents and section numbering, and use inline R code inside your sentences. In Chapter 13, you will practice using these skills. 

## Create a Quarto File (🌱)

For those of you who **have GitHub** and a course project, navigate to your course folder and open your course project. This will open RStudio. Click "Pull" in the Git pane to get the latest version of the course materials. After completing work, save, stage, commit, and push. 

If you **don't have GitHub** or a course project, just open RStudio and set your working directory to your course folder.

Open a new Quarto document in your course folder and save it as: `chapter13_notes.qmd` (`File > New File > Quarto Document`).

## YAML Header (🌱)

Copy and paste the YAML header from this file into your new Quarto document. Update the title, author, date, and team fields in the YAML header.

```{r echo=FALSE, results='asis'}
cat("```yaml
---
title: \"Chapter 13 Notes\"
subtitle: \"group_by() and summarize()\"
author: \"YOUR NAME\"
date: today
toc: true
toc-depth: 2
number-sections: true
format:
  html:
    code-fold: true
execute:
  warning: true
  message: true
---
```")
```

## Headings (🌱)

Use the section headings in this file to create the structure of your notes. You can copy and paste the section headings into your new Quarto document.

```markdown

# Loading Packages

# Discriptive Statistics

## Addition

## Mode

## Mean

## Variance

## Standard Deviation

## Median

## Range

### Interquartile Range (IQR)

### Percentiles

# summarize()

# group_by()

# multiple statistics

# multiple variables

```

Work through the sections under each heading. Render to HTML often to see the results.

## 📦 Packages

Install once (if needed): `install.packages("dplyr")`; `install.packages("dslabs")`

🎯 Load the packages 

```{r message=FALSE, warning=FALSE}
# ⚡
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
```

# 📖 Foundations

Before you use `summarize()` and `group_by()`, you need to understand what your data represents, what kind of variable you are working with, and what a summary value actually means. These opening sections build that foundation. They introduce key terms and descriptive statistics so you can make informed choices later in the chapter. 

As you work through the foundations, focus on matching the kind of data we have to the kind of summary that makes sense. That preparation will help you move beyond remembering definitions and toward actually using `dplyr` to answer questions with data.

**Option A**: If you have little to no statistics background, practice the review by following along. Type learning memos and code into your notes document. This practice will also help you understand the context and purpose of the functions you will be using.

**Option B**: If you have some statistics background, you can skim the review and move to the `dplyr` coding sections. You can refer back to the foundations as needed while you work through the coding sections.
 
## Data Collection

Data is a collection of facts, such as numbers, words, measurements, observations, or even just descriptions of things. Data can be qualitative (descriptive) or quantitative (numerical). It can be collected through various methods such as surveys, experiments, or observations.

**Population**: A population is a collection of objects or individuals that we want to study. It can be finite or infinite. The unknown population number is called a *parameter*.

* **Census**: A census is a complete enumeration of all the elements in a population. It is used to collect data from the entire population.

**Sample**: A sample is a subset of the population that we actually observe or measure. It is used to make inferences about the population. We use samples because it is often impractical or impossible to collect data from the entire population. The known sample number is called a *statistic*.

* Population Size **N**: The number of elements in the population.

* Sample Size **n**: The number of elements in the sample.

## Data Types

Your data set is most likely a sample from a larger population and not a census of the entire population.

1. **Unordered Sample**: A sample where the order of the elements does not matter. For example, if we have a population of 5 elements and we take a sample of 3, the set `{1, 2, 3}` **is** the same as `{3, 2, 1}`.

🎯 Create the vector `heights_sample` to store the set values: $5.9ft$, $6.1ft$, $5.8ft$, $6.0ft$, $5.7ft$. This is a sample of heights of *5 individuals* from a population of all individuals in a city.

```{r}
# ⚡
heights_sample <- c(5.9, 6.1, 5.8, 6.0, 5.7) # heights in feet
```
The order has no inherent meaning.

2. **Ordered Sample**: A sample where the order of the elements does matter. Time series data, growth measurements, finishing positions in a race, and data sorted from smallest to largest for calculating quantiles or medians are examples of ordered samples.

🎯  If we take 5 samples of the height of *one* youth as they grow up, we have an ordered sample. 

```{r}
# ⚡
growth_data <- c(42, 45, 48, 51, 54)  # Heights in inches by age
```

🗣 The order of the values matters because they represent measurements taken at different ages. The first value (42) corresponds to the height at the youngest age, and the last value (54) corresponds to the height at the oldest age. The set `{42, 45, 48, 51, 54}` is **NOT** the same as `{45, 54, 48, 42, 51}`.

Data that can be divided into categories or groups is called **categorical data**. For example, the colors of cars in a parking lot (red, blue, green, etc.). In R, categorical data is often represented as factors. Factors can be ordered or unordered.

🎯 A list of placements in a race is an ordered sample because the order of the placements matters. The first place finisher is different from the second place finisher, and so on. 

```{r}
# ⚡
race_positions <- c("Kim", "Alex", "Jordan", "Taylor")  # Order indicates ranking
```

🗣 The set `{Kim, Alex, Jordan, Taylor}` is **NOT** the same as `{Taylor, Jordan, Alex, Kim}`.

**Ordinal data** is about whether your data values have a meaningful natural ranking order, the intervals between categories aren't necessarily equal. Other examples include Likert scale or types of other hierarchical values like academic hierarchies.

`heights_sample` and `heights_sample` are **numerical data**. This is data that can be measured or counted. Numerical data can be further classified into:

* **Discrete Data**: Numerical data that can take on only specific values. For example, the number of cylinders in a car (4, 6, 8, etc.). In R, discrete data is often represented as integers. In R, we store a vector of counts or whole numbers by including an "L" after the number to indicate that it is an integer (e.g., `c(1L, 2L, 3L)`).

* **Continuous Data**: Numerical data that can take on any value within a range. For example, the height of individuals in a population (e.g., `heights_sample` and `heights_sample`).

___

When describing data, you will need to choose the appropriate statistical method based on the type of sample data you have. 

1. Check the type of sample data

2. Apply the appropriate descriptive statistics

# Summary Values

Descriptive statistics are used to summarize and describe the main features of a dataset. They provide a way to understand the distribution and variability of the data. This section introduces some common descriptive statistics so that you know what kind of summary you want when use `dplyr` verbs `group_by()` and `summarize()`.

## Counts

Use count-based summaries when your question is about how many observations you have or how often values appear. Length, sum, count are all ways to add up values in a dataset. 

### Length

**Length** counts the number of values in the set of numbers. In other words, it tells us how many observations we have. In a data table, it tells us how many rows are in a column.

🎯 Count the number of values in the `mpg` column of the `mtcars` dataset.

```{r}
# ⚡ length (Step 1a)
n <- length(mtcars$mpg)
```

___

🌳 Use this inline R code to dynamically report the number of values in the `mpg` column and include it in a sentence. Write the sentence in your notes document outside of a code chunk and without `#`.

```markdown

🗣 The `mpg` column in the `mtcars` dataset contains `{r} length(mtcars$mpg)` values.

```

When rendered to HTML, this will show as:

🗣 The `mpg` column in the `mtcars` dataset contains `r length(mtcars$mpg)` values.

Test rendering with the inline code to make sure it works. 
___

*You can also use `nrow(mtcars)` to get the number of rows in the dataset, which should give the same result as `length(mtcars$mpg)` since each row has one value for `mpg`.*

Later in this chapter, when you work inside `summarize()`, you will often use `n()` to count rows within groups.

### Count and Mode

When you want to know how often each value appears in a column, use the **`dplyr`** function `count()`. This is especially useful for categorical variables and discrete variables. A frequency table shows how many times each unique value appears.

The **mode** is the value that appears most often. In other words, once you create a frequency table, the mode is the value with the largest count.

📜 The SYNTAX: 

`data %>%`  
  `count(variable)`

*Use `?count` to learn more about the function.*

🎯 Use `count()` to create a table of the **frequency** of each unique value in the `cyl` column of the `mtcars` dataset. Determine the mode.

```{r}
# ⚡ Count discrete values 
cyl_counts <- mtcars %>%
  count(cyl)
# ✅ 
cyl_counts
```

🗣 In this table, `n` is the number of times that each unique value appears in the `cyl` column. 8 cylinders occur most frequently (14 times) in the `cyl` column of the `mtcars` dataset.

*Note: If no value occurs more than once, then all are the mode or no mode. You get to pick.*

---

🪞 What is the difference between `length()` `sum()` and `count()`? 
 
 - `length()` counts the number of values in a column, 
 
 -`sum()` adds up the values in a column, and 
 
 - `count()` counts the frequency of each unique value in a categorical column.
 
 For `cyl`, we care about how many cars in the dataset have 4, 6, or 8 cylinders (`count()`) and not about the total of all cylinder values added together (`sum()`). A frequency table helps us describe categorical data, and the mode helps us identify which category appears most often.
 
Mode is one measure of central tendency, or a measure of center. The next sections introduce two other measures of center that are more appropriate for numerical data.
 
## Center

Measures of **central tendency**, or measures of center, are used to describe the typical value in a dataset. They give us an idea of where the center of the data is located. Common measures of central tendency include the mean, median, and mode.

### Median

The **median** is the middle value of a set of numbers when they are arranged in order. The median is often useful when the data may contain unusually high or low values. 

**Step 1**: Sort the data in ascending order.

**Step 2**: Check the size of the sample (n). If there is an even number of values, the median is the average of the two middle values.

**Step 3**: Calculate the median using the appropriate formula based on whether n is odd or even.

* If n is odd, the median is the middle value: $$\text{median} = x_{(\frac{n+1}{2})}$$

* If n is even, the median is the average of the two middle values: $$\text{median} = \frac{x_{(\frac{n}{2})} + x_{(\frac{n}{2}+1)}}{2}$$

Where $x_{(i)}$ represents the ith ordered value in the sample.

___

🌳  Type this in your notes document to see how to write the median calculation mathematically.

```math
$$
\text{median} = x_{(\frac{n+1}{2})}
$$

$$
\text{median} = \frac{x_{(\frac{n}{2})} + x_{(\frac{n}{2}+1)}}{2}
$$
```

___

🎯 Calculate the median of the `mpg` column in the `mtcars` dataset.

#### Step 1

```{r}
# ✅  Pre-check
print(mtcars$mpg)
```

```{r}
# ⚡ Median Step 1: Sort the data
sorted_data <- sort(mtcars$mpg)  
# ✅ 
sorted_data
```

___

🌳 Use this inline code to dynamically report the size of the sample.

```markdown
🗣 The `mpg` column in the `mtcars` dataset sorted in ascending order is: `{r} sorted_data`.
```

When rendered to HTML, this will read:

🗣 The `mpg` column in the `mtcars` dataset sorted in ascending order is: `r sorted_data`.

___

#### Step 2

🎯 Find the size of the sample to determine how to calculate the median.

```{r}
# ⚡ Median Step 2: Get the size of the sample
length(sorted_data)
```

___

🌳 Use this inline code to dynamically report the size of the sample.

```markdown
🗣 There are `{r} length(sorted_data)` values.
```

When rendered to HTML, this will read: 

🗣 There are `r length(sorted_data)` values.

___

We need to calculate the median for a dataset with an even number of data points.

#### Step 3

To do this manually, we find the two middle values and take their average. The two middle values are the 16th and 17th values in the sorted list.

🎯 Find the median using the bracket operator `[]` to access the 16th and 17th values in the sorted list.

```{r}
# ⚡ Median Step 3: Calculate the median for an even number of values
median_value <- (sorted_data[16] + sorted_data[17]) / 2
# ✅
median_value
```

R simply gives us the median value without needing to do the manual calculations.

🎯 Use `median()` function to find the middle of the distribution of the `mpg` column in the `mtcars` dataset.

```{r}
# ⚡ Or use the median() function
median(mtcars$mpg)
```

___

🌳 Use this inline code to dynamically report the median value.

```markdown
🗣 The median of the `mpg` column in the `mtcars` dataset is `{r} median(mtcars$mpg)`.
```

When rendered to HTML, this will show as:

🗣 The median of the `mpg` column in the `mtcars` dataset is `r median(mtcars$mpg)`.

___

The median is a measure of center that is sometimes selected when the data contains outliers. The most common measure of center is the mean, which is the average of the values in a dataset. The next section introduces the mean and how to calculate it.

### Mean

The **mean** is the average of a set of numbers. It is calculated by summing all the values and dividing by the number of values.

**Step 1a** is to count the number of values in the dataset.

**Step 1b** is to sum the values in the dataset.

**Step 2** is to divide the sum by the number of values in the dataset.

#### Step 1a

We calculated the number of values in the `mpg` column of the `mtcars` dataset in the previous section

```{r}
# ⚡ length (Step 1a)
n <- length(mtcars$mpg)
```

#### Step 1b

**Summation** is the process of adding a sequence of numbers together. Summation is represented by the Greek letter sigma (Σ). It indicates that we are summing a series of values. 

___

🌳  Type this in your notes document to see how to write summation in Markdown with LaTeX syntax.
```math
$$
\sum_{i=1}^{n}
$$

```

$$\sum_{i=1}^{n}$$

When you see $\sum_{i=1}^{5} a$, this means:

* Start with $i=1$ (the second value in the set of numbers)

* Add the value $a$ to your sum

* Increase $i$ by 1

* Repeat until you reach $i=5$ (the fifth value in the set of numbers)

* Then stop

This is adding $a$ four times (once each for $i=2$, $i=3$, $i=4$, and $i=5$).

```math
$$
\sum_{i=1}^{5} a => a_1 + a_2 + a_3 + a_4 + a_5 
$$
```
$$
\sum_{i=1}^{5} a => a_1 + a_2 + a_3 + a_4 + a_5 
$$

___

🎯 Sum the values in the `mpg` column of the `mtcars` dataset. Store the result as sum_sample. 

```{r}
# ⚡ Sum (Step 1b)
sum_x <- sum(mtcars$mpg)
# ✅ 
sum_x
```

`sum(mtcars$mpg)` is the total value from adding all the values in the `mpg` column of the `mtcars` dataset.

___

🌳 Use this inline code to dynamically report `sum_sample`.

```markdown
🗣 The sum of the values in the `mpg` column of the `mtcars` dataset is `{r} sum_x`.
```

When rendered to HTML, this will show as:

🗣 The sum of the values in the `mpg` column of the `mtcars` dataset is `r sum_x`.

You can also use `sum(mtcars$mpg)` directly in the inline code without storing it in a variable.

```markdown
🗣 The sum of the values in the `mpg` column of the `mtcars` dataset is `{r} sum(mtcars$mpg)`
```
___


#### Step 2: Divide by the number of values

___

🌳  Type this in your notes document to see how to write mean calculation mathematically using LaTeX syntax.

```math
$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i
$$
```
$$\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$$

___

🎯 Calculate the mean of the `mpg` column in the `mtcars` dataset.

```{r}
# ⚡ Mean (Step 2) Divide Step 1 values
x_bar <- sum_x / n 
```

🗣 We used `sum_x` and `n` objects created earlier.

🎯 Calculate the mean of the `mpg` column in the `mtcars` dataset using the `mean()` function.

```{r}
# ⚡
# OR use mean() function directly
x_bar <- mean(mtcars$mpg) 
# ✅ 
x_bar
```

___

🌳 Use this inline code to dynamically report the mean value.

```markdown
🗣 The mean of the values in the `mpg` column of the `mtcars` dataset is `{r} mean(mtcars$mpg)`.
```

When rendered to HTML, this will show as:

🗣 The mean of the values in the `mpg` column of the `mtcars` dataset is `r x_bar`.

___

The next section, which explores measures of variability describe how spread out the values in a dataset are.

# Spread

**Spread** measures tell you how much individual data points deviate from the central tendency. Low spread indicates consistent, predictable data, while high spread suggests more variability and uncertainty. Unusual spread patterns can reveal data quality issues, outliers, or measurement errors that need investigation. Models built on data with known spread characteristics allow analysts to provide confidence intervals and uncertainty estimates. Comparing the spread of different groups, time periods, or conditions helps identify which scenarios are more stable or predictable. Understanding data spread helps choose appropriate analytical techniques and transformation methods.

## Standard Deviation

The **standard deviation** is a measure of how spread out the values in a dataset are. It is calculated as the square root of the variance.

**Step 1** is to calculate the variance, which is the average of the squared differences from the mean.

**Step 2** is to take the square root of the variance to get the standard deviation.

### Step 1

The **variance** is a measure of how spread out the values in a dataset are. It is calculated as the average of the squared differences from the mean.

Variance is calculated by first finding the difference between each value and the mean, squaring those differences, summing them up, and then dividing by the number of values minus one (for a sample variance).

___

🌳  Type this in your notes document to see how to write the variance calculation mathematically.

```math
$$
s^2 = \frac{\sum_{i=1}^{n} (x_i - \bar{x})^2}{(n-1)}
$$
```

$$s^2 = \frac{\sum_{i=1}^{n} (x_i - \bar{x})^2}{(n-1)}$$

___

🎯 Calculate the variance of the `mpg` column in the `mtcars` dataset.

```{r}
# ⚡
# Variance step 3: Calculate the variance
variance <- sum((mtcars$mpg - x_bar)^2) / (n - 1)
# ✅ 
variance
```

🗣 We used `x_bar` (mean) and `n` (column length), objects created earlier. Let's break down this nested function to make sure we understand what is happening. 

First, the `mtcars$mpg - x_bar` calculates the difference between *each value* in the `mpg` column and the *mean* of the `mpg` column. 

Rapping that list of difference values `( )` to the power of two `(mtcars$mpg - x_bar)^2` squares each of the differences. 

These squared differences are summed together. 

The sum of squared differences is divided by the length of the column minus 1, `(n - 1)`.  

This results in the sample variance ($s^2$).

🎯 Calculate the variance of the `mpg` column in the `mtcars` dataset using the `var()` function.

```{r}
# ⚡
# OR use var() function directly
variance <- var(mtcars$mpg)
# ✅ 
variance
```

___

🌳 Use this inline code to dynamically report the variance value.

```markdown
🗣 The variance of the values in the `mpg` column of the `mtcars` dataset is `{r} var(mtcars$mpg)`.
```

When rendered to HTML, this will show as:

🗣 The variance of the values in the `mpg` column of the `mtcars` dataset is `r var(mtcars$mpg)`.

You can round the variance to two decimal places for better readability.

```markdown
🗣 The variance of the values in the `mpg` column of the `mtcars` dataset is `{r} round(var(mtcars$mpg), 2)`
```
___


### Step 2

Find the **standard deviation** by taking the square root of the variance. 

___

🌳 Type this in your notes document to see how to write the standard deviation calculation mathematically.

```math
$$
s = \sqrt{s^2} = \sqrt{\frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}}
$$
```

$$s = \sqrt{s^2} = \sqrt{\frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}}$$

___

🎯 Calculate the standard deviation of the `mpg` column in the `mtcars` dataset.

```{r}
# ⚡
# Standard Deviation step 4: Calculate the standard deviation
std_dev <- sqrt(sum((mtcars$mpg - x_bar)^2) / (n - 1))
std_dev
```

🗣 This is the formula for $s^2$ inside of the function `sqrt()` which calculates the square root of a value.


🎯 Calculate the standard deviation of the `mpg` column in the `mtcars` dataset using the `sd()` function.

```{r}
# ⚡
# OR use sd() function directly 
std_dev <- sd(mtcars$mpg)
# ✅ 
std_dev
```

___

🌳 Use this inline code to dynamically report the standard deviation value.

```markdown
🗣 The standard deviation of the values in the `mpg` column of the `mtcars` dataset is `{r} sd(mtcars$mpg)`
```

When rendered to HTML, this will show as:

🗣 The standard deviation of the values in the `mpg` column of the `mtcars` dataset is `r sd(mtcars$mpg)`

You can round the standard deviation to two decimal places for better readability.

```markdown
🗣 The standard deviation of the values in the `mpg` column of the `mtcars` dataset is `{r} round(sd(mtcars$mpg), 2)`
```

___

Standard deviation works seamlessly with other statistical calculations (confidence intervals, hypothesis tests, regression). It uses every data point, providing a complete picture of variability. It is essential for parametric statistics and probability calculations, enabling standardized comparisons across different scales. But standard deviation is heavily influenced by extreme values, which can be misleading. It is less meaningful for skewed or non-normal distributions. Also, squared units can be less intuitive to interpret. 

In the next section, we will explore measures of spread that are less sensitive to outliers and skewed data.

## Range

This simplest measure of variability is the **range**, which is the difference between the largest and smallest values in a dataset. $x{(n)}- x{(1)}$

Sensitive to outliers, or extreme values, which can skew the range. 

### Interquartile Range (IQR)

The **interquartile range** (IQR), or the *fourth spread*, is the range of the middle 50% of the data. It is calculated as:

$$IQR = Q_3 - Q_1$$

IQR is less sensitive to outliers than the range because it focuses on the central portion of the data.

Where $Q_1$ is the **first quartile** (25th percentile) and $Q_3$ is the **third quartile** (75th percentile).

Remove lower and upper 25% removes extreme values and determine the range between the first and third quartiles.

### Percentiles

A **percentile** is a value below which a given percentage of observations fall. The 100 × p-th percentile (where p is between 0 and 1) is the value such that p × 100% of the data falls below that value.

100% of the data are less than or equal to the 100% percentile. Divide the data into 100 equal parts.

**Quartile**: divide the whole data range into four equal parts.

$Q_1$ is the lower quartile (25th percentile), $Q_2$ is the median (50th percentile), and $Q_3$ is the upper quartile (75th percentile).

#### Step 1 

🎯 Sort the data from the smallest to the largest

```{r}
# ⚡
# Sort the mpg column in ascending order
sorted_mpg <- sort(mtcars$mpg)
```

```{r}
# ✅ Verify data is sorted
print(sorted_mpg)
```

#### Step 2

🎯 Find the position value

```{r}
# ⚡
# Find minimum value of mpg in mtcars
min_mpg <- sorted_mpg[1]
# ✅ 
min_mpg
```

🎯 Find the minimum value of the `mpg` column in the `mtcars` dataset using the `min()` function.

```{r}
# ⚡
# Or simply use min() function
min_mpg <- min(mtcars$mpg)
# ✅ 
min_mpg
```

___

🌳 Use this inline code to dynamically report the minimum value.


```markdown
🗣 `{r} min(mtcars$mpg)` is the minimum value in the `mpg` column of the `mtcars` dataset.
```

When rendered to HTML, this will show as:

🗣 `r min(mtcars$mpg)` is the minimum value in the `mpg` column of the `mtcars` dataset.

___


#### Step 3

🎯 Find the position of the 25th percentile (Q1) in the sorted data and then find the value at that position.

*Type ?ceiling to learn about the `ceiling()` function.*

```{r}
# ⚡ Find the 25th percentile (Q1) of mpg in mtcars

## Step 1: Get the size of the sample
n <- length(mtcars$mpg)

## Step 2: Calculate the position of Q1
Q1_position <- 0.25 * (n + 1)


## Step 3: Find the value at the Q1 position
Q1_value <- sorted_mpg[ceiling(Q1_position)]

# ✅ 
Q1_value
```

🎯 Find the position of the 25th percentile (Q1) using the `quantile()` function.

*Type ?quantile to learn about the `quantile()` function.*

```{r}
# ⚡
# Or simply use quantile() function
Q1_value <- quantile(mtcars$mpg, 0.25)
# ✅ 
Q1_value
```

___

🌳 Use this inline code to dynamically report the 25th percentile.

```markdown
🗣 The 25th percentile (Q1) of the `mpg` column in the `mtcars` dataset is `{r} Q1_value`.
```

When rendered to HTML, this will show as:

🗣 The 25th percentile (Q1) of the `mpg` column in the `mtcars` dataset is `r Q1_value`.

___


#### Step 4

🎯 Find the position of the 50th percentile (median) in the sorted data and then find the value at that position.

```{r}
# ⚡ Find the median (50th percentile) of mpg in mtcars

## Step 1: Calculate the position of the median
median_position <- 0.5 * (n + 1)

## Step 2: Find the value at the median position# 
median_value <- sorted_mpg[ceiling(median_position)]

# ✅ 
median_value
```

🎯 Find the position of the 50th percentile (median) using the `median()` function.

```{r}
# ⚡
# Or simply use median() function
median_value <- median(mtcars$mpg)
# ✅ 
median_value
```

___

🌳 Use this inline code to dynamically report the median.

```markdown
🗣 The 50th percentile (median) of the `mpg` column in the `mtcars` dataset is `{r} median_value`.
```

When rendered to HTML, this will show as:

🗣 The 50th percentile (median) of the `mpg` column in the `mtcars` dataset is `r median_value`.

___

#### Step 5

🎯 Find the position of the 75th percentile (Q3) in the sorted data and then find the value at that position.

```{r}
# ⚡ Find the 75th percentile (Q3) of mpg in mtcars

## Step 1: Calculate the position of Q3
Q3_position <- 0.75 * (n + 1)

## Step 2: Find the value at the Q3 position
Q3_value <- sorted_mpg[ceiling(Q3_position)]
# ✅ 
Q3_value
```
🎯 Find the position of the 75th percentile (Q3) using the `quantile()` function.

```{r}
# ⚡
# Or simply use quantile() function
Q3_value <- quantile(mtcars$mpg, 0.75)
# ✅ 
Q3_value
```

___

🌳 Use this inline code to dynamically report the 75th percentile.

```markdown
🗣 The 75th percentile (Q3) of the `mpg` column in the `mtcars` dataset is `{r} Q3_value`.
```

When rendered to HTML, this will show as:

🗣 The 75th percentile (Q3) of the `mpg` column in the `mtcars` dataset is `r Q3_value`.

___


#### Step 6

🎯 Find the position of the maximum value in the sorted data and then find the value at that position.

```{r}
# ⚡ Find maximum value of mpg in mtcars
max_mpg <- sorted_mpg[n]
# ✅ 
max_mpg
```

🎯 Find the position of the maximum value using the `max()` function.

```{r}
# ⚡
# Or simply use max() function
max_mpg <- max(mtcars$mpg)
# ✅ 
max_mpg
```

___

🌳 Use this inline code to dynamically report the maximum value.

```markdown
🗣 `r max(mtcars$mpg)` is the maximum value in the `mpg` column of the `mtcars` dataset.
```
When rendered to HTML, this will show as:

🗣 `r max(mtcars$mpg)` is the maximum value in the `mpg` column of the `mtcars` dataset.

___


### `summary()`

You can use the `summary()` function to get a quick overview of the distribution of a numeric variable, including the minimum, 1st quartile, median, mean, 3rd quartile, and maximum values.

🎯  Use `summary()` to get a quick overview of the distribution of the `mpg` column in the `mtcars` dataset.

*Type `?summary` to learn about the `summary()` function.*

```{r}
# ⚡
summary(mtcars)  # Summarizes all columns at once
```

🎯 Use `summary()` to get a quick overview of the distribution of the `mpg` column in the `mtcars` dataset.

```{r}
# ⚡
summary(mtcars$mpg)  # Summarizes just the mpg column
```
_____

So far, you have learned how to calculate descriptive statistics manually and using built-in functions. You have also learned how to use inline R code to dynamically report these statistics in your Quarto document.

In the remaining of this chapter, we will compute specific summary statistics. You will learn how to group your data by certain variables before summarizing. This is where `dplyr`'s `summarize()` and `group_by()` functions come in handy.

# dplyr's summarize()

You can use `summarize()` to compute summary statistics like the mean, median, etc.

`?summarize` 

📜  The SYNTAX
`data %>%` 
  `summarize(new_column_name = some_operation(column_name))`
  
🎯 Find the overall mean mpg for all cars.

```{r}
# ⚡
mtcars %>% 
  summarize(mean_mpg = mean(mpg))
```

🗣 The code above creates a new column called `mean_mpg` that contains the mean of the `mpg` column from the `mtcars` dataset.

**We name the new column using `new_name = calculation`.**

🎯 Find several summary statistics for mpg: mean, max, and count of cars.

```{r}
# ⚡
mtcars %>% 
  summarize(mean_mpg = mean(mpg),
            max_mpg = max(mpg),
            count = n())
```

🧐 **Notice:** There is not piping `%>%` inside of the `summarize()` function. Instead, a comma `,` is used between the `mean()`, `max()` and `n()` calculations. 

🪞 `summary()` is a base R function that gives a quick overall description of an object, such as basic statistics for each variable in a data frame. `summarize()` from `dplyr` is used to create your own summary results, usually after `group_by()`, such as the mean or count for each group.

# dplyr's group_by()

`group_by()` changes the behavior of summary functions.

📜 The SYNTAX

`data %>%` 
  `group_by(variable1, variable2, ...) %>%`
  `some_operation()`

🎯 Find the mean mpg for each cylinder count.

```{r}
# ⚡
cyl_means <- mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))
```

🗣 `group_by(cyl)` splits the cars into groups based on cylinder count, and `summarize()` calculates `mean(mpg)` for each group. `mean_mpg` is the name of the new summary column.

🧐 **Notice:**  that we chain `group_by()` and `summarize()` together using the pipe operator `%>%`. This allows us to first group the data by the specified variable(s) and then apply the summary function to each group.

```{r}
# ✅ 
cyl_means  
```

🪞 Grouping cars by `cyl` is useful because cars with different numbers of cylinders often have different fuel efficiency, so one overall average mpg would hide those differences. By grouping first, you can compare the typical mpg for 4-, 6-, and 8-cylinder cars separately.

## Explicit vs. Default Naming

If you do not give the summary column a name, **`dplyr` uses the summary expression itself as the column name** by default. So `summarize(mean(mpg))` creates a column named `mean(mpg)` instead of a cleaner name like `mean_mpg`.

🎯 Find the mean mpg for each cylinder count without naming the summary column.

```{r}
# ⚡
mtcars %>%
  group_by(cyl) %>%
  summarize(mean(mpg))      
```

🗣 The default name is the code itself. Default names can be confusing when sharing results. Naming columns makes tables clearer. Use it whenever you share results.

# Multiple Statistics

🎯 We can create several summary columns in one call, separated by commas.

*`?n` to learn about the `n()` function.*

```{r}
# ⚡
cyl_stats <- mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg),
            sd_mpg   = sd(mpg),
            n        = n())
# ✅
cyl_stats
```

🗣 `n()` counts rows in each group. This is similar to the function `length()` which counts the number of values in a vector or row in a data set. It is different from length because `n()` works within `summarize()` and counts the number of rows in each group created by `group_by()`. In this way, `n()` is like `count()`, but it is used within `summarize()`. Using `n()` is a quick way to see how many observations are in each group. 

# Multiple Variables

`group_by` one column creates one group for each unique value. `group_by` two columns creates one group for each unique combination. 

❓How many groups will `group_by(cyl, am)` create? 

🎯 Predict how many groups will be created by `group_by(cyl, am)` before running the code.

```{r}
# ✅  Pre-check
# Recall that mtcars$cyl has three unique values:
unique(mtcars$cyl)
# Recall that mtcars$am has two unique values:  
unique(mtcars$am)
```

💬 Two variables with 3 and 2 unique values creates 3 x 2 = 6 groups.

🎯 Find the mean mpg for each combination of cylinder count and transmission type.

```{r}
#⚡
mtcars %>%
  group_by(cyl, am) %>%
  summarize(avg_mpg = mean(mpg),
            count = n())
```

🗣 Comment (example):
- Group 1: cyl = 4, am = 0 (4-cylinder automatic cars)

- Group 2: cyl = 4, am = 1 (4-cylinder manual cars)

- Group 3: cyl = 6, am = 0 (6-cylinder automatic cars)

- Group 4: cyl = 6, am = 1 (6-cylinder manual cars)

- Group 5: cyl = 8, am = 0 (8-cylinder automatic cars)

- Group 6: cyl = 8, am = 1 (8-cylinder manual cars)

## ungroup()

With two grouping vars, the default keeps the earlier group(s). With `.groups = "drop"` (or `ungroup()`), the result is not grouped.

A. Default behavior: keeps earlier groups

🎯 Show that with two groups `(cyl, gear)`, `summarize()` keeps the earlier group `(cyl)` by default.

```{r}
# ⚡
# A) Default behavior: "drop_last" (keeps earlier groups)
cyl_gear_stats <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg))   # no .groups set
```


```{r}
# ✅ 
group_vars(cyl_gear_stats)  # what groups remain?
```

```{r}
# ✅ Verify
is_grouped_df(cyl_gear_stats) # expect TRUE
```

🗣 The `Default = drop_last` from `(cyl,gear)` down to `(cyl)`. 

You’ll want to be explicit when you don’t want any grouping left.

B. Make it explicit inside `summarize()`

🎯 Return a plain, ungrouped table by telling `summarize()` to drop all groups.

```{r}
# ⚡
# B) Make it explicit: drop *all* groups in the result
cyl_gear_stats_drop <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg), .groups = "drop")
```


```{r}
# ✅ 
group_vars(cyl_gear_stats_drop)  # expect character(0)
```

```{r}
# ✅ 
is_grouped_df(cyl_gear_stats_drop) # expect FALSE
```

🗣 Using `.groups = "drop"` tells `summarize()` to remove all grouping after the summary operation, resulting in a plain data frame without any grouping structure.

C. Alternative: drop groups after summarizing

🎯 Achieve the same ungrouped result using `ungroup()` after `summarize()`

```{r}
# ⚡
# C) Alternative: drop groups after summarizing
cyl_gear_stats_ungroup <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg)) %>%
  ungroup()
```


```{r}
# ✅ 
group_vars(cyl_gear_stats_ungroup)  # expect character(0)
```

```{r}
# ✅ Verify
is_grouped_df(cyl_gear_stats_ungroup) # expect FALSE
```

🗣 Using `ungroup()` after `summarize()` removes all grouping from the resulting data frame, similar to using `.groups = "drop"` within `summarize()`. Both methods achieve the same outcome of returning an ungrouped data frame.

Pick either B) or C) based on your coding style preference.

# Summary



# Chapter Terms

**Categorical Data**: Data whose values place observations into groups or categories. The categories may have no natural order (nominal) or may have an order with uneven gaps between categories (ordinal). In R, categorical data are often stored as factors.

**Census**: A complete count or measurement of every member of a population, rather than just a subset.

**Central Tendency**: A measure that represents the center or typical value of a dataset. Common measures of central tendency include the mean, median, and mode.

**Continuous Data**: Numerical data that can take any value within a range, including values between other values. Measurements such as height, weight, and time are typical examples.

**Descriptive statistics**: Numerical summaries that describe the main features of a dataset, such as measures of central tendency (mean, median, mode) and measures of variability (range, variance, standard deviation).

**Discrete Data**: Numerical data that can take only specific, separate values, often counts such as 0, 1, 2, 3, and so on. In practice, discrete data are commonly integer-valued.

**Frequency**: The number of times a value or category appears in a dataset.

**Interquartile Range (IQR)**: The range of the middle 50% of the data, calculated as `Q3 - Q1`. Because it ignores the lowest 25% and highest 25%, it is less sensitive to extreme values than the full range.

**Length**: The number of elements in an object. For a vector, `length(x)` is the count of elements in the vector; for a list, it’s the number of list components; for a data frame, it returns the number of columns.

**Mean**: The arithmetic average of a set of numbers, found by adding all values and dividing by the number of values.

**Median**: The middle value in an ordered dataset. If there is an even number of values, the median is the average of the two middle values.

**Mode** (statistic): The value or category with the greatest frequency in a dataset. A dataset can have one mode, more than one mode, or no mode if no value occurs more often than the others.

**Numerical Data**: Data recorded as numbers, where the values have meaningful magnitudes and consistent intervals. Numerical data are also called *quantitative* data.

**Ordered Sample**: A sample in which the order of the selected elements does matter, so `(1, 2, 3)` is different from `(3, 2, 1)`. In counting language, this is the idea behind a *permutation*.

**Ordinal Data**: Categorical data with a meaningful ranking or order, but where the differences between adjacent categories are not necessarily equal or measurable. Examples include Likert-scale responses such as “strongly disagree” to “strongly agree.”

**Parameter**: A numerical summary of a population.

**Percentile**: A value that marks the point at or below which a given percentage of observations fall. For example, the 50th percentile is the median.

**Population**: The entire collection of individuals, cases, or objects you want to study. 

**Quartile**: One of the cut points that divides ordered data into four equal parts. The first quartile (`Q1`) is the 25th percentile, the second quartile (`Q2`) is the median, and the third quartile (`Q3`) is the 75th percentile.

**Range**: A simple measure of spread found by subtracting the smallest value from the largest value in a dataset.

**Sample**: A subset of the population that you actually observe or measure.

**Spread**: How much the values in a dataset vary or scatter rather than cluster tightly together. Common measures of spread include range, interquartile range (IQR), variance, and standard deviation.

**Standard Deviation**: A measure of spread equal to the square root of the variance. Because it takes the square root, standard deviation is expressed in the same units as the original data.

**Statistic**: A numerical summary computed from a sample.

**Summation**: The process of adding a sequence of values together. In formulas, summation is written with the capital Greek letter sigma, `Σ`, as a compact way to show a sum.

**Unordered Sample**: A sample in which the order of the selected elements does not matter, so `{1, 2, 3}` represents the same selection as `{3, 2, 1}`. In counting language, this is the idea behind a *combination*.

**Variance**: A measure of spread that describes how far values tend to vary from the mean, based on squared deviations. In introductory statistics, sample variance is commonly calculated using `n - 1` in the denominator.

# 📝 Practice Space

This Practice Space helps you turn Chapter 13 concepts into a reproducible analysis workflow. In this chapter, you practiced [Complete this]

This work matters in this course because [you are not only learning to write R code. You are learning to document your reasoning, check your results, and produce work that can be re-run and understood later]Complete this]. That connects directly to learning outcomes skills: [Complete this].

These tasks give you practice with the main skills from Chapter 13. These are common data-analysis tasks. Analysts often need to [Complete this]

## Setup

1. Create a new Quarto file in your course folder and save it as: 

- `chapter13_practice.qmd`

- Test Render your document often to see how your Practice Space report is coming together.

- Submit `chapter13_practice.html`

2. Use the heading structure taught earlier in the chapter.
Your Quarto setup for Chapter 13 should include headings, section numbering, and a table of contents, and you should render often to check both formatting and errors.

3. 📦 Library it up!

- Make sure there is script in your document that loads `dplyr` and `dslabs` packages so their functions / datasets load.

4. Memo your work with the 🎯/⚡/✅ /🗣

- a memo before the code chunk that states the goal of the code and what you are trying to learn or understand

- the code chunk itself

- any needed checks after the code

- a memo after the code chunk that explains what you learned, discovered, or confirmed

- answers to all comment questions included in the task

If you used help while working, document it in a memo. For example, if you used an LLM or another outside resource to understand an error or concept, briefly disclose that use and explain what help you received. Type all code yourself. This is part of the course expectation for AI-supported learning and memo-based workflow.

___

Follow this template for each coding task. 

🎯 What is the goal of the code? What am I trying to learn, check, or demonstrate?

```{r, eval=FALSE}
#⚡Code 
your_code <- "here"
```

```{r, eval=FALSE}
# ✅ Include Checks after and pre-checks as needed
colnames(your_code)
head(your_code)
table(your_code)
class(your_code)
unique(your_code)
```


🗣/🪞 What I learned / Next step. 1–2 sentences.

___


## Task 1

🤔 Reflect: How many rows would you expect if each state has 6 diseases? Make a prediction and test it. Then write a memo 🗣 Commenting about the results or 🪞 Explaining why this will be useful.

✍️ Practice: Pipeline. Fill in the Blanks

For `mtcars`, compute minimum and maximum horsepower for each transmission type (0 = automatic, 1 = manual). Store as **`transmission_hp`**.

```{r , eval=FALSE}
transmission_hp <- mtcars %>%
  ____(__) %>%
  ____(min_hp = min(____),
            max_hp = max(____))

transmission_hp
```

🚀 Explore and Play: Try changing horse power to miles per gallon or weight and rerun. Create a memo note, 🗣 Commenting about the results or 🪞 Explaining why this will be useful.

✍️ Practice: Replace the blanks (___) and run. 

You'll use the `heights` dataset and use the functions `group_by`, `summarize`, `mean`, `sd`, `n`

```{r , eval=FALSE}
data("heights")
?heights
```
```{r , eval=FALSE}
heights %>%
   ___(sex) %>%
   ___(mean_height = ___(height),
             sd_height   = ___(height),
             n           = ___()) -> height_descriptives

```



## Task 2

**Single grouping**

🌟 Using `stars`, compute mean temperature **AND** mean magnitude for each type. Name the result **`twinkle_twinkle`**.

🎯 Restate the Goal

```{r, eval=FALSE}
#⚡ Code
____ <- _____
```


```{r, eval=FALSE}
# ✅ Checks

```

🗣 Did it work? What happened? 

## Task 3

**Single grouping + three stats**

✔️ Pick ONE numeric variable in `olive` (not region, area), `group by region`, return `mean`, `sd`, and `n`.  Store as **`Olive_Garden`**.

🎯 Restate the Goal

```{r, eval=FALSE}
#⚡ Code
____ <- _____
```

```{r, eval=FALSE}
# ✅ Checks

```

In a short comment, note the region with the highest mean value.

🗣 "___"

## Task 4

**Multiple grouping**

⚠️ From `us_contagious_diseases`, `group_by` `state` AND `disease` then  `summarize` the mean `count`. Keep only the three diseases with the highest national mean `counts`. Store as **`Infectious_Burden`**.

⚡ Hint: **National means by disease.** Group the dataset by `disease` and summarize the mean of `count` across all rows. Pick the top 3 diseases. Arrange the diseases by their national mean and keep the top three.


```{r, eval=FALSE}
data("us_contagious_diseases")
# National means per disease
__________ <- us_contagious_diseases %>%
  _________(______) %>%
  _________(national_mean = ____(____), _____ = _______)%>%
  _________(desc(__________))%>%
  head(n = __)

```

```{r, eval=FALSE}
# ✅ Checks

```

🗣 Did it work? What happened?


## Task 5

**Putting it all together**

🚀 Using `dplyr`'s `starwars` data, create a table of **per-group descriptive statistics** by **two grouping variables**. Use several statistics and more than one numeric variable in a single `summarize()` call. Return an **ungrouped** result. Name it `starwars_summary`.

1. Inspect: Open the help page and glimpse the data. Identify **two categorical** columns you can group by (e.g., `species` and `sex`) and **two numeric** columns to summarize (e.g., `height` and `mass`).

```{r}
data("starwars")
```

2. In one code chunk: 

- Use `filter()` to remove rows with missing values so your groups are well-defined.

- `group_by()` your two categorical variables.

- `summarize()` to compute *at least* **two descriptive statistics** (e.g., mean, sd, n) for **each numeric variable**.

- Drop groups to return an ungrouped result.

- Name the result **`starwars_summary`**.

```{r, eval=FALSE}
#⚡ Code
____ <- _____
```

```{r, eval=FALSE}
# ✅ Checks

```

🗣 Did it work? What happened?

## Task 6

🪞 Reflection

📝 Write a short paragraph reflecting why we need to do `group_by()` come BEFORE `summarize()`?

✍️ EXPLANATION: "___"

