---
title: "Chapter 12"
subtitle: "Select & Filter"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Chapter 12"
subtitle: "mutate and casewhen"
author: "by Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: false
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
# ---
title: "Chapter 12"
subtitle: "Select & Filter"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```


# Overview


In Chapter 11, you learned about about subsetting with dplyr (`select()` & `filter()`). This week, we will look at other functions in the dplyr package (`mutate()` and `case_when()`). Real-world data often needs new or cleaner variables before analysis. The dplyr verb `mutate()` lets you create or modify columns, while `case_when()` supplies the logic for quick recoding. Together they form a duo you will use in nearly every data project.

**Lesson Overview**

By the end of Chapter 12 you will be able to:

1. ⚡ Remember – State the purposes of `mutate()` and `case_when()`

2. 🔍 Understand – Explain how `case_when()` replaces values based on a test.

3. 🛠 Apply – Overwrite or add columns using `mutate()` + `case_when()`

4. 📊 Analyze – Build multi‑level recodes with nested `case_when()`.

5. 🧠️ Evaluate – Choose between overwriting vs. keeping the original data.

Keep these goals in mind as you move through each section.

# Quarto: Part 2

In Chapter 11, you learned how to create a Quarto file, keep the YAML header, use Markdown headings, write memos outside code chunks, and control chunks with #| options. In this chapter, you will build two more Quarto skills that make your notes more readable and more reproducible:

1. add a table of contents and section numbering

2. use inline R code inside your sentences

🎯 Create a Quarto File (🌱)

1. For those of you who **have GitHub** and a course project, navigate to your course folder and open your course project. This will open RStudio. Click "Pull" in the Git pane to get the latest version of the course materials. After completing work, save, stage, commit, and push. 

If you **don't have GitHub** or a course project, just open RStudio and set your working directory to your course folder.

2. Open a new Quarto document in your course folder and save it as: `chapter12_notes.qmd` (`File > New File > Quarto Document`).

## toc: true (🌳)

As your notes get longer, it becomes harder to scroll and find what you need. Quarto can automatically build a table of contents from your headings, and it can also number your sections for you. The main YAML options for this are `toc: true`, `toc-depth:`, and `number-sections: true`. These options are commonly placed in the document options near the top of the YAML.

Use this updated YAML header for Chapter 12:

```{r echo=FALSE, results='asis'}
cat("```yaml
---
title: \"Chapter 12 Notes\"
subtitle: \"mutate() and case_when()\"
author: \"YOUR NAME\"
date: today
toc: true
toc-depth: 2
number-sections: true
format:
  html:
    code-fold: true
execute:
  warning: true
  message: true
---
```")
```

### What each part does:

- `toc:` true adds a table of contents automatically.

- `toc-depth:` 2 tells Quarto to include level 1 and level 2 headings in that table of contents.

- `number-sections: true` adds section numbers to your headings in the rendered document.

The `toc-depth: 2` is a good default. It keeps your notes organized without making the table of contents too crowded.

## Dynamic memoing (🌳)

So far, you have used code chunks to run R code. Quarto also lets you place small pieces of R code directly inside normal sentences. This is called inline code. Inline code is useful when you want your written memo to automatically include a value computed by your R code. Quarto evaluates inline code during rendering, so the sentence stays up to date with your analysis.

In Quarto, inline code uses one pair of backticks and the engine name inside the backticks. For R, the pattern is:

```markdown

`{r} expression`

```

Using inline R code helps ensure that your report always shows the exact values used to in calculations. 

### Inline R code

🎯 Imagine you want to write a sentence that says how many rows are in the filtered `mtcars` dataset. Instead of hardcoding the number, you can use inline code to compute it. 

```{r message=FALSE, warning=FALSE}
# ⚡ Type this in a Quarto code chunk
library(dplyr)
num_rows <- mtcars %>% 
  filter(cyl == 6) %>% 
  nrow()
```

Then, in your written memo, you can write the expression: 

```markdown
🗣 The filtered dataset has `{r} num_rows` rows.
```

When rendered to HTML, this will show as:

🗣 The filtered dataset has `r num_rows`  rows.

💡 **Tip:** Keep inline code simple. Use it to display values that you already created in a regular code chunk. Do not try to do lots of work inside the sentence itself. 

For this chapter, set up your notes file so that:

1. the YAML includes a table of contents and section numbering

2. your headings organize the chapter clearly

3. your code chunks do the main work

4. your memos use inline R code whenever you want to report a value from your analysis

___

Now that you have created (*and saved*) your quarto file, ... [finish this]

# Set up 

To get started, 

## Headings

A simple outline for Chapter 12

```markdown

# Load Data

# 

```

## 📦 Packages

In this chapter, you will use `dplyr` verbs and `dslabs` datasets.

Remember that you only need to install once `install.packages("dplyr")`; `install.packages("dslabs")`.

🎯 Load the required packages at the beginning of your QMD.

```{r , eval=FALSE}
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
```

[transition statement]

# Vectorized Conditional functions

In earlier chapters, you learned how to use bracket indexing to subset, extract, and insert values. With **logical indexing**, you built a **logical vector** of `TRUE` and `FALSE` values and then used that pattern to keep or modify selected positions. For example, `x[x > 3]` extracts only the values greater than `3`, and `x[x > 3] <- "Large"` would replace matching positions for the word "Large". 

```{r}
# 📖 Review
## Create vector x
x <- 1:6
## logical vector
x > 3
```
```{r}
# 📖 Review
## extract values
x[x > 3]
```
```{r}
# 📖 Review
## insert values
x[x > 3] <- "Large"
## ✅  print result
x
```
      
In Chapter 11, you used `filter()` for the same basic idea at the data-frame level: it keeps the rows that satisfy a condition and drops the rest. 




Subsetting answers the question, “Which values or rows do I keep?”

This chapter introduces two new vectorized conditional functions: `ifelse()` and `case_when()`. They still rely on logical conditions checked element-by-element across a vector, but they do something different from subsetting. Instead of keeping only the matching values, they return a new vector the same length as the input, filling each position according to a rule. 

Base R describes `ifelse()` as returning a value with the same shape as the test, filled from `yes` or `no`, and `dplyr` describes `case_when()` as a general vectorized if-else that evaluates cases in order and uses the first match for each element.



[Write an intro]



## `ifelse()` tests 

Conditonal element selection

📜 The SYNTAX: `ifelse(test, yes, no)`

- *test*  – logical vector (`TRUE`/`FALSE`/`NA`)

- *yes*   – value to insert where test is `TRUE`

- *no*    – value to insert where test is `FALSE` or `NA`

🎯 Goal: Label numbers > 3 as "Large" and others as "Small".

💬 Prediction: I expect to see a vector of "Small" and "Large" labels. Specifically, I expect only 4 and 5 to be "Large".

```{r , eval=FALSE}
ifelse(test = c(1, 2, 3, 4, 5) > 3, yes = "Large", no = "Small")
```

✅ Check / 🗣 Comment: Confirm the output positions for 1:3 are "Small"  and 4:5 are "Large".

## `case_when()` tests

dplyr::case_when() can do exactly what `ifelse()` is doing. 

📜 The SYNTAX: `case_when(condition1 ~ value1, condition2 ~ value2, TRUE ~ default_value)`

The `~` symbol means "maps to" or "yields".

🎯 Goal: Label numbers > 3 as "Large" and others as "Small".

💬 Prediction: I expect to see a vector of "Small" and "Large" labels. Specifically, I expect only 4 and 5 to be "Large".

```{r , eval=FALSE}
case_when(c(1, 2, 3, 4, 5) > 3 ~ "Large", TRUE ~ "Small")
```

✅ Check / 🗣 Comment: Confirm the output positions for 1:3 are "Small"  and 4:5 are "Large"

So, what's the point? Why do we need both `ifelse()` and `case_when()`? `case_when()` is usually clearer once you have 3+ categories.


## Multi‑Level Recoding 

Nested `ifelse()`

🎯 Goal: Re-code numbers 1 to 5 as "One", "Two", "Three", "Four", or "Five".

💬 Prediction: I expect to see a vector of the words "One" through "Five" corresponding to the numbers 1 through 5.

```{r , eval=FALSE}
# ⚡Code: 
object2 <- c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5)
length(object2) # Check the length of object2
```


```{r , eval=FALSE}
# ⚡Code
ifelse(object2 == 1, "One",
       ifelse(object2 == 2, "Two",
              ifelse(object2 == 3, "Three",
                     ifelse(object2 == 4, "Four", "Five"))))
```


✅ Check / 🗣 Comment:  Confirm the output length is 10 and the labels match the numbers.

Nested `case_when()`

🎯 Goal: Re-code numbers 1 to 5 as "One", "Two", "Three", "Four", or "Five".

💬 Prediction: The `case_when()` output should be identical in length and order to the nested `ifelse()` version.

```{r , eval=FALSE}
# ⚡Code
case_when(object2 == 1 ~ "One",
          object2 == 2 ~ "Two",
          object2 == 3 ~ "Three",
          object2 == 4 ~ "Four",
          object2 == 5 ~ "Five")
```

✅ Check / 🗣 Comment:  Confirm the output is identical to the `ifelse()` version.

🪞 What I learned: `case_when()` is easier to read because the function name is not repeated at each line.

Apply this nested `case_when()` in Task 3 of the assignment.

# `mutate()`

`mutate()` is a function that adds new columns or modifies existing ones

📜 The SYNTAX: `mutate(data, new_col = value)`

## Overwrite Columns

### Copy A Column

🎯 Goal: Create a new column that is a copy of an existing column.

```{r , eval=FALSE}
# Preview Column Names
names(mtcars) # Check the column names
```


💬 Prediction: In a new data set called `mtcars2`, a **new** column named `new_column_copy` is created that is identical to the `am` column.

```{r , eval=FALSE}
# ⚡ Code: 
mtcars2 <- mtcars %>% 
  mutate(new_column_copy = am)
```

```{r , eval=FALSE}
# ✅ Verify:
names(mtcars2) # Check the column names again
```

🗣 Comment: Did you see the new column in the output?

```{r , eval=FALSE}
# ✅ Verify: is am == to n_column_copy?
mtcars2$am == mtcars2$new_column_copy
```


```{r , eval=FALSE}
# ✅ Verify: is am identical to new_column_copy?
identical(mtcars2$am, mtcars2$new_column_copy)
```

🗣 Comment: `mutate(new_column_copy = am)` created an identical column of `am` called `new_column_copy`

### Overwrite Original Column Values

🎯 Goal: Recode the `am` column in `mtcars` from 0/1 to "Automatic"/"Manual".

```{r , eval=FALSE}
# Check the unique values in the am column BEFORE
unique(mtcars$am) 
```

💬 Prediction: In a new data set called `cars_overwrite`, the `am` column is changed from 0/1 to "Automatic"/"Manual". In a second new data set called `cars_keep`, a **new** column named `transmission` is created that has "Automatic"/"Manual" values, while the original `am` column remains unchanged.

```{r , eval=FALSE}
# ⚡ Code: 
cars_overwrite <- mtcars %>%
  mutate(am = case_when(am == 0 ~ "Automatic", TRUE ~ "Manual"))
```

```{r , eval=FALSE}
# ✅ Verify:
unique(cars_overwrite$am)
```

### New Column & New Values

🎯 Goal: Create a new column `transmission` that recodes `am` from 0/1 to "Automatic"/"Manual", while keeping the original `am` column unchanged.

```{r , eval=FALSE}
# ⚡ Code: 
cars_keep <- mtcars %>%
  mutate(transmission = case_when(am == 0 ~ "Automatic", TRUE ~ "Manual"))
```

```{r , eval=FALSE}
# ✅ Verify:
cars_keep %>% select(am, transmission)
```

🗣 Comment: What is the difference between the code that creates `cars_overwrite` and `cars_keep`?

🪞 Metacognition: Why might you want to create a copy of a column when changing the values?

## Factors and Levels

In Lesson 2, we learned about factors for vectors. A factor is a categorical variable that can take on a limited number of values, called levels. For example, in the vector `tshirt_sizes_data`, the possible values are "S", "M", "L", and "XL".

```{r eval=FALSE}
# ⚡ Creating a character vector
tshirt_sizes_data <- c("M", "L", "S", "M", "L", "XL", "S")
# ⚡ Changing the character vector to a factor class with levels
tshirt_sizes <- factor(tshirt_sizes_data, levels = c("S", "M", "L", "XL"))
```

```{r eval=FALSE}
# ✅ Investigating your factor class with levels
unique(tshirt_sizes)
```
 
In the dataset `iris`, the `Species` column is a factor with three levels: "setosa", "versicolor", and "virginica".
 
```{r eval=FALSE}
data(iris)
?iris
```

```{r eval=FALSE}
# ✅Investigating the Species column
unique(iris$Species)
class(iris$Species)
levels(iris$Species)
```

### Overwrite with Factors

🎯 Goal: Take a 100-row sample from `movielens`, then overwrite the numeric `rating` column with an ordered factor: "Bad" if `< 4`, "Good" if `>= 4`, using `case_when()` inside `mutate()`.

💬 Prediction: Before: `rating` is numeric. After: `rating` is an ordered factor with levels `Bad < Good`. Row count stays 100; no `NA`s created.

📜 SYNTAX reminder:
`mutate(data, new_col = case_when(cond1 ~ value1, cond2 ~ value2, TRUE ~ default))`

`name_data <- data %>% mutate(new_col = case_when(cond1 ~ value1, cond2 ~ value2, TRUE ~ default))`

```{r , eval=FALSE}
data("movielens")
?movielens
```

```{r , eval=FALSE}
set.seed(99)

# Sample 100 movies from the movielens dataset
smaller_movies <- movielens %>% 
  sample_n(100) %>% 
  select(title, rating)

head(smaller_movies)
```

```{r , eval=FALSE}
# ✅ Checks (before)
names(smaller_movies)                # expect "title", "rating"
class(smaller_movies$rating)         # expect "numeric"
summary(smaller_movies$rating)       # quick distribution check
```

🎯 Goal: We'll apply `mutate()` to `smaller_movies` to recode the rating column to "Good" or "Bad".

```{r , eval=FALSE}
# ⚡ Overwrite numeric rating with an ordered factor via case_when()
movies_overwrite <- smaller_movies %>%
  mutate(
    rating = factor(
      case_when(
        rating >= 4 ~ "Good",
        rating < 2 ~ "Bad",
        TRUE        ~ "Ok"
      ),
      levels = c("Bad", "Ok", "Good"),
      ordered = TRUE
    )
  )
```


```{r , eval=FALSE}
# ✅ Checks (after)
names(movies_overwrite)              # still "title", "rating"
```


```{r , eval=FALSE}
# ✅ Checks (after)
is.ordered(movies_overwrite$rating)  # expect TRUE
```


```{r , eval=FALSE}
# ✅ Checks (after)
levels(movies_overwrite$rating)      # expect "Bad", "Ok", "Good"
```


```{r , eval=FALSE}
# ✅ Checks (after)
table(movies_overwrite$rating)       # counts in each level
```

🗣 Comment: Overwriting keeps the column name but replaces its contents and type (numeric → ordered factor). 

🪞 Metacognition: `case_when()` reads cleaner than nested `ifelse()` for multi-branch recodes and makes me specify a default. In the future, I could keep the original numeric ratings (e.g., `rating_num`) and add a new labeled column instead of overwriting, to preserve provenance.


```{r , eval=FALSE}
# ⚡ Overwrite numeric rating with an ordered factor via case_when()
movies_rename <- smaller_movies %>%
  mutate(
    rating_words = factor(
      case_when(
        rating >= 4 ~ "Good",
        rating < 2 ~ "Bad",
        TRUE        ~ "Ok"
      ),
      levels = c("Bad", "Ok", "Good"),
      ordered = TRUE
    )
  )
```

### Bonus Example

🎯 Goal: Create a new column rating_label in smaller_movies using `case_when()` that labels ratings `>= 4` as "Good" and the rest as "Bad". Keep the original numeric rating column.

💬 Prediction: After running, the data will still have 100 rows, keep both `title` and `rating`, and add `rating_label` (character or factor). No `NA`s should appear.

```{r , eval=FALSE}
# ✅ Checks (before)
names(smaller_movies)                # expect "title", "rating"
class(smaller_movies$rating)         # expect "numeric"
```

```{r , eval=FALSE}
# ⚡ Add a NEW column using case_when()
smaller_movies %>%
  mutate(
    rating_label = case_when(
      rating >= 4 ~ "Good",
      TRUE        ~ "Bad"
    )
  ) -> movies_newcol
```

🧐 NOTICE: What is happening here `-> movies_newcol`?

🛠Break Things! How can you reorganize this script? 

```{r , eval=FALSE}
# ✅ Checks (after)
names(movies_newcol)                 # expect "title", "rating", "rating_label"
class(movies_newcol$rating)          # expect "numeric"
class(movies_newcol$rating_label)    # expect "character" or "factor"
```
🗣 The code used -> movies_newcol, which assigns right-to-left.

🛠 Break Things! What happens with `head(movies_newcol) -> movies_newcol`?

🪞 Metacognition: I’ll stick with the usual `<-` to make my code easier to read

## Calculations

🚗 Using `mtcars`, let's create a column `l_per_100km` that converts from US units (miles per gallon) to metric units (liters per 100km). 235.215 is the conversion factor between these units

🎯 Goal: Identify the most fuel-efficient car in L/100 km terms.

💬 Prediction: The new column `l_per_100km` will be added to `mtcars_units`, and the most fuel-efficient car will have the lowest value in this column.


```{r , eval=FALSE}
# ⚡ Code
mtcars_units <- mtcars %>%
  mutate(l_per_100km = round(235.215 / mpg, 1))
```

```{r , eval=FALSE}
# ✅ Check
head(mtcars_units)
```

👀 Check‑in: Which car is most fuel‑efficient in l/100 km terms?

```{r , eval=FALSE}
# Find the most fuel-efficient (smallest L/100km)
best_idx <- which.min(mtcars_units$l_per_100km)
best_car <- rownames(mtcars_units)[best_idx]
best_value <- mtcars_units$l_per_100km[best_idx]
```

```{r , eval=FALSE}
"best car"
best_car
```

```{r , eval=FALSE}
"best value"
best_value
```

🪞 Metacognition: Unit conversions are easy to keep alongside original units with `mutate()` (don’t overwrite).

# Summary

# Chapter Terms

**Logical Indexing**: Indexing a vector using a logical vector of `TRUE`/`FALSE` values (for example, `x[x > 0]`) so R keeps the elements where the index is `TRUE` and drops those where it is `FALSE`. If the logical index is a different length, R may recycle it, which can lead to surprising results.

**Logical Vector**: A vector that stores Boolean elements: `TRUE` or `FALSE` (and sometimes `NA` for missing). Logical vectors are commonly produced by comparisons (for example, `x > 0`) and are heavily used for filtering and conditional code.

**Subsetting**: Selecting a smaller piece of an R object using indices (positions, names, or a logical mask). With data frames, subsetting most commonly uses `[` in the form `df[rows, cols]` and typically preserves the object’s structure (e.g., you still get a data frame when you subset rows and columns).

# 📝 Practice Space

Replace each ____ placeholder with working code or a short written answer. Run each section; be sure the requested objects appear in the Environment. When finished, submit practice as `chapter12_practice.qmd` along with the `chapter12_practice.RData` workspace.

Template (use around every code chunk):

🎯/❓ Goal. One sentence stating the outcome.

💬 Prediction. (Optional) What you expect to see before you run code.

```{r}
#⚡Code attempts.

# your code here
```

✅Checks (before/after). Show what you’re verifying (names, head(), table(), class(), etc.).

🗣/🪞What I learned / Next step. 1–2 sentences.

## Task 0

✍️ Practice: Recode the mpg column. Use `mutate()` and `case_when()`

Fill in the Blanks. 

🎯 Goal: Create `mpg_class` with three labels using `case_when()`:
"High MPG" if `mpg > 25`; "Low MPG" if `mpg < 15`; otherwise "Medium MPG".

```{r , eval=FALSE}
# ⚡ Code
mtcars %>%
  ______(mpg_class = _____(
    ___ > 25 ~ "High MPG",
    mpg < 15 ~ "_______",
    TRUE     ~ "Medium MPG"
  )) -> mpg_recode
```

```{r , eval=FALSE}
# ✅ Checks
table(mpg_recode$____)  # expect 3 levels
```





## Task 1

📦 Library it up!

Make sure there is script in your document that loads `dplyr` and `dslabs` packages so their functions / datasets load.

## Task 2

📝  Recode ranges (overwrite)

Using `quakes`, recode `mag` into THREE ordered levels of _your choice_ (e.g., "Low", "Mid", "High") **overwriting** the original column. Use `mutate()` and `case_when()` chain and pipe (`%>%`). Store as **`quake_recode`**.

```{r, eval=FALSE}
data("quakes")
?quakes
```

```{r, eval=FALSE}
____ <- _____
```

```{r, eval=FALSE}
# Quick check
table(quake_recode$mag)   # should show exactly 3 levels
```

## Task 3

Recode categories (new col)

🌟 With `stars`, turn the one-letter *type* into a descriptive phrase of *your own* in a **new** column called `full_type`.  Keep the original type. Store as **stars_recode**.

```{r, eval=FALSE}
data("stars")
?dslabs::stars
```

```{r, eval=FALSE}
____ <- _____
```

```{r, eval=FALSE}
# Quick check
head(select(stars_recode, type, full_type))
```

## Task 4

Unit conversion

Create `international_friendly_heights` by adding a `height_cm` column to `heights` (inches × 2.54). Leave original height.

```{r, eval=FALSE}
data("heights")
?heights
```

```{r, eval=FALSE}
____ <- _____
```

🤔 Reflect: Why add—not overwrite—the height column? 

✍️ EXPLANATION: "___"

# Save and Upload

1. You will be submitting **both** the R Markdown | Quarto Document and the workspace file. The workspace file saves all the objects in your environment that you created in this lesson. You can save the workspace by running the following command in a code chunk of the R Markdown | Quarto Document document:

```{r , eval=FALSE}
save.image("Assignment9_Workspace.RData")
```

Or you can click the "Save Workspace" button in the Environment pane. 

💾 **Always save the R documents before closing.**

2. Find the assignment in this week's module in Canvas and upload **both** the RMD and the workspace file.

# Today you practiced:

- Practiced logical tests and the structure of `case_when()`.

- Created binary and multi‑level recodes.

- Used `mutate()` to overwrite or create columns.

- Performed inline calculations for unit conversion.

🛠 Continue experimenting. Recoding and mutation are the building blocks of clean, analysis‑ready data!
