---
title: "Lesson 6 Grading"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Warm up




# ➤ Assignment

## Task 0

🚗 Setup: Load the built-in data set `mtcars` and take a quick look. Identify what the columns contain and meaning of that content.

```{r ,eval=FALSE}
data("mtcars")   # already in memory but this keeps the workflow explicit
?mtcars        # help file for variable descriptions
View(mtcars)   # spreadsheet view (optional)
```

🗣 Comment: Explain R’s basic subsetting template. Hint: It looks like `data[ rows , columns ]`

## Task 1 

⚖️ Manual vs. Automatic

1. Create `Manual_Cars`: all rows where `am == 1` (manual), all columns.

```{r}
# 1. All rows where am == 1, all columns
Manual_Cars <- mtcars[ mtcars[ , "am"] == 1 , ]

```

2. Use `nrow()` to record how many manual cars there are (store in `n_manual`).

```{r}
# 2. Count rows
n_manual <- nrow(Manual_Cars)
```

## Task 2 

🔍 MPG by Cylinder

Extract mpg for each cylinder group into separate objects. Hint: Repeat the bracket pattern

```{r , eval=FALSE}
Four_Cyl_MPG  <- mtcars[ mtcars[ , "cyl"] == 4 , "mpg" ]
Six_Cyl_MPG   <- mtcars[ mtcars[ , "cyl"] == 6 , "mpg" ]
Eight_Cyl_MPG <- mtcars[ mtcars[ , "cyl"] == 8 , "mpg" ]
```

## Task 3

🧐 Analyze Fuel Efficiency

1. Compute the mean mpg for each object.

```{r , eval=FALSE}
mean_4 <- mean(Four_Cyl_MPG)
mean_6 <- mean(Six_Cyl_MPG)
mean_8 <- mean(Eight_Cyl_MPG)
```

2. 🗣 Comment: Which cylinder group is most fuel‑efficient? Answer in one word or number.

Most efficient: 4 or "Four"

## Task 4

🚦 Safe vs. Direct Editing

1. Make a safe copy of `mtcars` called `cars_safe`.

```{r , eval=FALSE}
# 1. Safe copy
cars_safe <- mtcars
```

2. Rename the `'am'` column in `cars_safe` to `"Transmission_Type"`.

```{r , eval=FALSE}
# 2. Rename 'am' to 'Transmission_Type' using indexing
colnames(cars_safe)[ colnames(cars_safe) == "am" ] <- "Transmission_Type"
```

3. 🗣 Comment: Explain *why* working on cars_safe (a copy) is safer than editing mtcars directly.  

Editing a copy prevents accidental, irreversible changes to the original data (mtcars). If something breaks, you still have the untouched source.

## Task 5

📝 Practice `which()`

Using `cars_safe`, extract the rows for V‑shaped engines (`vs == 0`) *and* horsepower (`hp`) above `150`. Use `which()` for the row indices and keep only `mpg`, `hp`, and `cyl` columns. Name the object `hi_power_v`.

1. Create `row_idx` for the condition above.

- `cars_safe[ , "vs"]` ? `0` 

- `&` 

- `cars_safe[ , "hp"]` ? `150`

- `cars_safe[row_idx , c("mpg","hp","cyl")]`


```{r , eval=FALSE}
# 1. Row indices: V-shaped engines (vs == 0) AND hp > 150
row_idx <- which( cars_safe[ , "vs"] == 0 & cars_safe[ , "hp"] > 150 )
```

2. Subset cars_safe with `row_idx` and columns `mpg`, `hp`, `cyl`. Name object `hi_power_v`.

```{r , eval=FALSE}
# 2. Keep only mpg, hp, cyl
hi_power_v <- cars_safe[ row_idx , c("mpg", "hp", "cyl") ]
head(hi_power_v)
```

## Task 6

🛠 Insertion Example

Add a new column called `"Efficiency_Class"` to `cars_safe`. Label `"High"` if `mpg >= 25`, `"Low"` otherwise.

Hint: use `ifelse()` and `cars_safe[ , "mpg"] >= 25, "?", "?"`

```{r , eval=FALSE}
cars_safe[ , "Efficiency_Class"] <- ifelse(
  cars_safe[ , "mpg"] >= 25, "High", "Low"
)
```
 











