---
title: "Wildfire Signature Assignment"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  bookdown::html_document2: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../../assets/header.html
      after_body: ../../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Wildfire Signature Assignment"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
output: 
   bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
 
Or
title: "Wildfire Signature Assignment"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  bookdown::html_document2: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```

# Overview

🔥 Idaho is well known for its wildfires. They play a ecological role, shaping our forests and grasslands while presenting significant hazards to communities and resources. In recent years, the Idaho landscape has experienced frequent and sometimes severe wildfire seasons due to both natural and human activity. While fires are part of the natural ecosystem, they have become more destructive as the climate warms. Wildfire preparedness and management is a continuing challenge for the state. For more information this [research at Boise State](https://www.boisestate.edu/coas/2024/12/04/boise-state-experts-provide-perspective-on-idaho-mountain-wildfires/).


**How far could a new wildfire run in 8 hours? How many new wild fire ignitions will grow larger than 10 km (straight line) in 8 hours?**

In this three-part Core Signature Assignment activity option, you can work in groups but each person submits their own work independently. 


By the end of this Wildfire Signature Assignment you will be able to:

1. 🧭 Organize – Set up a reproducible RStudio workflow (project, files, data, RData) that can be re-run on another machine.

2. 🛠 Wrangle – Import the WHP sample data, create quantile-based hazard categories, and summarize areas and probabilities with tidy R code.

3. 📈 Model – Specify and use lognormal models (p/q/d/r *lnorm* functions) to answer distance and tail-probability questions by hazard category.

4. 🎲 Simulate – Build a multi-step Monte Carlo simulation (with `replicate()`) that mixes categories by area weights and estimates the chance of many fast-growing fires in a season.

5. 🗣 Communicate – Knit a clear Rmd/Qmd notebook that explains assumptions, code choices, and results, with appropriate metadata and citations for data and help sources.

6. 🪞 Reflect – Document troubleshooting, AI use, and learning-skill growth (curiosity, courage, tenacity, thoroughness, open-mindedness) through short memos embedded in the workflow.

# Setting Up Your Notes

To begin Wildfire guided activity, follow these steps:

1. Open your course project for RStudio 

For those of you who **have GitHub** and a course project, navigate to your course folder and open your course project. This will open RStudio. Click "Pull" in the Git pane to get the latest version of the course materials. After completing work, save, stage, commit, and push.

If you **don't have GitHub** or a course project, just open RStudio and set your working directory to your course folder.

2. Create a new file 

Create a new Quarto document for Wildfire Signature Assignment and save it with the file name `wildfire.qmd`

3. Follow along

Read the background material below, then complete the tasks that follow. Replace each ____ placeholder (and any TODO comments) with working code or a short written answer. Run each section; be sure the requested objects appear in the Environment. 

Use the template 🎯 Goal → ⚡ Code → ✅ Checks → 🗣 Comment to met the Complete criteria

1. 🎯 **Goal** – In your own words, one sentence about what you are trying to do.

2. ⚡ **Code** – Code that completes the task (replace each `____` with your own code).

3. ✅ **Checks** – Code that convinces you your solution works (structure, counts, summaries, etc.).

4.  🗣 **Comment** – 1–2 sentences either:
  - data-focused (what you learned about the wildfire data), **or**
  - code-focused (what the code does / why it works).

**Supporting**

- 📝 Format your document with headings, code chunks, and narrative text. Use proper grammar, spelling, and punctuation. Include a document title, author name, and 🕰 date in the heading of your report.

- 👀 Cite resources (including referencing lessons, peers, instructor guidance, internet websites, AI summaries, the help file in R) in memos. State specifically what or who helped and in what ways. Always compose your own code (no copy and paste). 

- 🚀 Code trials show how you explored ideas, asked questions independently of the assignment. Include code comments to explain your thinking.

- 🪞 Reflect on learning process, connects ideas, discusses challenges, and describes improvement strategies.

- 🧠 Growth mindset means you demonstrate effort, persistence, and a positive attitude toward learning new skills. You show courage.  

- 💬 Prediction means you make hypotheses about what the code will do or what the data will look like before running it, and then reflect on whether your predictions were correct.

- ✍️ Record what you tried, even if it didn’t work. This is proof of thoroughness and trustworthiness, prevents future duplication of errors, contextualizing successes, and exposes hidden discoveries (unanticipated effect).

## Dataset

Wildfire managers need tools that show where fires are more likely and how intense they could become. The U.S. Forest Service’s Wildfire Hazard Potential (WHP) is one such tool: a nationwide raster layer where each grid cell stores a unitless index of relative wildfire hazard (higher value = higher hazard). For this activity we use an Idaho-only WHP file prepared by Dr. Matt Williamson [SPASES Lab Page](https://www.spaseslab.com/). 

Figure 1 shows the Idaho WHP raster created in R with the packages `terra` (raster operations), `sf` (vector data), `tidyverse` (data wrangling), `tidyterra` (ggplot helpers for rasters), and `tigris` (state/county boundaries). The GeoTIFF uses an Albers Equal Area projection (units: meters) with 240 × 240 m cells. Within Idaho, valid cells cover about 216,315 km²; cells outside the state within the file’s bounding box are NoData. WHP values in this subset range from 0 to 64,185.656; the median is about 397, and the upper tail is steep (e.g., 95th percentile ≈ 7,753, 99th ≈ 22,429). 

**Formal dataset citation**

Dillon, Gregory K.; Gilbertson-Day, Julie W. 2020. Wildfire Hazard Potential for the United States (270-m), version 2020. 3rd Edition. Forest Service Research Data Archive. [https://doi.org/10.2737/RDS-2015-0047-3](https://doi.org/10.2737/RDS-2015-0047-3).

Idaho subset (whp_id.tif) prepared and provided by Matt Williamson, SPASES Lab, Boise State University.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'hide', fig.cap='Idaho Wildfire Hazard Potential (WHP), continuous index (unitless) from the USFS WHP v2020 raster, clipped to the Idaho boundary. Colors use a viridis gradient (dark purple = lower WHP; yellow = higher WHP); values span approximately 0–64,186. Raster resolution is 240 m in an Albers Equal Area projection; NoData cells are transparent. Red outline shows the TIGER/Line state boundary reprojected to the raster CRS. Idaho subset prepared by Matt Williamson (SPASES Lab, Boise State University); original dataset: Dillon & Gilbertson-Day (2020), USFS Research Data Archive (DOI: 10.2737/RDS-2015-0047-3).'}
## **First Install R Packages: 'terra', 'tidyterra', 'tigris'
library(terra)
library(tidyverse)
library(sf)
library(tidyterra)
library(tigris)

terraOptions(progress = 0)           # terra: no text progress
options(tigris_use_cache = TRUE)     # tigris: avoid re-download chatter


whp_idaho <- rast("whp_id.tif") 


state_bounds <- states() %>% ## download state cartographic boundaries form the TIGER shapefiles
  filter(STUSPS %in% c("ID")) %>% ##filter for the ID state boundary
  st_transform(., crs = crs(whp_idaho)) ## reproject to the raster projection

plot_whp <- ggplot() + #open the ggplot environment
  geom_spatraster(data = whp_idaho, mapping = aes(fill = WHP_ID)) + #use tidyterra to convert the data to an object that can be used in the geom
  geom_sf(data = state_bounds, fill = "transparent", color = "red", lwd = 0.5) + #add the state boundaries
  scale_fill_viridis_c(option = "C", na.value = "transparent") + #change the colorramp for the data and plot NA as transparent
  theme_bw() #use a nicer theme
plot_whp

# a raster of wildfire hazard potential for Idaho

```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# --- Assumes the packages in your script are already loaded ---
# terra, tidyverse, sf, tidyterra, tigris

# Alias for clarity
whp <- whp_idaho

# 1) BASIC METADATA -----------------------------------------------------------
meta_tbl <- tibble(
  name        = names(whp),
  n_layers    = terra::nlyr(whp),
  n_cells     = terra::ncell(whp),
  is_factor   = is.factor(whp),
  has_levels  = !length(terra::levels(whp)),  # should be FALSE now
  crs_wkt     = terra::crs(whp, proj = TRUE),
  is_lonlat   = terra::is.lonlat(whp),
  res_x_m     = terra::res(whp)[1],          # your file shows 240; units are meters
  res_y_m     = terra::res(whp)[2],
  xmin        = terra::ext(whp)[1],
  xmax        = terra::ext(whp)[2],
  ymin        = terra::ext(whp)[3],
  ymax        = terra::ext(whp)[4],
  value_units = "WHP index (unitless)",
  description = "Wildfire Hazard Potential (continuous index); Idaho subset"
)
```


```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# 2) SUMMARY STATS & QUANTILES -----------------------------------------------
basic_stats <- terra::global(whp, fun = c("min","max","mean","sd"), na.rm = TRUE) |>
  as_tibble() |>
  mutate(across(everything(), as.numeric))

probs <- c(0, .01, .05, .10, .25, .50, .75, .90, .95, .99, 1)


vals <- terra::values(whp, mat = FALSE, na.rm = TRUE)
quant_tbl <- tibble(
  prob  = probs,
  value = as.numeric(stats::quantile(vals, probs = probs, na.rm = TRUE))
)
```


```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# 3) AREA-WEIGHTED DECILES (robust to terra column names) ---------------------
deciles <- seq(0, 1, 0.1)
edges   <- as.numeric(stats::quantile(vals, probs = deciles, na.rm = TRUE))

whp_decile <- terra::app(
  whp,
  fun = function(v) as.integer(cut(v, breaks = edges, include.lowest = TRUE, labels = FALSE))
)
names(whp_decile) <- "decile"

# cell areas in km^2
area_km2_rast <- terra::cellSize(whp, unit = "km")

# Sum area by decile
area_by_decile <- terra::zonal(area_km2_rast, whp_decile, fun = "sum", na.rm = TRUE) |>
  as_tibble(.name_repair = "minimal")
names(area_by_decile)[1:2] <- c("decile", "area_km2")  # <— robust renaming

area_by_decile <- area_by_decile |>
  mutate(area_acres = area_km2 * 247.105381,
         area_pct   = round(100 * area_km2 / sum(area_km2), 3))

# Add WHP range per decile (robust renaming again)
min_by_dec <- terra::zonal(whp, whp_decile, fun = "min", na.rm = TRUE) |>
  as_tibble(.name_repair = "minimal"); names(min_by_dec)[1:2] <- c("decile","whp_min")
max_by_dec <- terra::zonal(whp, whp_decile, fun = "max", na.rm = TRUE) |>
  as_tibble(.name_repair = "minimal"); names(max_by_dec)[1:2] <- c("decile","whp_max")

decile_table <- area_by_decile |>
  left_join(min_by_dec, by = "decile") |>
  left_join(max_by_dec, by = "decile") |>
  mutate(whp_range = paste0(whp_min, "–", whp_max)) |>
  select(decile, whp_range, area_km2, area_acres, area_pct) |>
  arrange(decile)

```


```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# 4) AREA WITH MISSING DATA (if any) -----------------------------------------
na_area_km2 <- {
  na_r <- terra::ifel(is.na(whp), area_km2_rast, NA)
  as.numeric(terra::global(na_r, "sum", na.rm = TRUE)[1,1])
}
na_tbl <- tibble(area_km2 = na_area_km2,
                 area_acres = na_area_km2 * 247.105381)

# 5) OPTIONAL: COUNTY-LEVEL MEANS (Idaho) ------------------------------------
id_counties <- tigris::counties(state = "ID", cb = TRUE) |>
  sf::st_transform(crs = terra::crs(whp))
county_v <- terra::vect(id_counties)

county_mean <- terra::extract(whp, county_v, fun = mean, na.rm = TRUE) |>
  as_tibble() |>
  bind_cols(sf::st_drop_geometry(id_counties) |> select(COUNTYFP, NAME)) |>
  rename(mean_whp = 2, county_fips = COUNTYFP, county = NAME) |>
  arrange(desc(mean_whp))

# 6) WRITE TABLES -------------------------------------------------------------
readr::write_csv(meta_tbl,      "whp_idaho_metadata.csv")
readr::write_csv(basic_stats,   "whp_idaho_basic_stats.csv")
readr::write_csv(quant_tbl,     "whp_idaho_quantiles.csv")
readr::write_csv(decile_table,  "whp_idaho_area_by_decile.csv")
readr::write_csv(county_mean,   "whp_idaho_county_mean.csv")
readr::write_csv(na_tbl,        "whp_idaho_missing_area.csv")


library(terra)
library(sf)
library(tigris)
library(dplyr)
library(readr)

# --- 1) Sample raw WHP values (random, non-NA) -------------------------------
set.seed(2025)
N <- 150000

samp_pts <- terra::spatSample(
  whp_idaho,
  size      = N,
  method    = "random",
  na.rm     = TRUE,
  as.points = TRUE,
  values    = TRUE      # attaches WHP_ID attribute to the points
)

# --- 2) Add county attributes to each sampled point --------------------------
id_counties <- tigris::counties(state = "ID", cb = TRUE) |>
  sf::st_transform(crs = terra::crs(whp_idaho))

# spatial join (point -> county); keep only county name & FIPS
pts_sf   <- sf::st_as_sf(samp_pts)
pts_join <- sf::st_join(
  pts_sf,
  id_counties[, c("NAME","COUNTYFP")],
  left     = TRUE,
  largest  = TRUE
)

# --- 3) Optional: add per-cell area (km^2) as a weight -----------------------
# (Equal-area CRS + uniform cell size => constant ~0.0576 km^2, but we compute it explicitly)
area_km2_rast <- terra::cellSize(whp_idaho, unit = "km")
cell_area_km2 <- terra::extract(area_km2_rast, samp_pts, ID = FALSE)[,1]

# --- 4) Build the RAW tibble (no bins, no summaries) -------------------------
whp_raw_labeled <- pts_join |>
  sf::st_drop_geometry() |>
  transmute(
    whp          = WHP_ID,          # unitless WHP index
    county       = NAME,            # county name
    county_fips  = COUNTYFP,        # 3-digit county FIPS (Idaho state code = 16)
    cell_area_km2 = cell_area_km2   # area weight per row (useful for probabilities)
  )

# Save for students (raw data only)
write_csv(whp_raw_labeled, "whp_raw_sample.csv")

head(whp_raw_labeled)
```


# Part A

Explore and wrangling of the raw WHP sample data. 

## Task 1

**Load Data**

Task instructions:

1. Use **`readr`** to read the file `whp_raw_sample.csv` from your working directory.  

2. Store the result in an object named **`county_raw`** in your Environment.  

3. Run at least two **structural checks** to make sure the data imported correctly (for example: number of rows, variable names, a summary of key columns).  

4. Write a short comment describing what kind of dataset `county_raw` is (size, what a row represents, high-level purpose).

🎯 **Goal.** Write your own one-sentence goal here (e.g., what you want to load and understand).

```{r eval=FALSE, message=FALSE, warning=FALSE}
#⚡ Code: readr::read_csv
____ # 1. e.g., load any needed package(s)
____ # 2. e.g., read the CSV into county_raw
```

```{r eval=FALSE, message=FALSE, warning=FALSE}
# ✅  Explore Data
____ # 3. e.g., structure or glimpse of county_raw
____ # e.g., check column names or a summary of whp / cell_area_km2
```

🗣 Comment (TODO).
In 1–2 sentences, describe what county_raw represents (rows, key columns, why it might be useful).

___

### **Column meanings**

**whp** – The Wildfire Hazard Potential value for a single sampled raster cell. This is a unitless, continuous index (higher = greater relative hazard). Values come directly from the WHP GeoTIFF.

**county** – The county name (e.g., “Idaho”, “Twin Falls”) obtained by spatially joining each sampled cell to the TIGER/Line county boundary. Use for readable reports.

**county_fips** – The county’s 3-digit FIPS code from TIGER/Line. *Option: To create the standard 5-digit FIPS for Idaho, prepend the state code 16: e.g., 049 → 16049. Need to use the 5-digit code when joining to external datasets (ACS, MTBS summaries, etc.).*

**cell_area_km2** – The area (in square kilometers) represented by that raster cell, computed from the raster’s equal-area projection. With 240 m × 240 m cells this is ~0.0576 km²; small numeric differences reflect rounding. Summing this column by any grouping yields true area totals, and dividing by the statewide sum yields area shares.

___

## Task 2

**Basic data audit summaries**

Task instructions:

1. Using `dplyr::summarize()` (or base R equivalents), compute and report the following for county_raw:

- Total number of rows.

- Number of missing values in `whp`.

- Minimum value of `cell_area_km2`.

- Maximum value of `cell_area_km2`.

2. Store the summary tibble (or data frame) in a clearly named object (e.g., `county_summary`).

🎯 Goal. Write your own one-sentence goal about what you are summarizing.

```{r eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Code: summarize key audit quantities for county_raw
____ # 1/2. assign county_summary and use county_raw
____ # 1. compute n_rows
____ # 1. compute n_na_whp
____ # 1. compute min_area
____ # 1. compute max_area
```

✅ Checks: print the summary object and confirm the summary makes sense.

🗣 Comment (TODO).
In 1–2 sentences, interpret what these numbers say about the dataset quality or structure (missingness, range of area, etc.).

## Task 3

**Compute WHP quantile edges** (category cut points)

Task instructions

1. Use `stats::quantile()` on the `whp` column of `county_raw` to compute the quantiles for `prob` using `0, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 1.00`.

2. Store the result in an object named `e` (for “edges”).

3. Use` na.rm = TRUE` so missing values are ignored.

4. Look at `e` and make sure it has 8 ordered values that look reasonable for a right-skewed variable.

📜 The SYNTAX: quantile(x, probs, na.rm)

🎯 Goal.
Write your own goal explaining quantiles/edges goal.

```{r eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Code: Base R stats::quantile 
____ # 2. store WHP quantile edges in object e
____ # 1. call quantile() on county_raw's whp column
____ # 1. specify the probs vector
____ # 3. include remove NA from calculation
```

```{r eval=FALSE, message=FALSE, warning=FALSE}
# ✅  Verify
____ # print e
```

🗣 Comment (TODO).
In 1–2 sentences, describe what `e` represents and how these cut points will be used in later tasks.

Task 3 might have been challenging! Consider other kinds of memos: 🪞 Reflect; 🧠 Growth mindset; 👀 Cite resources; 🚀 Code trials; ✍️ Record what you tried.

___

✅  Verify Option: Each element corresponds to a specific percentile:

- `e[1]` = minimum WHP (0th percentile)

- `e[2]` = 25th percentile

- `e[3]` = 50th percentile (median)

- `e[4]` = 75th percentile

- `e[5]` = 90th percentile

- `e[6]` = 95th percentile

- `e[7]` = 99th percentile

- `e[8]` = maximum WHP (100th percentile)


___

## Task 4

**Add a categorical WHP risk label**

Task instructions:

1. Start from `county_raw`, create a new data frame called `county_cats`.

2. Use `dplyr::mutate()` and `dplyr::case_when()` to add a new variable `Category` that assign a label based on `whp` and the edges in `e`.

3. Use the following intervals (closed on the left (>=), open on the right (<), except the last, which is closed on both sides (>=; <=)):

- If a row’s `whp` value is *at least* the minimum WHP AND *less than* the 25th percentile, label it "Very Low".

- If `whp` is *at least* the 25th percentile AND *less than* the 50th percentile (median), label it "Low".

- If `whp` is *at least* the 50th percentile AND *less than* the 75th percentile, label it "Moderate".

- If `whp` is *at least* the 75th percentile AND *less than* the 90th percentile, label it "High".

- If `whp` is *at least* the 90th percentile AND *less than* the 95th percentile, label it "Very High".

- If `whp` is *at least* the 95th percentile AND *less than* the 99th percentile, label it "Extreme".

- If `whp` is *at least* the 99th percentile AND *less than or equal to* the maximum WHP, label it "Ultra (Top 1%)".

- If a value of `whp` does not fall in any of these ranges (e.g., `NA`), `Category` should be `NA`.

4. Convert `Category` to an **ordered factor** so the levels follow the intended risk order.
- `"Very Low"`, `"Low"`, `"Moderate"`, `"High"`, `"Very High"`, `"Extreme"`, `"Ultra (Top 1%)"`.

🎯 Goal.
Write your own sentence(s) goal for turning numeric WHP into ordered risk categories.


```{r eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Code:  dplyr::mutate + dplyr::case_when
____ # 1. assign county_cats and pipe county_raw into it 
____ # 2. use mutate() to create Category 
____ # 3. with case_when() and e[ ] bounds
____ # 4. convert Category to an ordered factor with the specified levels
```

```{r eval=FALSE, message=FALSE, warning=FALSE}
# ✅  Verify: confirm Category looks right
____ # e.g., count() or table()
____ # optional: str() or glimpse()
```

🗣 Comment (TODO).
Either: (a) explain briefly how the `case_when` logic translates WHP into categories, or (b) interpret what the distribution of categories says about Idaho’s wildfire hazard.

Task 4 might have been challenging! Consider other kinds of memos: 🪞 Reflect; 🧠 Growth mindset; 👀 Cite resources; 🚀 Code trials; ✍️ Record what you tried.

## Task 5

**Aggregate area and compute category probabilities**

1. Start from `county_cats`

2. Use `dplyr::group_by()` with `Category`.

3. Use `summarize()` to compute, for each `Category`:

- `area_km2` equals the sum of `cell_area_km2`.

- `n_cells` equals the number of grid cells using `n()`.

- Use `.groups = "drop"` so that the result is ungrouped.

4. Use `dplyr::mutate()` to create a new variable called `Probability` that is the `area_km2` divided by `sum(area_km2)`.

5. Use `dplyr::select()` to keep only `Category`, `Probability`, `area_km2`, `n_cells` in this order.

🎯 Goal.
Write your own one-sentence goal for summarizing WHP categories into areas and probabilities.

💬 Prediction.

```{r , eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Code: dplyr::group_by + dplyr::summarize + dplyr::mutate
____ # 1. store prob_tbl and pipe county_cats into it
____ # 2. group_by()
____ # 3. summarize()
____ # 4. mutate())
____ # 5. select()
```

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ✅  Verify: validate prob_tbl
____ # print prob_tbl
____ # check that probabilities sum to ~1
```

🗣 Comment (TODO).
In 1–2 sentences, describe what `prob_tbl` tells you about the share of Idaho’s area in each hazard category and how that connects to “probability of a random ignition’s category.”

# Part B

The table below summarizes the quantile-based wildfire hazard categories derived from Idaho’s continuous USFS WHP raster and the distance distributions used for an 8-hour run. The **Category** column lists seven bins (Very Low to Ultra) defined from the raster’s cumulative distribution; the WHP index itself is unitless. The **Probability** column is the statewide area share for each bin—expressed as a percent of valid raster area (NA excluded)—and is interpreted as the chance that a random ignition falls in that category.

```{r echo=FALSE, message=FALSE, warning=FALSE, tab.cap='Table 1. Quantile-based WHP categories for Idaho and 8-hour spread distance parameters. Probability = statewide area share (% of valid raster area); distances in kilometers; WHP is unitless.'}
library(terra); library(dplyr); library(readr); library(knitr); library(kableExtra)

# Alias for clarity
whp <- whp_idaho

# whp is your SpatRaster (continuous WHP for Idaho)
probs <- c(0, .25, .50, .75, .90, .95, .99, 1.00)
qv <- as.numeric(stats::quantile(values(whp, na.rm = TRUE), probs = probs, na.rm = TRUE))

# Build a categorical raster from those thresholds
whp_cat <- app(whp, fun = function(v) as.integer(cut(v, breaks = qv, include.lowest = TRUE, labels = FALSE)))
names(whp_cat) <- "cat"

# Compute area per category (km^2), excluding NA
area_km2 <- cellSize(whp, unit = "km")
area_by_cat <- zonal(area_km2, whp_cat, fun = "sum", na.rm = TRUE) |>
   tibble::as_tibble(.name_repair = "minimal")
names(area_by_cat)[1:2] <- c("cat", "area_km2")
area_by_cat <- mutate(area_by_cat, area_pct = area_km2 / sum(area_km2))

# Human labels + WHP bounds
labels <- c("Very Low","Low","Moderate","High","Very High","Extreme","Ultra (Top 1%)")
bounds <- tibble(
  cat = 1:7,
  whp_lo = qv[1:7],
  whp_hi = qv[2:8]
)

# Your distance parameters (update as needed)
param_tbl <- tibble::tibble(
  cat = 1:7,
  Category = labels,
  normal_mean_km = c(2.0, 3.0, 4.5, 7.0, 12.0, 18.0, 28.0),
  normal_sd_km   = c(1.0, 1.0, 1.5, 2.0, 3.0, 4.0, 6.0),
  mu_log  = c(0.582, 1.046, 1.451, 1.907, 2.455, 2.866, 3.310),
  sd_log  = c(0.472, 0.325, 0.325, 0.280, 0.246, 0.220, 0.212)
) |>
  left_join(bounds, by = "cat") |>
  left_join(area_by_cat, by = "cat") |>
  transmute(
    Category,
    Probability = round(100 * area_pct, 2),  # measured %
    `Mean (km)` = normal_mean_km,
    `SD (km)`   = normal_sd_km,
    `Mean log` = mu_log,
    `SD log`   = sd_log
  )

# Render as a real table so LaTeX/HTML uses the chunk caption
kbl(param_tbl,
    booktabs = TRUE,
    align    = c("l","r","r","r","r","r","r","r")) |>
  kable_styling(full_width = FALSE,
                latex_options = c("hold_position","scale_down"))

```

The table also defines a simple fire-spread scenario using "plausible values" for an 8-hour straight-line distance by hazard category. These distances are not estimated from the WHP raster. They are modeling assumptions chosen for this assignment. Specifically, we assume that (1) ignitions are equally likely anywhere within the valid WHP raster (so Probability reflects area share), (2) higher WHP categories correspond to larger 8-hour straight-line spread distances, and (3) within each category, 8-hour distances follow either a zero-truncated normal distribution with the given mean and SD or a lognormal distribution.

___

The Normal mean (km) and Normal SD (km) columns give the mean and standard deviation, in kilometers, for a zero-truncated normal distance model representing straight-line spread over 8 hours. The Mean log and SD log columns give the parameters \( \mu \) and \( \sigma \) of a lognormal distance model in which

\[
\log(\text{distance in km}) \sim \mathcal{N}(\mu,\sigma^2).
\]

Labels “Extreme” and “Ultra” denote upper-tail bins (approximately the 95–99% and 99–100% quantiles, respectively).

___

## Statistical Methods

A continuous random variable is a variable that can take on an infinite number of values within a given range. Unlike discrete random variables, which can only take on specific, separate values (🎲 like the roll of a die), continuous random variables can take on any value within a certain interval. Examples of continuous random variables include ⚖️ weight or 🕓 time. In this assignment, we will use 📏 distance in km. 

_You do not need to understand the text descriptions of the equations to complete the tasks. There are there for those who are studying statistics more deeply (me, your instructor)._

___

Let $X$ be the continuous random variable of straight-line kilometer distance in an 8-hour period. Then a probability distribution of $X$ is a function $f(x)$ such that for any two numbers $a$ and $b$ with $a \leq b$, 

$$
P(a \leq X \leq b) =  \int_{a}^{b} f(x) \, dx
$$

The probability that $X$ takes on a value in the interval $[a, b]$ is given by the area above this interval and under the graph of the density function. 

___

## Lognormal Distribution

In this assignment, we will use **lognormal distribution** to model the 8-hour straight-line distance of new ignitions in High wildfire hazard categories. We will explore the list of R functions for the lognormal distribution. You can access the documentation for these functions by running the following command in R:

```{r , eval=FALSE}
?Lognormal
```

A **lognormal distribution** is used to model a variable that is always positive and becomes normal after taking the logarithm of the values. 

```{r echo=FALSE}
# Parameters
meanlog <- 1.5
sdlog   <- 0.6
p_cut   <- 0.80   # chosen percentile

# Simulate data
x <- rlnorm(3000, meanlog = meanlog, sdlog = sdlog)
log_x <- log(x)

# Percentile on original scale
x_p <- qlnorm(p_cut, meanlog = meanlog, sdlog = sdlog)

# Grids for smooth curves
x_max   <- qlnorm(0.995, meanlog, sdlog)
x_grid  <- seq(0.001, x_max, length.out = 1000)
y_grid  <- dlnorm(x_grid, meanlog, sdlog)

x_shade <- seq(0.001, x_p, length.out = 500)
y_shade <- dlnorm(x_shade, meanlog, sdlog)

# Layout: two plots on top, one caption strip below
layout(matrix(c(1, 2, 3, 3), nrow = 2, byrow = TRUE),
       heights = c(4, 1.15))

par(oma = c(0, 0, 2.2, 0),
    mgp = c(2.4, 0.8, 0),
    tcl = -0.3,
    cex.main = 1.05)

# -------------------------
# Panel 1: Lognormal distribution
# -------------------------
par(mar = c(4.2, 4.2, 2.8, 1.2))

hist(x,
     breaks = 28,
     probability = TRUE,
     xlim = c(0, x_max),
     col = "gray88",
     border = "white",
     main = "Lognormal distribution",
     xlab = "Original scale (X > 0)",
     ylab = "Density")

# Shaded probability area
polygon(c(0, x_shade, x_p),
        c(0, y_shade, 0),
        col = adjustcolor("goldenrod1", alpha.f = 0.45),
        border = NA)

# Re-draw histogram lightly so bars remain visible
hist(x,
     breaks = 28,
     probability = TRUE,
     xlim = c(0, x_max),
     col = adjustcolor("lightblue3", alpha.f = 0.35),
     border = "white",
     add = TRUE)

# Density curve and percentile line
lines(x_grid, y_grid, col = "navy", lwd = 2)
abline(v = x_p, col = "firebrick", lwd = 2, lty = 2)

# Small percentile label only
usr1 <- par("usr")
text(x_p, usr1[4] * 0.93,
     labels = paste0(round(100 * p_cut), "th percentile"),
     pos = 4, cex = 0.82, col = "firebrick")

# -------------------------
# Panel 2: After taking logs
# -------------------------
par(mar = c(4.2, 4.2, 2.8, 1.2))

hist(log_x,
     breaks = 28,
     probability = TRUE,
     col = "gray82",
     border = "white",
     main = "After taking logs",
     xlab = "log(X)",
     ylab = "Density")

curve(dnorm(x, mean = meanlog, sd = sdlog),
      add = TRUE, col = "darkgreen", lwd = 2)

# -------------------------
# Panel 3: Caption / key strip
# -------------------------
par(mar = c(0.4, 1, 0.2, 1))
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))
box(col = "gray85")

# Row 1
rect(0.04, 0.63, 0.08, 0.78,
     col = adjustcolor("lightblue3", alpha.f = 0.35), border = NA)
text(0.10, 0.705, "Bars show the distribution of X",
     adj = 0, cex = 0.9)

segments(0.39, 0.705, 0.45, 0.705, col = "navy", lwd = 2)
text(0.47, 0.705, "Height of the curve = concentration (density)",
     adj = 0, cex = 0.9)

# Row 2
rect(0.04, 0.20, 0.08, 0.35,
     col = adjustcolor("goldenrod1", alpha.f = 0.45), border = NA)
text(0.10, 0.275,
     labels = paste0("Shaded area = probability = P(X <= percentile) = ", p_cut),
     adj = 0, cex = 0.9)

segments(0.58, 0.275, 0.64, 0.275, col = "firebrick", lwd = 2, lty = 2)
text(0.66, 0.275,
     labels = paste0("Percentile = x-value where area reaches ", p_cut),
     adj = 0, cex = 0.9)

mtext("A lognormal variable is positive, and log(X) is approximately normal",
      outer = TRUE, line = 0.5, cex = 1.05)
```


## Task 1

**Tail probability with `plnorm()`**

`plnorm()` computes the cumulative distribution function (CDF) for a lognormal distribution. It gives the probability that a lognormally distributed random variable is less than or equal to a certain value.

❓ Question a: If a new ignition starts in a **High-hazard area**, what is the probability that its 8-hour straight-line spread is at least 10 km?

___

Let $X$ be the continuous random variable giving the 8-hour straight-line spread distance (in km) of a new ignition in a High-hazard area. Suppose $X$ has the probability density function $\mathcal{f_{high}}(x)$. Then for any two distances $a$ and $b$ with $0 \leq a \leq b$,

$$ P(a \leq X \leq b) = \int_{a}^{b} \mathcal{f_{high}}(x) dx. $$ 

In particular, the probability that the 8-hour straight-line spread is at least 10 km is given by

$$ P(X \geq 10) = \int_{10}^{\infty} \mathcal{f_{high}}(x) dx, $$ 

and in this problem, $\mathcal{f_{high}}(x)$ is the density of a lognormal distribution. 

___

Task instructions

1. Treat \(X\) as the 8-hour distance (km) for a new ignition in a High-hazard area.

2. Model \(X\) with a lognormal distribution using the **High** row’s **Mean log** and **SD log** from **Table 1**.

3. Use `plnorm()` with `lower.tail = FALSE` to compute  
  \( P(X \geq 10) = P(X > 10) \).

4.  Store the result as **`p_high_exact`**.

5.  Check both the raw probability and the corresponding percentage.

🎯 **Goal.**  
In your own words, state what you are finding the probability of, including time and distance and model type.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Test: lognormal tail probability for High category
____  # store p_high_exact, use plnorm(), 10 km threshold, High Mean log, High SD log

```

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ✅  Verify: eexamine probability and percentage
____ # display p_high_exact
____ # convert p_high_exact to a percentage
```

🗣 Comment (TODO).
Interpret what your probability means for High-hazard ignitions.

## Task 2

`qlnorm()` computes the quantile function (inverse CDF) for a lognormal distribution. It gives the value below which a certain percentage of the distribution falls.

❓ Question b: We’ll treat distances above the 95th percentile as ‘extreme fast runs’. For ignitions in High-hazard areas, what 8-hour distance is only exceeded 5% of the time? In other words, what is the 95th percentile run length?

___
  
Let $X$ be the continuous random variable giving the 8-hour straight-line spread distance (in km) of a new ignition in a High-hazard area. Suppose $X$ has the probability density function $\mathcal{f_{high}}(x)$. The cumulative distribution function (CDF) of $X$ is given by

$$ F(x) = P(X \leq x) = \int_{0}^{x} \mathcal{f_{high}}(t) dt. $$

We define the 95th percentile of $X$, denoted by $q_{0.95}$ of this distribution to be the value satisfying

$$ P(X \leq q_{0.95}) = F(q_{0.95}) = 0.95. $$

equivalently, 

$$
\int_{0}^{q_{0.95}} \mathcal{f_{high}}(x) dx = 0.95
$$

In particular, the 95th percentile of the 8-hour straight-line spread distance for new ignitions in High-hazard areas is given by 

$$ q_{0.95} = F^{-1}(0.95), $$ 

and in this problem $\mathcal{f_{high}}(x)$ is the density of a lognormal distribution.

___

**Task instructions**

1. Continue modeling the High-hazard 8-hour distance \(X\) with a lognormal distribution.

2. Use `qlnorm()` to find the 95th percentile distance (km) for \(X\).

3. Use the High row’s Mean log and SD log from Table 1.

4. Store the result as `q95_high`.

5. Check that this distance is plausible given the High category’s mean and SD.

🎯 Goal.
Write your own sentence(s) goal for finding a distance that 95% of High-hazard ignitions stay below.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Test: 95th percentile distance for High category
____ # store q95_high, use qlnorm() with p = 0.95, High Mean log, High SD log
```

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ✅  Verify: print and sanity-check the 95th percentile
____ # display q95_high
```

🗣 Comment (TODO).
Explain in words what this percentile distance means in the context of High-hazard ignitions.

## Task 3

`dlnorm()` computes the probability density function (PDF) for a lognormal distribution. It gives the relative likelihood of the random variable taking on a specific value.

❓ Question c: Under our High-hazard lognormal model, how concentrated is the distribution locally around the shorter 3 km run versus the longer 10 km distance in 8 hours? _Note: Higher density indicates a higher likelihood of the value in a small neighborhood around that point._

___

Let $X$ be the continuous random variable giving the 8-hour straight-line spread distance (in km) of a new ignition in a High-hazard area. Suppose $X$ has the probability density function $\mathcal{f_{high}}(x)$. Then for any two distances $a$ and $b$ with $0 \leq a \leq b$,

$$ P(a \leq X \leq b) = \int_{a}^{b} \mathcal{f_{high}}(x) dx. $$

The values $\mathcal{f_{high}}(x)$ are densities, not probabilities. They represent the height of the distribution at the point x. In particular, the density at 3 km and 10 km are given by

$$ \mathcal{f_{high}}(3) \text{ and } \mathcal{f_{high}}(10). $$

___

**Task instructions**

1. Under the High-hazard lognormal model, use `dlnorm()` to compute the density at 3 km and 10 km.

2. Use the High row’s Mean log and SD log from Table 1.

3. Store the results as `d_3` and `d_10`.

4. Compare `d_3` and `d_10` to decide which distance is more locally likely.

🎯 Goal.
Write your own sentence(s) goal for comparing how locally likely 3 km vs 10 km runs are under the High-hazard model.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Test: compare densities at 3 km and 10 km
____ # compute d_3 using dlnorm() at 3 km with High Mean log, High SD log
____ # compute d_10 using dlnorm() at 10 km with High Mean log, High SD log

```

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ✅ Checks: print and compare heights

____ # display d_3
____ # display d_10
```

🗣 Comment (TODO).
Use your two density values to describe which distance is more locally likely and what that implies.

## Task 4

`rlnorm()` generates random numbers from a lognormal distribution. **Note**: This task has three parts.

❓ Question d: If we simulate 450 new ignitions in High-hazard areas this season, about how many would be classified as ‘fast-growing’ (≥10 km in 8 hours)?

___

Let $X$ be the continuous random variable giving the 8-hour straight-line spread distance (in km) of a new ignition in a High-hazard area. Assume

$$ X \sim \text{Lognormal}(\mu = \text{meanlog High}, \sigma = \text{ sdlog  High}), $$

with density function $\mathcal{f_{high}}(x)$ and CDF $F_{high}(x)$. 

The probability that a single new ignition is 'fast-growing' (i.e., spreads 10 km or more in 8 hours) is given by

$$ P(X \geq 10) = 1 - F_{high}(10)$$

Now suppose we observe (or simulate) $n = 450$ new ignitions in High-hazard areas this season. Let $X_1, X_2, \ldots, X_{450}$ be random variables, each with the same lognormal distribution as $X$. 

Define indicator variables for 'fast-growing' ignitions:

$$ I_i = \begin{cases} 1 & \text{if } X_i \geq 10 \\ 0 & \text{if } X_i < 10 \end{cases} \quad i = 1,\dots,450.$$

Then, the total number of 'fast-growing' ignitions, $S$, is given by

$$ S = \sum_{i=1}^{450} I_i $$

Each $I_i$ is a Bernoulli random variable with success probability $P(I_i = 1) = P(X_i \geq 10) = P(X \geq 10),$. Therefore, the **theoretical expected number** of 'fast-growing' ignitions is

$$ \mathbb{E}[S] = \sum_{i=1}^{450} \mathbb{E}[I_i] = 450 \cdot P(X \geq 10). $$

___

In Tasks 4.1 and 4.2, you will simulate one season of 450 ignitions and count how many simulated ignitions are fast-growing. This simulated count may differ from the theoretical expected value.

### Task 4.1

**Simulate High-hazard distances**

Task instructions

1. Use `rlnorm()` to simulate 450 independent 8-hour distances for ignitions in the High category.

2. Use the High row’s Mean log and SD log from Table 1.

3. Set a seed with `set.seed(2025)` for reproducibility.

4. Store the simulated distances in a vector `High_Lognorm`.

5. Summarize the simulated distances to understand the range, center, and skew of this simulated season’s distances.

🎯 Goal.
Write your own sentence(s) goal for Task 4.1.

```{r eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Test: simulate 8-hour distances for High category
____ # set a seed for reproducibility
____ # store High_Lognorm, use rlnorm() to generate 450 values with High Mean log, High SD log
```

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ✅  Verify: look at basic summaries
____ # summarize High_Lognorm (e.g., summary())
____ # confirm the length of High_Lognorm is 450)
```

🗣 Comment (TODO).
Describe briefly how the simulated distances behave (range, skew, center).

### Task 4.2

**Count fast-growing ignitions** (≥10 km)

**Task instructions**

1. Define a “fast-growing” ignition as one with an 8-hour distance ≥ 10 km.

2. Use a logical condition on `High_Lognorm` to identify fast-growing runs.

3. Use `sum()` on that logical vector to count how many simulated ignitions are fast-growing.

4. Store the count as `fast_growing_count`.

5. Optionally, compute the simulated proportion of fast-growing ignitions and compare it to `p_high_exact` from Task 1.

🎯 Goal.
Write your own sentence(s) goal for Task 4.2.

```{r eval=FALSE, message=FALSE, warning=FALSE}
# ⚡ Test: count fast-growing ignitions (≥ 10 km)
____ # store as fast_growing_count,use sum() to count TRUEs of logical vector 
```

```{r eval=FALSE, message=FALSE, warning=FALSE}
# ✅  Verify: inspect count (and optional proportion)

____ # display fast_growing_count
____ # optionally compute the proportion fast_growing_count / length(High_Lognorm)
```

🗣 Comment (TODO).
Explain what your simulated count suggests about how many fast-growing events you might see in a season.

### Task 4.3

**Visualize simulated distances**

**Task instructions**

1. Use `hist()` to create a histogram of `High_Lognorm`.

2. Choose a reasonable bin setting and axis labels.

3. Use the plot to comment on the distribution shape (e.g., skew, tail length).

4. Connect the shape of the histogram to your numeric summaries from Task 4.1.

🎯 Goal.
Write your own sentence(s) goal for visualizing the simulated High-hazard distances.


```{r, eval=FALSE, message=FALSE, warning=FALSE}
# ⚡/✅ Code histogram of simulated High_Lognorm distances
____ # call hist() on High_Lognorm, choosing breaks/title/xlab as you see fit
```

🗣 Comment (TODO).
In one sentence, relate the histogram’s shape to your earlier summaries and to what you expect from a lognormal model.


# Part C

`rlnorm` + `replicate()` 

❓ Question e: In a season with 450 new ignitions **statewide**, what is the probability we see at least 50 fast-growing events (≥10 km in 8 hours) when ignitions can occur in any hazard category according to their statewide area weights?

Let there be $L$ hazard categories indexed by $i = 1, 2, \ldots, L$ with:

- statewide area weights $p_j$ such that $\sum_{j=1}^{L} p_j = 1$.

- lognormal spread model in category $j$: 

$ X | \text{category} = j \sim \text{Lognormal}(\text{meanlog}_j, \text{sdlog}_j). $

Consider a season with $N = 450$ new ignitions **statewide**. For each ignition $i = 1, 2, \ldots, N$:

1. Draw a hazard category 

$$ C_i \in \{1, 2, \ldots, L\}, \space \space P(C_i = j) = p_j. $$

independently across ignitions.

2. Conditional on $C_i = j$, draw an 8-hour straight-line spread distance $X_i$ as

$$ X_i | C_i = j \sim \text{Lognormal}(\text{meanlog}_j, \text{sdlog}_j). $$

independently across ignition given their categories.

Define the indicator variable for a 'fast-growing' ignition (≥ $k = 10$ km in 8 hours) as

$$ I_i = \begin{cases} 1 & \text{if } X_i \geq k \\ 0 & \text{if } X_i < k \end{cases} \quad i = 1,\dots,450.$$

Then, the total number of 'fast-growing' ignitions in the season, $S$, is given by

$$ T = \sum_{i=1}^{450} I_i $$

We are interested in the probability that a season has at least 5 fast-growing events:

$$ P(T \geq 50). $$

and is approximated by Monte Carlo simulation.

$$
\widehat{P(T \ge 50)}
\;\approx\;
\frac{1}{S} \sum_{s=1}^{S} \mathbf{1}\{T^{(s)} \ge 50\},
$$

where $T^{(s)}$ is the simulated count of fast-growing ignitions in season $s$, and $S$ is the total number of simulated seasons.


## Task 1

**Create category labels**

**Task instructions**

1. Using **Table 1**, create a character vector named **`labels`** that contains the seven wildfire hazard categories in the same order as the **Category** column.

2. Make sure the spelling and capitalization of each label exactly matches Table 1.

3. Check that the vector prints correctly.

🎯 **Goal.**  
Write your own sentence(s) goal for creating a vector of category labels.

```{r, eval=FALSE}
# ⚡ Test: vector of category labels
____  # create vector <- c( ... ) using the Category names from Table 1
```

```{r, eval=FALSE}
# ✅  Verify: inspect labels

____ # display labels

```

🗣 Comment (TODO).
Explain briefly why matching the order and spelling of these labels will matter later.


## Task 2

**Create category probabilities**

**Task instructions**

1. Using the Probability column in Table 1 (as proportions that sum to 1), create a numeric vector **`prob`**.

2. The entries of `prob` must be in the same order as `labels`.

3. Check that the probabilities sum to 1 (within rounding error).

🎯 Goal.
Write your own sentence(s) goal for building vector.


```{r, eval=FALSE}
# ⚡ Test: vector of category probabilities
____ # create vector <- c( ... ) using the Probability column from Table 1

```

```{r, eval=FALSE}
# ✅  Verify: : confirm they form a valid distribution
____ # display vector
____ # check that sum() is 1
```


## Task 3

**Create lognormal parameter vectors**

**Task instructions**

1. From Table 1, read off the Mean log values for each category in the same order as `labels` and store them in a vector **`meanlog`**.

2. Similarly, read off the SD log values and store them in a vector **`sdlog`**.

3. Check that both vectors line up with **`labels`** (same length, same order).

🎯 Goal.
Write your own sentence(s) goal for vectors.

```{r, eval=FALSE}
# ⚡ Test: vectors of lognormal parameters
____ # create vector 1 <- c( ...) from the Mean log column
____ # create vector 2 <- c( ...) from the SD log column
```

```{r, eval=FALSE}
# ✅  Verify: inspect and compare
____ # display vector 1
____ # display vector 2
____ # check length of labels vector, with vector 1 and 2.
```

🗣 Comment (TODO).
Describe any pattern you see.

## Task 4

**Set global simulation constants**

Task instructions

1. Let **`N`** be the number of ignitions per season (use the value given in the problem statement).

2. Choose a large number **`S`** of simulated seasons (as specified in the assignment).

3. Set the “fast-growing” threshold **`k`** in km (use the value given in the problem statement).

4. Store these three values in objects `N`, `S`, and `k`.

🎯 Goal.
Write your own sentence(s) goal for setting up the core constants. 

```{r, eval=FALSE}
# ⚡ Test: number of ignitions, seasons, and fast threshold
____ # set N to the number of ignitions per season 
____ # set S to the number of simulated seasons
____ # set k to the fast-run distance threshold in km 
```

```{r, eval=FALSE}
# ✅  Verify: inspect constants
____ # display N
____ # display S
____ # display k
```


## Task 5

**Sample categories for one season**

**Task instructions**

1. Simulate one season of ignitions by drawing `N` categories according to their probabilities.

2. Use `sample.int()` (or `sample()`) to draw indices from `1:length(labels)` with replacement and probability vector `prob`.

3. Store the category indices in a vector named **`idx`**.

4. Use `table()` with `labels[idx]` to see how many ignitions fall into each category.

🎯 Goal.
Write your own sentence(s) goal for simulating a single season’s mix of categories.

```{r, eval=FALSE}
# ⚡ Test: sample category indices for one season
____ # create idx using sample.int() (or sample()) with size = N, prob = prob
```

```{r, eval=FALSE}
# ✅  Verify: examine realized category mix

____ # inspect the first few indices
____ # use table() to see counts per category
```

🗣 Comment (TODO).
Compare your simulated category counts to the theoretical proportions from prob.

## Task 6

**Draw distances for one season**

**Task instructions**

1. For the same season, convert each category index in `idx` to an 8-hour distance using a lognormal model.

2. Use `rlnorm()` to draw `N` distances.

3. Use `meanlog[idx]` and `sdlog[idx]` so that each ignition uses the parameters for its own category.

4. Store the distances in a vector **`d`**.

5. Inspect the first few values and a summary of `d`.

🎯 Goal.
Write your own sentence(s) goal for turning category indices into distances for one season.

```{r, eval=FALSE}
# ⚡ Test: simulate distances for one season
____ # store d, use rlnorm() with n = N, meanlog = meanlog[idx], sdlog = sdlog[idx]
```
```{r, eval=FALSE}
# ✅  Verify: examine distances

____ # show head()
____ # summarize d 
```

🗣 Comment (TODO).
Describe what you notice about the distribution of distances across all categories.

## Task 7

**Count fast runs for one season**

**Task instructions**

1. Using the vector `d`, mark which ignitions are “fast-growing” using the condition `d >= k`.

2. Store the logical vector in **`fast_flags_one`**.

3. Use `sum()` on `fast_flags_one` to count fast ignitions and store the result as **`fast_count_one`**.

4. Compute `mean()` of `fast_flags_one` to get the fraction of fast ignitions.

🎯 Goal.
Write your own sentence(s) goal for counting fast-growing ignitions.

```{r, eval=FALSE}
# ⚡ Test: count fast runs in one season
____ # create fast_flags_one as a logical vector using d >= k
____ # compute fast_count_one using sum() of fast_flags_one
```
```{r, eval=FALSE}
# ✅  Verify: inspect count and proportion
____ # display fast_count_one
____ # optionally compute mean() of fast_flags_one to get proportion

```

🗣 Comment (TODO).
Comment on how many fast runs you saw in this particular simulated season.

## Task 8

**Bundle one-season logic in a single block**

Braces `{ ... }` group several lines together so they act as one expression and return the final result.

**Task instructions**

1. Combine the one-season computation into a single block that:

- samples `idx` for one season,

- simulates distances `d`,

- returns the count of fast ignitions.

2. Wrap the block in braces `{}` and assign the result to **`fast_count_one`**.

3. Run the block and confirm it returns a single number (the fast count).

🎯 Goal.
Write your own sentence(s) goal for packaging the one-season simulation into one code block.

```{r, eval=FALSE}
# ⚡ Test: bundled one-season simulation returning a single fast count
____ <- { #store vector and start block
____ # sample idx for one season
____ # simulate distances d using meanlog, sdlog with brackets and idx
____ # return the count of d values that are greater than or equal to k
} #end block
```
```{r, eval=FALSE}
# ✅  Verify: confirm the block works
____ # display fast_count_one
```

🗣 Comment (TODO).
Explain why having this one-season block is useful when you move to many seasons. 


Task 8 might have been challenging! Consider other kinds of memos: 🪞 Reflect; 🧠 Growth mindset; 👀 Cite resources; 🚀 Code trials; ✍️ Record what you tried.

## Task 9

**Use `replicate()` to simulate many seasons**

**Task instructions**

1. Use `replicate()` to repeat your one-season block `S` times.

2. Store the resulting vector of counts in **`counts_fast_mix`**.

3. Compute the expected number of fast runs per season as the mean of `counts_fast_mix`.

4. Estimate the probability that a season has at least 50 fast-growing events by computing the fraction of seasons where the count is ≥ 50.

🎯 Goal.
Write your own sentence(s) goal for simulating many seasons and estimating the chance of at least 50 fast fires.

```{r, eval=FALSE}
# ⚡ Test: many-season simulation with replicate()
set.seed(2025)
____ # store resulting vector, use replicate() 
____ # with n = number of seasons
____ # and expr = { ... one-season block ... } as before
)
```
```{r, eval=FALSE}
# ⚡ Test:  Answer Question
# summarize seasonal fast counts and estimate probabilities
____ # inspect summary() of counts_fast_mix
____ # compute mean() for expected fast runs
____ # compute mean() for Monte Carlo estimate of P(T >= 50)  
```

🗣 Comment (TODO).
Interpret both the expected fast count and the estimated probability of at least 50 fast fires in words.

Task 9 might have been challenging! Consider other kinds of memos: 🪞 Reflect; 🧠 Growth mindset; 👀 Cite resources; 🚀 Code trials; ✍️ Record what you tried.

## Task 10

**Visualize the distribution of seasonal fast counts**

**Task instructions**

1. Use `ggplot2::geom_histogram()` (or base `hist()`) to plot a histogram of `counts_fast_mix`.

2. If you use `ggplot2`, you could first build a small data frame with one column (`fast_count`).

3. Clearly label the axes and choose a reasonable bin width.

4. Draw a vertical line at 50 to mark the threshold and visually highlight seasons with ≥ 50 fast fires.

5. Relate the shape of the histogram to your numeric summaries from Task 9.

🎯 Goal.
Write your own sentence(s) goal for plotting the distribution of fast fires per season.

```{r, eval=FALSE}
# ⚡ Test: histogram of seasonal fast counts

____ # load ggplot2 if you choose to use it
____ # create a data frame with a column fast_count = counts_fast_mix

____ # build a histogram of fast_count (e.g., ggplot(...) + geom_histogram(...))
____ # Add a vertical line at the threshold of 50 fast fires
``` 

```{r, eval=FALSE}
# ✅  Verify: inspect histogram
____ # display the histogram
```

🗣 Comment (TODO).
In one sentence, explain how the histogram supports (or challenges) your numeric estimate of the probability of ≥ 50 fast fires.


# Save and Upload

Submit your (1) completed assignment notebook `QMD` (2) rendered `HTML` and (3) environment (`RData`) to appropriate Canvas pages, as directed. 

