---
title: "Is AI Appropriate? Group Discussion"
author: "Lorraine Gaudio"
date:   "`r paste(format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In this activity, you will learn how to make a defensible decision about when AI help supports your learning and when it replaces the learning work you need to do to build coding skill. In R, getting “working code” is not the same thing as understanding why it works and AI can easily create the illusion of competence if you copy solutions. This matters because your long-term ability to work in data analytics depend on you being able to explain and reproduce your work. You will practice a “tenacity-first” approach: attempt first, document what happened, then seek help (including AI) in a way that keeps you accountable for understanding and verification.




# AI, Tenacity, and Human-Centered Learning

In this course, you are the learner and decision-maker. AI is useful, fast, and sometimes wrong. A human-centered approach means:

- You stay accountable for understanding what the code is doing and for verifying results.

- AI supports your thinking instead of replacing it.

- Learning is the product. Getting “working code” is not the same as learning how to produce it.

When students use AI tools, they often engage in cognitive delegation: transferring mental labor (planning, interpreting, writing, debugging) to an artificial system. Delegation is not automatically bad. It becomes a problem when it prevents you from doing the mental work that builds skill.

Tenacity and AI belong together: tenacity tells you when to persist, when to recalibrate, and when to seek help. AI is one possible form of help if you use it in a way that keeps the learning with you.

## Tenacity-First AI rule

**Before you ask AI**, do enough work to ask a *good* question.

It’s appropriate to ask AI when you can provide:

1. 🎯 Your goal (what you were trying to do)

2. ⚡ The exact code you ran.

3. ⚠️ The exact error message/output.

4. 🧐 What you understand and do not understand.

5. 🛠 What you tried to gain understanding (e.g., reading the error message, checking intermediate objects, searching documentation).

If you can’t provide those details, you don’t have enough information yet to ask for help (AI or human). Do one more round of work first.

AI use is not appropriate (1) when you are using it to avoid thinking, (2) avoid experimenting yourself, or (3) avoid reading and interpreting the error message.

**After AI**, you must verify. 

- rerun from the top of the script (or restart and rerun),

- print or check the intermediate objects,

- compare the AI suggestion with the chapter content,

- explain your understanding in your own words

## AI Help and Harm

**AI is usually helpful when you use it as:**

- A translator (turning an error message into plain English)

- A debugging coach (suggesting systematic checks)

- A concept tutor (explaining why something works)

- A practice generator (creating additional examples for you to try)

- A prompt for reflection (asking you questions that help you reason)

**AI is usually harmful when you use it as:**

- A replacement for your attempt

- A copy-paste generator you run without understanding

- A shortcut around practice

- A confidence crutch (“I can’t do this, let the AI do it”)

## Is AI appropriate here?

### Vignette 1: 

**“Object not found” after running code out of order**

    While solving a Practice Space task, Maya runs `the_aftermath <- seven_eight - nine` and gets: 
    Error: object 'seven_eight' not found. She feels embarrassed and immediately asks AI: “Fix my code.”

**Is AI appropriate?**

AI is only appropriate after Maya works through the Tenacity-First steps. This error is relatively straightforward. Tenacity-First means she first reads the error and asks: “Did I create `seven_eight` in this session?” If she doesn’t know, she checks the Environment or runs the earlier lines. If she’s still confused, AI is useful as a translator: “What does ‘object not found’ mean and what should I check first?”

**Good AI use**: coaching her through a checklist.
**Bad AI use**: having AI rewrite everything without her learning the pattern.

### Vignette 2:

**“Write my whole script for me”**

    Jordan is tired. He asks AI: “Write the entire Module 1 script that creates objects, 
    saves them, and prints results. Make it perfect.”

**Is AI appropriate?**

No. This is cognitive delegation that removes the core learning: typing code, running it, fixing errors, and noticing how objects appear in the Environment. It prevents skill-building and produces fragile learning. You’ll have code you can’t reproduce. Tenacity means doing the uncomfortable part: building the script yourself, line-by-line, and debugging.

### Vignette 3:

**Not getting anywhere**

    Sam tries to install a package. It fails. He restarts RStudio, tries again, still fails.
    He’s repeating the same steps and getting the same error.
    He asks AI: “Here’s the error message ... and what I tried ... 
    Give me a structured debugging plan.”

**Is AI appropriate?**

Yes. This is exactly what tenacity recommends: persist, then recalibrate. Sam has evidence, not guesses. AI can propose a decision tree (“Check internet permissions → check R version → try installing from a different repo → read the full error → restart session”) and explain the logic. Sam still must test each step and confirm results. AI is acting as a troubleshooting guide, not a replacement brain.

___

Use AI when it keeps you thinking, speeds up understanding, and helps you debug systematically. Avoid AI when it replaces the struggle that produces learning. Tenacity is not “do everything alone.” Tenacity is “do the learning work, then seek the right help in the right way.”

# Is AI Appropriate? Discussion

You will complete four short tasks. Work silently first, then talk with your group. Your goal is to produce clear, specific decisions.

**How to work**

1. **Silent work** (5 minutes): Quietly complete Tasks 1–2 on your own notes.

2. **Group discussion**: Share answers, challenge vague claims, and reach a group consensus.

3. **Write up**: Each person will paste their Task 3 vignette into the Canvas assignment text box. Keep it brief and follow the template. *This is a participation check, not an essay.*

## Task 1

**Sorting 5 micro-scenarios**

Below are five micro-scenarios. Consider each one and decide:

      - Is AI is appropriate? yes/no/depends; 
      - What should the student do first; 
      - How should the student use the AI tool, and 
      - How should they verify the result.

1. “I paste my error + code + what I tried and ask AI for a checklist.”

2. “I ask AI to provide feedback on my R Script after I completed it independently. I plan to revise based on feedback."

3. “I ask AI what seq(from=0,to=10,by=1) returns and why. I paste the explanation into my memo.”

4. “I ask AI to generate extra practice problems.”

5. “I ask AI to write the script for Chapter 2.”

**Group goal**: After discussion, your group should be able to explain why each item belongs in its category.

## Task 2

Answer one or both of the following vignettes. Include one reason and one condition/limit. 

### **Vignette 4:**

**“Explain what my output means”**

    While working through a guided activity (Chapter reading), 
    Priya runs code that prints a number, 
    and she doesn’t understand why it’s that number. 
    She asks AI: 
    “Explain this output step-by-step using my variable names.”

**Is AI appropriate?** Write a couple sentences explaining your answer. 

```{r eval=FALSE, include=FALSE}
# Yes, with one condition.
# This is AI as a tutor. It can be useful for Priya to actively check the explanation by rerunning smaller “code trials” (printing intermediate objects and confirming each step). If she accepts the explanation without verifying, she’s delegating understanding. Tenacity means "trust, but verify."
```

### **Vignette 5:** 

**"Write my reflection"**

    Devin hits three errors in a row and feels discouraged. 
    After using AI to help him understand and solve the errors, he asks AI: 
    “Write my memo about what I learned and how I showed tenacity.”

**Is AI appropriate?** Write a couple sentences explaining your answer. 

```{r eval=FALSE, include=FALSE}
# Mostly no.
# The memo is where your learning becomes visible to you. Outsourcing it removes the reflection that turns confusion into growth. Tenacity includes emotionally appropriate persistence: naming what happened, what you tried, what worked, and what you’ll do next time.
# What is appropriate: asking AI to generate reflection prompts (questions) you answer yourself, like: “What did the error message literally say? What did you change? What did you learn about object creation and order?”
```

**Group goal**: Reach a consensus that includes a clear condition/limit and a non-negotiable verification step.

## Task 3

**Write your own vignette**

Write a short vignette from your own experience (or invent one) using the template above. Then trade with a partner and decide whether AI is appropriate, including one reason and one condition/limit. Compare your decisions and revise your condition/limit until it’s specific enough that a stranger could follow it.

Vignette template

    Name: [Name of person in vignette] 
    
    Context: What task was [Name] doing?

    Action: Provide code snippet 1–3 lines or describe the action if not code.

    Problem: State the error message or result

    Next Action: Did [Name] try something before asking for help

    AI Prompt: How does [Name] engage with AI?

    Evaluate: AI appropriate? (Yes/No/Depends/Unsure)

    Reason: Explain your evaluation
    
    Verification plan: How could [Name] check the AI output?
    
**Group goal**: Revise your condition/limit until it is specific enough that a stranger could follow it.   



---

# Wrap-Up: What You Should Take With You

The point of this activity is not to decide whether AI is “good” or “bad.” The point is to practice **responsible help-seeking** while you learn to code. In this course, AI is acceptable when it helps you *think* and it is not acceptable when it replaces the work that builds skill.

Going forward, use this rule-of-thumb: **attempt → document → ask → verify.** If you can’t show what you tried, you’re not ready to ask for help yet. If you do ask for help, you’re still responsible for verifying the result by rerunning code, checking intermediate objects, and explaining what changed in your own words. This is how you build confidence that lasts beyond this class.

