---
title: "Lesson 13"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Lesson 13"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
 
Or
title: "Lesson 13"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
team: "Fall 2025"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```


# (maybe a dice icon?) Welcome Back to R! 

In lesson 12, we did simple data visualizations using R’s built‑in function `hist()`. Today, we use the package `ggplot2` to turn data frames into clear, publication‑quality plots. Specificaly, we'll look at the grammar of graphics.

To begin Lesson 13, follow these steps:

1. Open your course project for RStudio 
   
2. Create a new file. From the file types we have used so far, pick which file type you want to use. (File > New File > ???).

3. Type in the code provided in this document as you follow along with the video. Pause the video at anytime to answer assignment questions, dig deeper or add memo notes.

**Lesson Overview**

By the end of Lesson 13 you will be able to:

1. ⚡ Remember – State the steps: `data` ➜ `aes()` ➜ `geom()`.

2. 🔍 Understand – Describe how `aes()` maps variables to plot elements.

3. 🛠 Apply – Build scatterplots, histograms, and boxplots.

4. 📊 Analyze –  Use color or fill aesthetics to compare groups.

5. 🧠️ Evaluate – Choose suitable geoms and themes for a question.

Keep these goals in mind as you move through each section.

# 📦 Packages

Install once (if needed): `install.packages("dplyr")`; `install.packages("dslabs")`

```{r , eval=FALSE}
install.packages("ggplot2")
```

Load the packages at the start of every session:

```{r , eval=FALSE}
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
library(ggplot2) # Data visualization
```

# Warm‑Up

What is ggplot2?

```{r , eval=FALSE}
?ggplot   # skim the Details section
```

📜 The SYNTAX (basic/general): `ggplot(data = NULL, mapping = aes(), ..., environment = parent.frame())`

🧐 NOTICE: Which argument supplies mapping information?

🧩 Explanation: In ggplot2, mapping refers to connecting your data variables to visual properties (aesthetics) of the plot. It's how you tell ggplot which column in your data should control which aspect of the visualization.

# The Blank Canvas

Start every plot by piping a data frame into `ggplot()`.


`df %>% ggplot()`

OR without piping

`ggplot(data = df)`


```{r , eval=FALSE}
divorce_margarine %>%
  ggplot()
```

The result is empty — we still need aesthetic mappings and a geom.

# Mapping Aesthetics

```{r , eval=FALSE}
?aes
```

📜 The SYNTAX (basic): `⁠ggplot(data = df, mapping = aes(x, y, other aesthetics))`

OR with piping

`df %>% ggplot(mapping = aes(x, y, other aesthetics))`

🎯 The GOAL: Map divorce rate to x and margarine consumption to y.

```{r , eval=FALSE}
divorce_margarine %>%
  ggplot(aes(x = divorce_rate_maine, y = margarine_consumption_per_capita))
```

# Scatterplot

`geom_point()` is a geometric object (or "geom") in ggplot2 that creates a scatter plot by placing points at the specified x and y coordinates.

```{r , eval=FALSE}
divorce_margarine %>%
  ggplot(aes(divorce_rate_maine, margarine_consumption_per_capita)) +
  geom_point()
```

🧩 Explanation: `geom_point()` understands these aesthetics:

-  x, y: Position (required)

-  alpha: Transparency

-  color: Point color

-  fill: Fill color (for shapes with fill)

-  shape: Point shape (0-25)

-  size: Point size

-  stroke: Border thickness

```{r , eval=FALSE}
divorce_margarine %>%
  ggplot(aes(x = divorce_rate_maine, 
             y = margarine_consumption_per_capita)) +
  geom_point(aes(color = year,         # Color points by year
                 size = divorce_rate_maine, # Size points by divorce rate
                 alpha = margarine_consumption_per_capita), # Transparency by consumption
             shape = 16)      
```


# Histogram 

```{r , eval=FALSE}
?geom_histogram
```

🎯 The GOAL: Build a small data frame of simulated heights.

```{r , eval=FALSE}
set.seed(2025)
Tallness <- data.frame(
  Country = c(rep("USA", 1000), rep("CAN", 1000)),
  Height  = c(rnorm(1000, 67, 2.5), rnorm(1000, 66, 1.5))
)
```

🧐 NOTICE: that the first country, "USA" will have a mean height of 67 inches

```{r , eval=FALSE}
Tallness %>%
  ggplot(aes(Height)) +                # only x for histograms
  geom_histogram(fill = "steelblue", color = "white")
```

`geom_histogram()` understands these aesthetics:

-  x: Position (required)

-  fill: Fill color

-  color: Border color

-  binwidth: Width of each bin (default is 1)

-  position: Position adjustment (e.g., "identity", "dodge", "fill")

-  alpha: Transparency

-  size: Border thickness

-  weight: Weight for each observation (optional)

-  linetype: Line type for borders (optional)

```{r , eval=FALSE}
Tallness %>%
  ggplot(aes(Height)) +                # only x for histograms
  geom_histogram(fill = "steelblue", 
                 color = "white",
                 binwidth = 0.5)       # Set bin width to 0.5
```

🚀 Explore and Play: Change `binwidth = 0.25` inside `geom_histogram()`. How does the distribution look?

🛠 Break Things! Test out your hand at adding aesthetics `geom_histogram()`  understands.

# Group Comparison

🎯 The GOAL: Overlay two histograms by mapping fill to Country. 

```{r , eval=FALSE}
View(Tallness)  # Recall that there are two countries: USA and CAN
```


```{r , eval=FALSE}
Tallness %>%
  ggplot(aes(Height, fill = Country)) +
  geom_histogram(position = "identity", alpha = 0.6)
```

🧩 Explanation: 

1. `position = "identity"` lets bars overlap

2. `alpha` controls transparency: Sets the transparency of all bars to 0.6 (0 == transparent, 1 == opaque) 

# Boxplot

```{r , eval=FALSE}
?geom_boxplot
```

```{r , eval=FALSE}
Tallness %>%
  ggplot(aes(Country, Height, fill = Country)) +
  geom_boxplot(show.legend = FALSE)
```

👀 Check‑in: Which country shows greater height variability?

# Labels and Themes


```{r , eval=FALSE}
Tallness %>%
  ggplot(aes(Country, Height, fill = Country)) +
  geom_boxplot(show.legend = FALSE) +
  xlab("North American Country") +
  ylab("Height (inches)") +
  ggplot2::theme_minimal()
```

🧩 Explanation: 

- `xlab() and `ylab()` set axis labels.

- `ggplot2::theme_minimal()` applies a clean, minimal theme to the plot.

- `ggplot2::` prefix explicitly specifies that you're using the `theme_minimal()` function from the `ggplot2` package. 

This is useful when multiple packages have functions with the same name. 

# 📝 Practice Space

✍️ Practice & 🚀 Explore and Play: 

  1. Build a density plot of the `heights` dataset (`dslabs`). Shade by `sex`.

  2. Try `geom_smooth()` on `divorce_margarine` to add a trend line.

  3. Use `iris` data to create side‑by‑side boxplots of `Sepal.Width` by `Species` and change the theme.

Add your code below and include at least one 👀 check‑in to predict an outcome.

```{r , eval=FALSE}


```

# ➤ Assignment

Replace each ____ placeholder (and any TODO comments) with working code or a short written answer. Run each section; be sure the requested objects appear in the Environment. When finished, save **BOTH** this script and your .RData workspace and upload.

When you’re done, your workspace should contain FOUR new objects: `Pretty_Flowers`, `Petal_Spread`, `CO2_Graph`, `Tall_Density`

## Task 0

📦 Packages

Attach `ggplot2`, `dplyr`, and `dslabs`

```{r, eval=FALSE}
____(____) 
____(____)
____(____)
```

## Task 1

🧠 Quick Recall

List the three core `ggplot2` steps

✍️ EXPLANATION: "___"

✍️ EXPLANATION: "___"

✍️ EXPLANATION: "___"

## Task 2

Scatterplot 

Build **`Pretty_Flowers`**: `iris` data ➜ `Sepal.Length` vs `Petal.Length,` color points by `Species`, `size = 3`, add labels for both axes.

Hint: use `ggplot2` pipe

```{r, eval=FALSE}
_____ <- _____

```

```{r, eval=FALSE}
print(Pretty_Flowers)
```

## Task 3

Boxplot: group comparison

Create **`Petal_Spread`**: `iris` ➜ `Petal.Width` by `Species` as green boxplots (no legend).  Use `fill = Species`, then `show.legend = FALSE`.

```{r, eval=FALSE}
____ <- _____
```

## Task 4

Density plot: split + transparency

Use CO2 data to make **CO2_Graph**:

- x = uptake

- fill by Type

- alpha = 0.5

- title = "CO2 Uptake by Plant Type"

```{r, eval=FALSE}
_____ <- _____ %>% 
  ggplot(_____(uptake, fill = _____)) +
  _____ +
  _____(title = "CO2 Uptake by Plant Type")
```

## Task 5

⚖️ Histogram vs. Density 

Build **Tall_Density** on heights (dslabs):

- map height only (x)

- add geom_histogram(binwidth = 1, fill = "grey70", color = "white")

- add geom_density(color = "red", size = 1)

- x-lab "Height (inches)"

```{r, eval=FALSE}
Tall_Density <- heights %>% 
  ggplot(_____(height)) +
  _____ +
  _____ +
  _____("Height (inches)")
```

## Task 6

🤔 Reflect: 

📝 Write a short paragraph reflecting on when would a density curve be more useful than a histogram?

✍️ EXPLANATION: "___"

# Save and Upload

1. You will be submitting **both** the Quarto Document and the workspace file. The workspace file saves all the objects in your environment that you created in this lesson. You can save the workspace by running the following command in a code chunk of the Quarto Document document:

```{r , eval=FALSE}
save.image("Assignment13_Workspace.RData")
```

Or you can click the "Save Workspace" button in the Environment pane. 

💾 **Always save the R documents before closing.**

2. Find the assignment in this week's module in Canvas and upload **both** the RMD and the workspace file.

# Today you practiced:

- Created a blank canvas with `ggplot()`.

- Mapped variables inside `aes()`.

- Added geoms for scatter, histogram, and boxplot.

- Used color / fill to compare groups.

- Applied labels and a minimal theme for clarity.


