---
title: "Chapter 8: Base R and bracket indexing"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Overview

*This is Module 4* Review Lesson 6. Lots of editing needed in this chapter!

TODO: WRITE AN OVERVIEW AND WHY LEARNING THE CHAPTER CONTENT IS IMPORTANT.



In Chapter Eight you will learn how to:

State R’s basic subsetting template: `x[rows, cols]`.




Explore with Base R tools and bracket indexing (`[]`) only



## Bracket Indexing `[]`

Chapter 3 introduced you to the **bracket operator**. You learned that the basic pattern is `x[i]`, where `x` is the vector and `i` is the index (or indices) of the elements you want to extract. A position used to identify elements in an object. In R, indices are typically integers (positions), names (characters), or logical values (TRUE/FALSE).

💡 **Tip:** positions start at 1 in R (not 0 like python).

### Vector Indexing Review

🎯 **Goal**  Practice using `[ ]` with **positional indexing** on vectors. 

```{r}
# ⚡ Type this in your R Notebook
z <- c(10, 20, 30, 40, 50) # create a vector
```

```{r}
# ⚡ Type this in your R Notebook
z[1]       
```

🗣 The first element by position is 10.

```{r}
# ⚡ Type this in your R Notebook
z[3]      
```

🗣 The third element by position is 30.

```{r}
# ⚡ Type this in your R Notebook
z[2:4]     
```

🗣 The second, third and fourth elements by position is 20, 30, and 40.

```{r}
z[-c(1, 5)] 
```

🗣 All *except* first and last elements is 20, 30, and 40.


### Data Frame Indexing

The **bracket operator** (`[ ]`) is used to select or replace parts of many kinds of objects including vectors, matrices, and data frames. For data frames, the common form: `df[rows, cols]` or `x[i, j, drop = ]` where `i` is the row index, `j` is the column index, and drop is a logical value that indicates whether to simplify the result to a vector if possible.

When you select a smaller part of an object, using indices such as numbers, names, or logical tests, you are **subsetting** the object. Subsetting is a fundamental operation in R that allows you to extract specific elements from a larger data structure.

- Rows and columns are coordinates:

`df[1:5, ]` first five rows, all columns

`df[ , 1:3]` all rows, first three columns

`df[1, 2]` one cell: first row, second column


## Load Data

🎯 **Goal** Load `field_sightings.csv` into R using `read.csv()` and assign it to a variable called `sightings`.

```{r}
# ⚡ Type this in your R Notebook
field_sightings <- read.csv("field_sightings.csv")
```

🗣 This code uses the `read.csv()` function to read the CSV file named "field_sightings.csv" and assigns the resulting data frame to a variable called `field_sightings`. Make sure that the file is in the same folder as your R Notebook.

🎯 **Goal** Print the first few rows of the `field_sightings` data frame using brackets to check that it loaded correctly

```{r}
# ✅ 
field_sightings[1:3, ]  
```

🗣 Shows first three rows, all columns.




- Selecting columns by name without $:

`df[ , c("age", "income")]`

- A short note on why this matters: this is the foundation for everything later (including dplyr and plotting).

(Cite base principles and then demonstrate; the goal here is fluency and precision.)

### Micro-checks that catch bad imports early

Covers:

- Are column names reasonable?

- Are numeric columns actually numeric?

- Do counts make sense (rows/columns)?



## AI competency: troubleshoot imports safely

AI can help explain error messages and suggest checks.

Identify the privacy risks of certain AI systems

AI should not receive private data or human-subjects data—ever.

What’s safe to share:

- the exact error message

- your import code

last chapter: Markdown structure # 

This chapter: Markdown bold/italics with * / ** so you can emphasize constraints and context in prompts.

### Example AI prompt for troubleshooting



# Memoing and documentation

Documenting your thinking processes (core)

## Memo template

Memo template: what every import needs

0) Heading structure using Markdown (##, ###, etc.)

1) Forethought (planning + strategic preparation)
- 🎯 Goal: explicit goal setting (task analysis).
- 💬 Prediction: anticipatory reasoning (“what do I expect will happen?”), which forces students to activate prior knowledge and commit to a mental model before execution.
- 🧠 Growth mindset (as you define it): motivational beliefs that support persistence—this sits squarely in Zimmerman’s forethought “self-motivational beliefs” bucket (self-efficacy, task value, outcome expectations).
2) Performance (strategy use + monitoring while working)
- ⚡ Code: strategy enactment (writing code to meet the goal).
- ✅ Checks: self-monitoring and control (debugging, validation, sanity checks).
- 🚀 Code trials + ✍️ Record what you tried: systematic self-monitoring and iterative control—this is SRL “self-control” and “self-observation” made visible.
3) Self-reflection (evaluation + adaptation)
- 🗣 Comment: self-evaluation (“why did this work / what did I learn?”) and causal attribution.
- 🪞 Reflect on learning process: explicit self-judgment and adaptive inference (“what will I do differently next time?”).
- 👀 Citing resources and stating how they helped: operationalizes help-seeking and tool-use regulation (key SRL behaviors). 


## Example memo 



# Summary


# Chapter Terms

**Bracket Operator**: The square-bracket indexing operator `[` used to extract or replace parts of an object in R (for example, `x[i]`). For vectors, it selects one or more elements and returns the same kind of object as `x` (a vector).

**Positional Indexing**: Indexing a vector using numeric positions (for example, `x[1]` for the first element or `x[c(1, 3)]` for the first and third). Positions are based on the current order of the vector.

# 📝 Practice Space

## Task 1

## Task 2

# References


