---
title: "Chapter 8"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Overview

*This is Module 4*

TODO: WRITE AN OVERVIEW AND WHY LEARNING THE CHAPTER CONTENT IS IMPORTANT.



In Chapter Eight you will learn how to:

- Courage Skill Highlight

Students can attempt a task first (even imperfectly), then seek help with a clear question and evidence (code + error + what they think is happening).

Students can share imperfect work in a discussion/lab setting and revise based on feedback.

- Git

- Github


## Folder Structure

Covers:

1.. Minimum viable structure (what you described): keep the data in the same folder as the R Notebook so relative paths are simple.

2. Best practice structure: create a `data/` folder and move raw files into it:

-`Intro_to_R/`
-- `chapter8_notes.Rmd`
-- `data/`
--- `myfile.csv`
--- `myfile.xlsx`

Introduce RStudio Projects: a way to keep everything together and keep a stable working directory.

### Privacy and Github

Covers (short, concrete):

- Why `data/` exists: so later you can ignore it when you start using GitHub.

- The mechanism they’ll use later: a .gitignore can exclude folders like `data/` from commits.

- A simple rule of thumb:

1. OK to upload: public/open data, de-identified data you’re allowed to share

2. Do not upload: human-subjects data, protected/controlled data, anything you promised to keep private

## Data source citation

Open R Notebook and document the data source in a memo.

Covers:

- What to record in a memo: where it came from, who made it, date accessed, license/terms if available

- Make the point: if you can’t answer “Where did this data come from?” you shouldn’t analyze it yet.

# Import in R

## Package Datasets

mtcars, iris, etc.

dslabs: data("murders")

## Read CSV

Importing a CSV with Base R (`read.csv`)

Covers:

- Why we start with Base R here: fewer moving parts; clearer mental model.

- The core code pattern (students type it twice: console then notebook):

`df <- read.csv("data/myfile.csv")`

- What `read.csv()` returns: a data frame.

`Quick “checks” immediately after import:

(Maybe Base R Version of head)

`dim(df)`, `names(df)`, `head(df)`, `str(df)`

- Common import issues to anticipate:

1. “cannot open file” (path/working directory problem)

2. weird column names or wrong separators (CSV isn’t really comma-delimited)

This is where you explicitly model Curiosity: “I don’t trust the import yet—I verify it.”

## Read Excel

Importing an Excel file (readxl::read_excel)

Covers:

- Why this requires a package: Base R doesn’t read .xlsx directly.

- Install/load once, then import:

`install.packages("readxl")`

`library(readxl)`

`df <- read_excel("data/myfile.xlsx")`

- What to do when there are multiple sheets: list sheets; choose `sheet = ...` (brief, not exhaustive)

- Why `readxl` is used: it’s designed for tabular Excel import and is broadly used in the R ecosystem.

- Common Excel-specific failures:

1. students try to import a Google Sheet link instead of a downloaded file

2. sheet names don’t match what they assume

# Data frames vs. tibbles

## Data Frames

“Data frame” as the chapter’s working object

Covers (minimal, first-year level):

A data frame is a rectangular table: rows = cases, columns = variables.

Reinforce: import functions usually give you a data frame (or something extremely close).

## Tibbles

Tibbles: same idea, different behavior/printing

Covers (brief + practical):

- A tibble is a modern version of a data frame with nicer printing and stricter behavior.

- Show how students might encounter it (Excel import often yields a tibble-like printout; tidyverse later will use tibbles).

- One conversion demo (optional):

`tib <- tibble::as_tibble(df)`

Promise: “You can treat these as the same for now—our exploration tools still work.”

# R Programming: 

Explore with Base R tools and bracket indexing (`[]`) only

## The “inspect first” toolkit (no subsetting yet)

Covers:

dim(), nrow(), ncol(), names(), head(), tail(), str(), summary()

Interpretation prompts that force curiosity:

“Which columns look categorical vs numeric?”

“Are there missing values? Where?”

## Bracket Indexing `[]`

## Sub-setting Vectors

🎯 **Goal**  Pull out the pieces you need.  Use `[ ]` with positions or a logical test. 

🧠 Remember: positions start at 1 in R (not 0).

Type the following into your R script and run each line. 

```{r , eval=FALSE}
z[1]       # first element by position

z[3]       # third element by position

z[2:4]     # last two elements by position 

z[-c(1, 5)] # all but first and last elements
```

🛠 Break Things! What happens if you try to index with a number larger than the vector length, like `z[6]`? ⚡ Explain any error message you see.

🧠 Challenge: Extract the last two fruits from *chars*.

### Vector Indexing 



### Logical Indexing

Type the following into your R script and run.

```{r , eval=FALSE}
numbers3[numbers3 > 0] # keep only positives
```

🔍 Look deeper: What happens if you run`y <- z[(!is.na(z))]`?  Hint: The function `is.na()` checks for missing values (NA) in the vector. The `!` operator negates the logical vector. Print `y`. 👀 What happened?⚡ Explain

```{r}
# Create a memo note in your R script if you did 🛠 Break Things and 🧠 Challenge 
# Or if you did the 🔍 Look deeper, and 👀 What happened?. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did in Indexing, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo
```


Brackets as a coordinate system: `df[rows, cols]`

Covers (teach as a pattern, not a list of tricks):

- Rows and columns are coordinates:

`df[1:5, ]` first five rows

`df[ , 1:3]` first three columns

`df[1, 2]` one cell

- Selecting columns by name without $:

`df[ , c("age", "income")]`

- A short note on why this matters: this is the foundation for everything later (including dplyr and plotting).

(Cite base principles and then demonstrate; the goal here is fluency and precision.)

### Micro-checks that catch bad imports early

Covers:

- Are column names reasonable?

- Are numeric columns actually numeric?

- Do counts make sense (rows/columns)?



## AI competency: troubleshoot imports safely

AI can help explain error messages and suggest checks.

Identify the privacy risks of certain AI systems

AI should not receive private data or human-subjects data—ever.

What’s safe to share:

- the exact error message

- your import code

last chapter: Markdown structure # 

This chapter: Markdown bold/italics with * / ** so you can emphasize constraints and context in prompts.

### Example AI prompt for troubleshooting



# Memoing and documentation

Documenting your thinking processes (core)

## Memo template

Memo template: what every import needs

0) Heading structure using Markdown (##, ###, etc.)

1) Forethought (planning + strategic preparation)
- 🎯 Goal: explicit goal setting (task analysis).
- 💬 Prediction: anticipatory reasoning (“what do I expect will happen?”), which forces students to activate prior knowledge and commit to a mental model before execution.
- 🧠 Growth mindset (as you define it): motivational beliefs that support persistence—this sits squarely in Zimmerman’s forethought “self-motivational beliefs” bucket (self-efficacy, task value, outcome expectations).
2) Performance (strategy use + monitoring while working)
- ⚡ Code: strategy enactment (writing code to meet the goal).
- ✅ Checks: self-monitoring and control (debugging, validation, sanity checks).
- 🚀 Code trials + ✍️ Record what you tried: systematic self-monitoring and iterative control—this is SRL “self-control” and “self-observation” made visible.
3) Self-reflection (evaluation + adaptation)
- 🗣 Comment: self-evaluation (“why did this work / what did I learn?”) and causal attribution.
- 🪞 Reflect on learning process: explicit self-judgment and adaptive inference (“what will I do differently next time?”).
- 👀 Citing resources and stating how they helped: operationalizes help-seeking and tool-use regulation (key SRL behaviors). 


## Example memo 



# Summary


# Chapter Terms


# 📝 Practice Space

## Task 1

## Task 2

# References


