---
title: "Chapter 9"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

*This is Module 4*


TODO: WRITE AN OVERVIEW AND WHY LEARNING THE CHAPTER CONTENT IS IMPORTANT. Why is this useful as a researcher? How does it fit with where we are in the class. 

In Chapter ____ you will learn how to:

- 

Frame the Module around one repeatable habit: Ask → Load → Verify → Explore → Document.



# Learning Skill



LS
Core
Document thinking process
Advanced

Thoroughness Skill Highlight

- Verify assumptions with checks (e.g., length/type/missingness) before interpreting output.


# R Programming


Lesson 6: (First part), Students load data (don't make the data)

# R Programming: 

Explore with Base R tools and bracket indexing (`[]`) only

## The “inspect first” toolkit (no subsetting yet)

Covers:

dim(), nrow(), ncol(), names(), head(), tail(), str(), summary()

Interpretation prompts that force curiosity:

“Which columns look categorical vs numeric?”

“Are there missing values? Where?”

## Bracket Indexing `[]`

## Sub-setting Vectors

🎯 **Goal**  Pull out the pieces you need.  Use `[ ]` with positions or a logical test. 

🧠 Remember: positions start at 1 in R (not 0).

Type the following into your R script and run each line. 

```{r , eval=FALSE}
z[1]       # first element by position

z[3]       # third element by position

z[2:4]     # last two elements by position 

z[-c(1, 5)] # all but first and last elements
```

🛠 Break Things! What happens if you try to index with a number larger than the vector length, like `z[6]`? ⚡ Explain any error message you see.

🧠 Challenge: Extract the last two fruits from *chars*.

### Indexing 





Brackets as a coordinate system: `df[rows, cols]`

Covers (teach as a pattern, not a list of tricks):

- Rows and columns are coordinates:

`df[1:5, ]` first five rows

`df[ , 1:3]` first three columns

`df[1, 2]` one cell

- Selecting columns by name without $:

`df[ , c("age", "income")]`

- A short note on why this matters: this is the foundation for everything later (including dplyr and plotting).

(Cite base principles and then demonstrate; the goal here is fluency and precision.)

### Micro-checks that catch bad imports early

Covers:

- Are column names reasonable?

- Are numeric columns actually numeric?

- Do counts make sense (rows/columns)?





# Memoing and documentation

Documenting your thinking processes (core)

Curiosity Skill Highlight Continued

## Memo template

Memo template: what every import needs

0) Heading structure using Markdown (##, ###, etc.)

1) Forethought (planning + strategic preparation)
- 🎯 Goal: explicit goal setting (task analysis).
- 💬 Prediction: anticipatory reasoning (“what do I expect will happen?”), which forces students to activate prior knowledge and commit to a mental model before execution.
- 🧠 Growth mindset (as you define it): motivational beliefs that support persistence—this sits squarely in Zimmerman’s forethought “self-motivational beliefs” bucket (self-efficacy, task value, outcome expectations).
2) Performance (strategy use + monitoring while working)
- ⚡ Code: strategy enactment (writing code to meet the goal).
- ✅ Checks: self-monitoring and control (debugging, validation, sanity checks).
- 🚀 Code trials + ✍️ Record what you tried: systematic self-monitoring and iterative control—this is SRL “self-control” and “self-observation” made visible.
3) Self-reflection (evaluation + adaptation)
- 🗣 Comment: self-evaluation (“why did this work / what did I learn?”) and causal attribution.
- 🪞 Reflect on learning process: explicit self-judgment and adaptive inference (“what will I do differently next time?”).
- 👀 Citing resources and stating how they helped: operationalizes help-seeking and tool-use regulation (key SRL behaviors).


# Advanced R Programming


# Summary


# Chapter Terms


# 📝 Practice Space

## Task 1

## Task 2

# References

