---
title: "Chapter 8: Base R and bracket indexing"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Overview

*This is Module 4* Review Lesson 6. Lots of editing needed in this chapter!

TODO: WRITE AN OVERVIEW AND WHY LEARNING THE CHAPTER CONTENT IS IMPORTANT.



In Chapter Eight you will learn how to:

State R’s basic subsetting template: `x[rows, cols]`.

Read and write the `x[rows, cols]` pattern to pull out what you need and safely overwrite values.


Explore with Base R tools and bracket indexing (`[]`) only

# Set Up

To begin this guided activity, you'll want to open, format, and save an R Notebook in your course folder. This will be your workspace for practicing the content of Chapter 8. Then, you'll practice loading data. We'll be using the `field_sightings.csv` dataset.

## Open an R Notebook

To follow along Chapter 8, 🎯 please open a new R Notebook in your course folder named `chapter8_notes.Rmd` and add a date to the YAML header. While your here, **include your name as the author**! See the example of a YAML below. Use this document to take notes and practice the chapter content. 

1. Open RStudio → File → New File → R Notebook.

2. Save it in your course folder as: `chapter6_notes.Rmd`.

3. In your YAML header at the top:


```{r echo=FALSE, results='asis'}
cat("```yaml
---
title: \"Chapter 8 Practice: Bracket Indexing\"
author: \"Your Name Here\"
date: \"`r Sys.Date()`\"
output: html_notebook
---
```")
```

4. Remove the default content in the R Notebook and replace it with your notes and code as you work through the chapter.

5. Use headings as you work. At minimum, create section headings that match the chapter structure (e.g., `# Bracket Indexing`, `# Subsetting`, `# Extraction`, `# Insertion`. This creates a readable outline.

6. Memo as you go.

- Outside chunks: write normal text.

- Inside chunks: use `#` for any code comments (otherwise R tries to run the text).

- If you hit an error: keep the broken line commented out and write a one-sentence note about what you changed to fix it.

## Load Data

The `field_sightings.csv` dataset contains **simulated** (randomly generated) values. The dataset theme is based on the pretend situation where field biologists observed and recorded a population of species of birds. Bird banding is a scientific method used to track and study individual birds, providing valuable data on their movements, behavior, and population dynamics. In our senerio, most birds observed were "Untagged," but a small number have been previously captured and marked with colored leg bands, "Tagged". The color of the bands might provide information about the bird's history, such as where it was banded or its age.

The simulated dataframe `field_sightings` has two columns:

- `Status` Describes whether the sighted bird is “Untagged” (the common case) or “Tagged” (the rare case).

- `Band_Color` If a bird is untagged, the value is “None.” If it is tagged, the value is one of several possible band colors.

🎯 **Goal** Download the `field_sightings.csv` file from Canvas and save it in the *same* folder as your R Notebook. Then, load the data into R.

🎯 **Goal** Load `field_sightings.csv` into R using `read.csv()` and assign it to a variable called `sightings`. Recall that you need to library `readr` before you can use `read.csv()`.

```{r message=TRUE, warning=TRUE}
# ⚡ Type this in your R Notebook
library(readr) # Load the readr package for reading CSV files
field_sightings <- read.csv("field_sightings.csv")
```

🗣 This code uses the `read.csv()` function to read the CSV file named "field_sightings.csv" and assigns the resulting data frame to a variable called `field_sightings`. It it normal to see *warning* messages with you load packages like `readr()`.

## Bracket Indexing `[ ]`

In R, indices are typically integers (positions), names (characters), or logical values (TRUE/FALSE).

### Positional Indexing 

Positional indexing is the most basic form of indexing in R, where you use numeric positions to select elements from a vector or data frame.

---

📖 **Review**: In Chapter 3, you practiced using the **bracket operator** with positional indexing to extract elements from vectors. Let's review that briefly before we move on to data frames. 

💡 **Tip:** Positions start at `1` in R. This is different from other coding languages like python which starts indexing values at `0`.

You learned that the basic pattern for **positional indexing** is `x[i]`, where `x` is the vector and `i` is the index (or indices) of the elements you want to extract. A position used to identify elements in an object. 

🎯 **Goal**  Practice using `[ ]` with positional indexing on vectors. 

```{r}
# ⚡ Type this in your R Notebook
z <- c(10, 20, 30, 40, 50) # create a vector
```

```{r}
# ⚡ Type this in your R Notebook
z[1]       
```

🗣 The first element by position is 10.

```{r}
# ⚡ Type this in your R Notebook
z[3]      
```

🗣 The third element by position is 30.

```{r}
# ⚡ Type this in your R Notebook
z[2:4]     
```

🗣 The second, third and fourth elements by position is 20, 30, and 40.

```{r}
# ⚡ Type this in your R Notebook
z[-c(1, 5)] 
```

🗣 All *except* first and last elements is 20, 30, and 40.

---

#### Data Frame Positional Indexing

The **bracket operator** (`[ ]`) is used to select or replace parts of many kinds of objects, not just vectors, including matrices, and data frames. For data frames, the common form is `df[rows, cols]` or `x[i, j, drop = ]` where `i` is the **row index**, `j` is the **column index**, and drop is a logical value that indicates whether to simplify the result to a vector if possible.

When you select a smaller part of an object, using indices such as numbers, names, or logical tests, you are **subsetting** the object. Subsetting is a fundamental operation that allows you to select specific elements from a larger data structure.

- Rows and columns are coordinates:

`df[1:5, ]` results in the first five rows, all columns

`df[ , 1:3]` shows all rows, first three columns

`df[1, 2]` prints only one cell's value: first row, second column

Let's see this in practice with our `field_sightings` data frame.

🎯 **Goal** Print the first six rows of the `field_sightings` data frame using the bracket operator.

```{r}
# ⚡ Type this in your R Notebook
field_sightings[1:6, ]  
```

🗣 Shows first three rows, all columns. 

### `head()` function

There is a function called `head()` that *basically* does the same thing as `field_sightings[1:6, ]` but is more concise and easier to read. 

📜 The SYNTAX (basic): 

`head(x, n = 6L)`

If you don't specify `n`, it defaults to 6, which means it will show the first 6 rows of the data frame. You can change this number to show more or fewer rows.

🎯 **Goal** Print the first three rows of the `field_sightings` data frame using the `head()` function to check that the dataset loaded correctly.

```{r}

# ⚡ Type this in your R Notebook
head(field_sightings, n = 3)
```

🗣 `head()` is commonly used to get a quick orientation of the dataset after loading. 

`head()` does not do arbitrary extraction or subsetting. `head()` is a function that provides a convenient way to view the first few rows of a data frame. 

Bracket indexing allows for flexible and specific subsetting of the data frame based on row *and* column indices.  `[` works on almost everything and supports rich indexing and is commonly used as part of an algorithm. You will use `[` when you need subsetting by columns, conditions, and arbitrary rows when you are writing analysis code.

### Character Indexing

So far, these examples have used **positional indexing**. You can also use **character indexing** to select columns by their names instead of their positions. This is often more readable and less error-prone, especially when the order of columns might change or when you have many columns.

---

📖 **Review** In Chapter 3, you created the `nerdy_palette` vector and then used character indexing to return the hexadecimal color code. 

```{r}
# 📖 Review
# Create nerdy_palette
nerdy_palette <- c(
"Octarine" = "#7F00FF",
"TARDIS Blue" = "#003B6F",
"Rebeccapurple" = "#663399",
"Hooloovoo" = "#4F97A3",
"Cosmic Latte" = "#FFF8E7",
"Vantablack" = "#000000",
"Matrix Green" = "#03A062",
"Source Engine Magenta" = "#FF00FF",
"Command Red" = "#CC0000",
"Impossible Green" = "#4D9E68"
)
# Use character indexing to return the hex code for Octarine
nerdy_palette["Octarine"]
```

🗣 The output shows the hexadecimal color code for “Octarine” retrieved from the `nerdy_palette` vector using its name.

---

#### Data Frame Character Indexing

We can use character indexing to select a column by name. 

🎯 **Goal** Use character indexing to return the band color in the `field_sightings` data frame. To reduce what prints, select on the the 46th through 55th rows.

```{r}
# ⚡ Type this in your R Notebook
field_sightings[46:55, "Band_Color"] 
```

🗣 A vector is printed showing the element values stored in rows 46 though 55 for the column "Band Color."

### `$` Operator

There is another operator called the **dollar sign operator** (`$`) that is used to extract columns from a data frame by name. It is a convenient shorthand for selecting a single column or element.

# here





- A short note on why this matters: this is the foundation for everything later (including dplyr and plotting).

(Cite base principles and then demonstrate; the goal here is fluency and precision.)

### Micro-checks that catch bad imports early

Covers:

- Are column names reasonable?

- Are numeric columns actually numeric?

- Do counts make sense (rows/columns)?



## AI competency: troubleshoot imports safely

AI can help explain error messages and suggest checks.

Identify the privacy risks of certain AI systems

AI should not receive private data or human-subjects data—ever.

What’s safe to share:

- the exact error message

- your import code

last chapter: Markdown structure # 

This chapter: Markdown bold/italics with * / ** so you can emphasize constraints and context in prompts.

### Example AI prompt for troubleshooting



# Memoing and documentation

Documenting your thinking processes (core)

## Memo template

Memo template: what every import needs

0) Heading structure using Markdown (##, ###, etc.)

1) Forethought (planning + strategic preparation)
- 🎯 Goal: explicit goal setting (task analysis).
- 💬 Prediction: anticipatory reasoning (“what do I expect will happen?”), which forces students to activate prior knowledge and commit to a mental model before execution.
- 🧠 Growth mindset (as you define it): motivational beliefs that support persistence—this sits squarely in Zimmerman’s forethought “self-motivational beliefs” bucket (self-efficacy, task value, outcome expectations).
2) Performance (strategy use + monitoring while working)
- ⚡ Code: strategy enactment (writing code to meet the goal).
- ✅ Checks: self-monitoring and control (debugging, validation, sanity checks).
- 🚀 Code trials + ✍️ Record what you tried: systematic self-monitoring and iterative control—this is SRL “self-control” and “self-observation” made visible.
3) Self-reflection (evaluation + adaptation)
- 🗣 Comment: self-evaluation (“why did this work / what did I learn?”) and causal attribution.
- 🪞 Reflect on learning process: explicit self-judgment and adaptive inference (“what will I do differently next time?”).
- 👀 Citing resources and stating how they helped: operationalizes help-seeking and tool-use regulation (key SRL behaviors). 


## Example memo 



# Summary


# Chapter Terms

**Bracket Operator**: The square-bracket indexing operator `[` used to extract or replace parts of an object in R (for example, `x[i]`). For vectors, it selects one or more elements and returns the same kind of object as `x` (a vector).

**Column index**: The index used to select columns in a data frame or matrix (the second slot inside brackets: `x[, j]`).

**Dollar Sign Operator** (`$`): An extraction (and replacement) operator used to access a single named component of a recursive object. Most commonly a data frame column (`df$col`) or a list element (`lst$name`). `$` requires a literal name (not a name stored in a variable), returns NULL if the name doesn’t exist, and by default may do partial matching (which can surprise you).

**Extraction**: Pulling values out of an object by subsetting on the right-hand side of an expression (e.g., `x[i]`), which returns the selected elements.

**Insertion**: Putting values into an object by subsetting on the left-hand side of an assignment (e.g., `x[i] <- value`), replacing or updating only the indexed part of the object (also called replacement or subassignment).

**Positional Indexing**: Indexing a vector using numeric positions (for example, `x[1]` for the first element or `x[c(1, 3)]` for the first and third). Positions are based on the current order of the vector.

**Row index**: The index used to select rows in a data frame or matrix (the first slot inside brackets: `x[i, ]`)

**Subsetting**: Selecting a smaller part of an R object (for example, specific elements of a vector, rows/columns of a data frame, or components of a list) using R’s subsetting operators like `[`, `[[`, and `$`.

# 📝 Practice Space

## Task 1

## Task 2

# References


