---
title: "Chapter 5"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Overview

*This is Module 4*

TODO: WRITE AN OVERVIEW AND WHY LEARNING THE CHAPTER CONTENT IS IMPORTANT.

In Chapter Five you will learn how to:

- 📊 Subset vectors by position or logical tests.

# Learning Skill

Curiosity

LS
Core
Document thinking process
Advanced
Curiosity Skill Highlight


LS
Core
Formating Documents
Advanced
Autonomy Skill Highlight
Review course content


# Design

R Markdown

# AI Competency

Human-centered Mindset

Core
Identify the privacy risks of certain AI systems
Advanced
Explain the difference between humans controlling AI versus AI controlling humans


# R Programming: 

Lesson 5

Lesson 6: (First part)

## Sub-setting Vectors

🎯 **Goal**  Pull out the pieces you need.  Use `[ ]` with positions or a logical test. 

🧠 Remember: positions start at 1 in R (not 0).

Type the following into your R script and run each line. 

```{r , eval=FALSE}
z[1]       # first element by position

z[3]       # third element by position

z[2:4]     # last two elements by position 

z[-c(1, 5)] # all but first and last elements
```

🛠 Break Things! What happens if you try to index with a number larger than the vector length, like `z[6]`? ⚡ Explain any error message you see.

🧠 Challenge: Extract the last two fruits from *chars*.

## Logical Indexing

Type the following into your R script and run.

```{r , eval=FALSE}
numbers3[numbers3 > 0] # keep only positives
```

🔍 Look deeper: What happens if you run`y <- z[(!is.na(z))]`?  Hint: The function `is.na()` checks for missing values (NA) in the vector. The `!` operator negates the logical vector. Print `y`. 👀 What happened?⚡ Explain

```{r}
# Create a memo note in your R script if you did 🛠 Break Things and 🧠 Challenge 
# Or if you did the 🔍 Look deeper, and 👀 What happened?. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did in Indexing, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo
```


