---
title: "Courage Lab"
author: "by Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 2 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Courage Lab"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
 
Or
title: "Courage Lab"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
team: "Fall 2025"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 2 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```

# 👥 Human-Centered AI

This course teaches AI competencies: Human-centered mindset, AI techniques and applications, and Ethics of AI. So far, we've focused on AI techniques and applications such and prompt engineering. Today, we take a step back and think about when and how to use AI in a way that centers human relationships and learning.

Help-seeking is a social act. In learning, research, and work settings, asking for help typically does more than solve a problem: it signals vulnerability; it activates reciprocity. it transmits norms; and it builds trust and belonging. When you ask your peer, instructor, or others for help, you are participating in a relational infrastructure. Quick questions before class, brief check-ins during office hours, and small moments of confusion shared with classmates all build a sense of community and shared purpose. If an LLM becomes the easiest, fastest, lowest-friction option, you will experience fewer micro-interactions, have fewer peer modeling opportunities, and fewer moments where norms, strategies, or values are socially transmitted. Using LLM exclusively will reduce the density of casual, learning-oriented contact, which is often where community is actually built. Those “small” interactions are how learning communities form.

An AI can give instrumental help (fast feedback, examples, explanations) without any social risk. That can be especially useful when you feel anxious, behind, or worried about looking foolish. LLMs don't judge, don't get tired, don't keep score, and don't *feel like* you are disclosing private information to others. But it also means AI can quietly replace the low-stakes human interactions that help you feel connected. In other words, LLMs provide private competence-building but may result in feelings of isolation. A shared learning ecology offers connection but requires volubility and time. 

# 🦁 Courage

Fear shows up in predictable ways: fear of struggling, fear of failure, fear of speaking up, fear of being “found out,” and fear of looking foolish in front of other people. The problem is not that fear exists. The problem is what fear does when it drives your behavior. It keeps you on the sidelines, makes you passive, and convinces you to protect your image instead of taking risks. Over time, that disengagement compounds. You miss opportunities to practice, you avoid asking questions, and you may fall behind. That is why fear is the enemy of deep learning.

Courage is not the absence of fear. Courage is the ability to manage fear in service of truth and understanding. In practical terms, courage means you act even when you feel exposed or uncertain. You volunteer an answer attempt, you ask the “dumb” question, you share work-in-progress, you bring an error message to a peer or to office hours, and you let feedback guide your next attempt. Courage also means responding to obstacles with agency: you stay involved, you try new strategies, and you use available resources instead of ruminating on mistakes or interpreting errors as evidence that you “*aren’t good at this.*”

Courage matters because it directly determines whether you will actually use the learning environment you’re in. A student can have access to strong teaching, helpful peers, office hours, and AI tools and still not benefit if fear prevents engagement. Courage is the bridge between resources and results. It is also a relationship skill. When you practice courage, you make help-seeking possible, and help-seeking builds community. That’s why courage sits next to “human-centered AI” in this course. AI can support learning, but courage is what keeps you connected to people and willing to learn in public, which is where much of the growth happens.

# 🛠 Lab Activity

## Purpose

This lab is about practicing courage and human-centered AI use. You will (1) identify what you’re stuck on, (2) name what makes asking hard, and (3) ask a real person for help using a clear, respectful message. AI tools can help you clarify your question, but in this lab a human is the target audience.


⏱️ 30 minute plan at a glance

 - 0-5 min — Set up your notebook + 
 
 - 8-18 min — Courage Inventory

 - 18-25 min — Reflection + prepare your help request

 - 25-30 min — Submission + actually send the help request
 
If you don’t have questions, be a 🚑 helper for someone who does to earn learning skill points.

## 🖥 Courage Lab Notebook

Please take 5 minutes to set up your notebook

1. Open RStudio. 

2. Start a new R Notebook named `courage_lab`

3. In the YAML header, title it "Courage Lab" and include your name + date.

4. Delete the template text.

5. Create a section header: ## Question Inventory

## 🔍 Courage Inventory

Please take 10 minutes to reflect on your recent work and identify a specific question you have right now.

Under ## Question Inventory, respond using markdown sub-headers. Use code chunks for any code or error messages.

1. Briefly review the assignment/activity you are working on or struggled with.

- What’s one moment from the last assignment/activity that you found difficult or that stuck with you? 

- What did you do when you got stuck? For example, did you keep trying, stop, search the web, ask a peer, ask the instructor, use ?help, other?

2. Identify *at least* one question you have right now.
    - Source: (Assignment __, feedback comment, class activity, etc.)
    - My question:
    - My code &/or error:
    
```{r}
# Include code example
```

3. What is your the main barrier to asking for help? Examples may include fear of *looking dumb*, not sure how to ask, procrastination, overwhelmed, time, “*I should know this.*”

4. Choose your help path (must choose ONE)
    A) Ask a peer (in class today)
    B) Ask the instructor (message or office hours request)

🆘 For this lab, if you use AI, only use it to improve the question you will ask a person. Please create a transparency memo regarding your AI use in your notebook.

## 📝 Prepare Your Help Request

Please take a few minutes to draft your help request message.

Using markdown, draft your help request message.

    Hi ___, I’m working on ___. I’m stuck on ___. I tried ___. Here is my code/error: ___.
    My best guess is ___. Could you help me understand what to try next?

Including a screenshot of your code/error is optional but can be helpful.

## 🪞 Reflect on Your Courage

Pleae take a few minutes to reflect on your courage practice. The questions below are prompts to guide your reflection. You can answer all or some of them or write your own.

What part of this felt uncomfortable or risky?

What is your next step before the next class?

Who did you connect with today, and what is one thing you learned about how they think?

# 🚑 Helper option

If you don’t have a question, your job is to help someone else ask a better one.

1. Find a peer who has a question.

2. Help them clarify their question using the Courage Inventory and Help Request templates.

3. Record your help-giving in your notebook:

I helped: ___

Their question was about: ___

What I did: (helped them write the help request / helped reproduce the error / explained a concept)

# 📤 Submission

1. Save your R Notebook.

2. Knit to HTML.

3. Upload the HTML file to the Courage Lab assignment in Canvas.

4. Send your help request. If you opted to message the instructor, please send it within 24 hours of completing this lab.

5. 🎉 Celebrate your courage. The outcome is not “having the right answer.” The outcome is staying engaged and practicing help-seeking.