---
title: "Chapter 9: Data Dictionary"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Spreadsheets are easy to analyze but hard to self-document. A separate data dictionary file makes the data interpretable later.

By the end of Chapter 9, you will be able to:

- explain how documentation supports analysis

- understanding subassignment


# Data Dictionary

A **data dictionary** is documentation that explains what each variable (column) means so someone else can correctly interpret and reuse the dataset. Its a codebook with variable names/labels, coding values (what 0/1 means), missing-value codes, and any special notes needed to avoid misanalysis. 

## Example Data Dictionary

In RStudio, the help pane is your first stop for built-in datasets. The `?` operator opens the documentation for a dataset, which includes a data dictionary. For example, `?mtcars` will show you the data dictionary for the `mtcars` dataset, which includes variable names, descriptions, and data types.

```{r, eval=FALSE}
# ⚡ Type this in your R Studio Console
?mtcars
```

The help page lists the dataset context (1974 Motor Trend road tests) and provides variable descriptions, units, and several coded values (e.g., `am` is 0/1 for automatic/manual).

# Create a report

Please open a new R Notebook in your course folder named `chapter9_notes.Rmd` and edit the YAML. Use this document to take notes and practice the chapter content. 

1. Open RStudio → File → New File → R Notebook.

2. Save it in your course folder as: `chapter9_notes.Rmd`.

3. In your YAML header at the top:


```{r echo=FALSE, results='asis'}
cat("```yaml
---
title: \"Chapter 9 Practice: Data Dictionary\"
author: \"Your Name Here\"
date: \"`r Sys.Date()`\"
output: html_notebook
---
```")
```

4. Remove the default content in the R Notebook and replace it with your notes and code as you work through the chapter.

5. Use headings as you work. 

6. Memo as you go.

# Data Collection Story

Read the story provided below. It describes the context of the dataset and how the data were collected. This information is crucial for understanding the meaning of each variable and how to interpret the data correctly. You will use this story to create a data dictionary for the `field_sightings_expanded.csv` dataset.

---

Just after sunrise in the Asterfall Wetlands of the fictional Cedarglass Basin, a small field crew made the same loop every week from April 1 through July 31, 2025—boots wet, clipboards out, scanning reed edges for movement and listening for the soft trill of the **Asterfall Reedlark (ARLK)**. A decade earlier, the Reedlark vanished from the basin after a fast-moving respiratory outbreak locals now call Marshfade. The 78 surviving birds were taken into a captive breeding program, and in 2024–2025 the team began releasing juveniles into six restored wetlands to test whether the habitat could support a comeback.

Each line in the dataset records one detection event—one bird seen during a site visit. This means the same individual can appear more than once across the season if it is seen again on a later date. Every detection event was assigned a unique tracking number, and each observer used an assigned ID so the project could audit field notes later.

Surveys were conducted in the early morning, and field notes record the calendar date and the local start time of the detection using a 24-hour clock. Locations are recorded using two-letter site abbreviations for the six wetlands: **Basin Rim (BR)**, **Lakeside Cove (LC)**, **Meadow Reach (MR)**, **North Pool (NP)**, **South Riparian (SR)**, and **Willow Channel (WC)**. Coordinates are recorded in decimal degrees from a handheld GPS at the observer’s station; they are close approximations rather than exact bird locations. Observers also estimated the distance to the bird in meters; when vegetation or glare prevented a reasonable estimate, distance was left blank.

The project’s focal species is the **Asterfall Reedlark (ARLK)**, but incidental detections were also logged for common wetland birds: **Song Sparrow (SOSP)**, **Red-winged Blackbird (RWBL)**, and **American Robin (AMRO)**.  Birds from the reintroduction program were fitted with one colored leg band. Band color encodes hatch year: **silver** for **2023**, **blue** for **2024**, and **green** for **2025**. Because wild hatch birds are unbanded, they have no band color recorded and no individual ID. Reintroduced birds, however, have an individual ID, and that ID may repeat across multiple detections because the same bird can be observed on different days or at different sites.

Age class was recorded using four categories—**hatch-year**, **second-year**, **adult**, or **unknown**—with numeric codes **1**, **2**, **3**, and **99** respectively. For reintroduced birds, **age class** is derived from the known hatch year. For wild hatch birds, age class is estimated visually and is sometimes marked unknown. Sex was recorded when it could be determined visually; sex uses numeric codes where **1** indicates **female** and **2** indicates **male**. When sex could not be determined it was left missing.

Release information applies only to reintroduced birds. Birds hatched in 2023 were released in 2024, and birds hatched in 2024 or 2025 were released in 2025. Each reintroduced bird is also assigned a **cohort label** that combines hatch year and release year (for example, **hatch-2024** / **release-2025**). Wild hatch birds have no release year or cohort label.

**Purpose of the study**: The conservation team is evaluating whether reintroduction plus wetland restoration is working in Cedarglass Basin—specifically, whether reintroduced Reedlarks are being detected repeatedly across the season and across multiple sites (suggesting short-term survival and dispersal), and whether detections of unbanded, wild hatch birds are becoming common enough to indicate the population may be moving toward a self-sustaining recovery.

```{r eval=FALSE, include=FALSE}
# field_sightings_expanded.csv (visual-only version for data dictionary activity)
set.seed(8)

n <- 4470
n_wild <- 4168
n_reintro <- 302

# --- Core status variables ---
Status <- sample(c(rep("Wild Hatch", n_wild), rep("Reintroduced", n_reintro)))
Tag_Flag <- ifelse(Status == "Reintroduced", 1L, 0L)

idx_reintro <- which(Tag_Flag == 1L)
idx_wild    <- which(Tag_Flag == 0L)

# Hatch-year is NOT stored, but it drives band color + age class for reintroduced birds
hatch_year <- rep(NA_integer_, n)
hatch_year[idx_reintro] <- sample(
  c(2023L, 2024L, 2025L),
  length(idx_reintro),
  replace = TRUE,
  prob = c(0.25, 0.45, 0.30)
)

# Band color encodes hatch year (Silver=2023, Blue=2024, Green=2025)
Band_Color <- rep("None", n)
Band_Color[idx_reintro] <- ifelse(
  hatch_year[idx_reintro] == 2023L, "Silver",
  ifelse(hatch_year[idx_reintro] == 2024L, "Blue", "Green")
)

# Reintroduced bird IDs repeat (same individual can be detected multiple times)
bird_pool <- paste0("B", sprintf("%05d", sample(10000:99999, 125)))
Bird_ID <- rep(NA_character_, n)
Bird_ID[idx_reintro] <- sample(bird_pool, length(idx_reintro), replace = TRUE)

# --- IDs and timing ---
Sighting_ID <- paste0("S", sprintf("%06d", 1:n))
Observer_ID <- paste0("O", sprintf("%02d", sample(1:12, n, replace = TRUE)))

date_seq <- seq(as.Date("2025-04-01"), as.Date("2025-07-31"), by = "day")
Date <- sample(date_seq, n, replace = TRUE)

# Most surveys are early morning (05:00–11:00)
mins <- sample((5*60):(11*60), n, replace = TRUE)
Time <- sprintf("%02d:%02d", mins %/% 60, mins %% 60)

# --- Sites in the fictional Asterfall Wetlands (Cedarglass Basin) ---
sites <- data.frame(
  Site_Code = c("BR","LC","MR","NP","SR","WC"),
  Site_Name = c("Basin Rim","Lakeside Cove","Meadow Reach","North Pool","South Riparian","Willow Channel"),
  lat0 = c(43.61, 43.58, 43.64, 43.66, 43.56, 43.62),
  lon0 = c(-116.21, -116.18, -116.24, -116.16, -116.22, -116.19)
)

Site_Code <- sample(sites$Site_Code, n, replace = TRUE, prob = c(0.18,0.16,0.19,0.15,0.17,0.15))
lat0 <- sites$lat0[match(Site_Code, sites$Site_Code)]
lon0 <- sites$lon0[match(Site_Code, sites$Site_Code)]
Latitude  <- round(lat0 + rnorm(n, 0, 0.008), 5)
Longitude <- round(lon0 + rnorm(n, 0, 0.008), 5)

# --- Species: target is fictional, others are incidental detections ---
Species_Common <- sample(
  c("Asterfall Reedlark", "Song Sparrow", "Red-winged Blackbird", "American Robin"),
  n, replace = TRUE, prob = c(0.70, 0.15, 0.10, 0.05)
)

Species_Code <- ifelse(Species_Common == "Asterfall Reedlark", "ARLK",
                ifelse(Species_Common == "Song Sparrow", "SOSP",
                ifelse(Species_Common == "Red-winged Blackbird", "RWBL", "AMRO")))

# Visual-only: estimated distance (meters); sometimes unknown due to obstructed view (NA)
Distance_m <- round(runif(n, 5, 120), 1)
idx_dist_na <- sample(1:n, size = round(0.03 * n))
Distance_m[idx_dist_na] <- NA_real_

# --- Age (with special code 99 = Unknown) ---
Age_Class <- rep(NA_character_, n)
Age_Code  <- rep(NA_integer_,  n)

# Reintroduced birds: age class follows hatch year relative to 2025
Age_Class[idx_reintro] <- ifelse(hatch_year[idx_reintro] == 2025L, "Hatch-year",
                          ifelse(hatch_year[idx_reintro] == 2024L, "Second-year", "Adult"))
Age_Code[idx_reintro] <- ifelse(Age_Class[idx_reintro] == "Hatch-year", 1L,
                         ifelse(Age_Class[idx_reintro] == "Second-year", 2L, 3L))

# Wild-hatched birds: age estimated in the field; 99 = Unknown
Age_Code[idx_wild] <- sample(c(1L, 2L, 3L, 99L), length(idx_wild), replace = TRUE, prob = c(0.28, 0.22, 0.40, 0.10))
Age_Class[idx_wild] <- ifelse(Age_Code[idx_wild] == 1L, "Hatch-year",
                      ifelse(Age_Code[idx_wild] == 2L, "Second-year",
                      ifelse(Age_Code[idx_wild] == 3L, "Adult", "Unknown")))

# --- Sex: introduce more NA, especially for Wild Hatch + Hatch-year ---
# Sex_Code: 1=F, 2=M, NA=Not determined
Sex_Code <- sample(c(1L, 2L, NA_integer_), n, replace = TRUE, prob = c(0.45, 0.45, 0.10))

# More unknown sex for Wild Hatch hatch-year birds (hard to sex visually)
idx_sex_hard <- which(Status == "Wild Hatch" & Age_Code == 1L)
Sex_Code[idx_sex_hard] <- sample(c(1L, 2L, NA_integer_), length(idx_sex_hard),
                                 replace = TRUE, prob = c(0.35, 0.35, 0.30))

Sex <- ifelse(is.na(Sex_Code), NA_character_, ifelse(Sex_Code == 1L, "F", "M"))

# --- Release information (only for reintroduced birds) ---
# Use NA instead of -99 for "not applicable" to increase realistic missingness
Release_Year <- rep(NA_integer_, n)
Release_Cohort <- rep(NA_character_, n)

Release_Year[idx_reintro] <- ifelse(hatch_year[idx_reintro] == 2023L, 2024L, 2025L)
Release_Cohort[idx_reintro] <- paste0("H", hatch_year[idx_reintro], "_R", Release_Year[idx_reintro])

# --- Build dataset (visual-only; no measurement columns; no observation-method columns) ---
field_sightings_expanded <- data.frame(
  Sighting_ID, Observer_ID, Date, Time,
  Site_Code, Latitude, Longitude,
  Species_Common, Species_Code,
  Status, Band_Color, Tag_Flag, Bird_ID,
  Distance_m,
  Age_Class, Age_Code,
  Sex, Sex_Code,
  Release_Year, Release_Cohort
)

write.csv(field_sightings_expanded, "field_sightings_expanded.csv", row.names = FALSE)
```

---

# Load the dataset

## Load Data

The `field_sightings_expanded.csv` dataset contains **simulated** (randomly generated) values. 

🎯 **Goal** Download the `field_sightings_expanded.csv` file from Canvas and save it in the *same* folder as your R Notebook. Then, load the data into R.

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
# 1) Confirm the CSV is in the same folder as your R Notebook
list.files()
```

🎯 **Goal** Load `field_sightings_expanded.csv` into R using `read_csv()` and assign it to a variable called `feild_sightings`. Recall that you need to library `readr` before you can use `read_csv()`.

```{r message=FALSE, warning=FALSE}
# ⚡ Type this in your R Notebook
library(readr) # Load the readr package for reading CSV files
field_sightings <- read_csv("field_sightings_expanded.csv")
```

## Inspect the dataset

### Basic inspection

🎯 **Goal** Do a first pass inspection to know what columns exist and what kinds of values they contain.

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
head(field_sightings) # View the first few rows of the dataset
colnames(field_sightings)  # Get the structure of the dataset (variable names, types
```

🗣 The column names are "Sighting_ID", "Observer_ID", "Date", "Time", "Site_Code", "Latitude", "Longitude", "Species_Common", "Species_Code", "Status", "Band_Color", "Tag_Flag", "Bird_ID", "Distance_m", "Age_Class", "Age_Code", "Sex", "Sex_Code",  "Release_Year", and "Release_Cohort".


### Detailed Checks

🎯 **Goal** Identify (1) coded values, (2) missing values, and (3) relationships between fields.

We'll use the base R function `table()`

📜 The SYNTAX: 

table(x, y, useNA = "ifany")

#### Sex_Code

🎯 **Goal** What values are used for sex codes (including missing)?

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
table(field_sightings$Sex_Code, useNA = "ifany")
```

🗣 `Sex_Code` is 1 (n = 1985) for female, 2 (n = 1891) for male, and `NA` (n = 684) for missing values.


#### Age_Code

🎯 **Goal** What values are used for age codes?

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook

table(field_sightings$Age_Code, useNA = "ifany")
```

🗣 `Age_Code` is 1 (n = 1238) for hatch-year, 2 (n = 1094) for second-year, 3 (n = 1742) for adult, and 99 (n = 396) for unknown. 

#### Band_Color and Status relationship

🎯 **Goal** Check the story claim: wild hatch birds have no band, reintroduced birds do.

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
table(field_sightings$Status, field_sightings$Band_Color, useNA = "ifany")
```

🗣 "Wild hatch" only appears with band `color = None` and "Reintroduced" appears with Silver / Blue / Green.

#### Latitude & Longitude

🎯 **Goal** Range check for coordinates and distance

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
range(field_sightings$Latitude, na.rm = TRUE)
range(field_sightings$Longitude, na.rm = TRUE)
range(field_sightings$Distance_m, na.rm = TRUE)
```


🗣 Latitude is around 43.53 to 43.69; Longitude is around -116.27 to -116.14; Distance is 5 to 120 meters

# Create your data dictionary

To create a data dictionary, we will start by creating a template that includes the variable names and their data types. Then, we will fill in the meanings, units, codes/values, missing values, valid values, and any notes or relationships for each variable.

## Create a template

🎯 **Goal** Create a blank data dictionary template. 

There should be one row per field in the dataset, with an empty place to write meanings later, units, codes, missingness, and notes.

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
data_dictionary <- data.frame(
  Field_Name = names(field_sightings),
  Meaning = "",
  Type = sapply(field_sightings, class),
  Units = "",
  Codes_Values = "",
  Missing_Values = "",
  Valid_Values = "",
  Notes_Relationships = "",
  stringsAsFactors = FALSE
)

print(data_dictionary)
```

🗣 `names(field_sightings)` grabs the dataset’s column names. Those become the first column of the dictionary.

🗣 `sapply(field_sightings, class)` finds the data type of each column (character, numeric, etc.) so I don’t have to guess.

🗣 All the documentation columns start as `""` (blank strings) because I will fill them in later.

🗣 `print(data_dictionary)` shows the template so I can verify it has one row per dataset field.

## Subassignment

**Subassignment** is when you replace part of an object by putting an index on the left-hand side of `<-`. R updates only the indexed piece, not the whole object.

🎯 **Goal** Fill in **one row** of the dictionary (the row for sex code) using subassignment. This will updating only specific cells instead of rewriting the entire table.

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
row <- which(data_dictionary$Field_Name == "Sex_Code")

data_dictionary[row, "Meaning"] <- "Sex of the bird when it could be determined visually."
data_dictionary[row, "Units"] <- "N/A"
data_dictionary[row, "Codes_Values"] <- "1 = female, 2 = male"
data_dictionary[row, "Missing_Values"] <- "NA means sex could not be determined"
data_dictionary[row, "Valid_Values"] <- "1, 2, or NA"
data_dictionary[row, "Notes_Relationships"] <- "Matches Sex (F/M); more NA for wild hatch hatch-year birds."
print(data_dictionary)
```

🗣 `which(...)` finds the row number where the field name is `Sex_Code`.

Each line like `data_dictionary[row, "Meaning"] <- ...` is subassignment: it changes one cell inside the data frame.

## Store and share your data dictionary

Save your data dictionary as a CSV file in your course folder. This will allow you to share it with others and use it for future reference.

```{r, eval=FALSE}
write.csv(data_dictionary, "field_sightings_dictionary.csv", row.names = FALSE)
```

# Chapter Terms

**Data Dictionary**: A companion document (often a table) that defines each variable in a dataset—its name, meaning/description, data type, allowable values or coding scheme, missing-value codes, and any notes needed to interpret the data correctly and reuse it later.

**Subassignment**: Replacing or updating part of an existing object by putting an index on the left-hand side of `<-` (e.g., `x[i] <- value` or `df[rows, "col"] <- value`). R changes only the indexed elements by using a replacement function (like `[<-`, `[[<-`, or `$<-`) and then stores the updated object back into the same name.


# 📝 Practice Space

Use the data dictionary template you created to fill in the meanings, units, codes/values, missing values, valid values, and any notes or relationships for the fields specified in the tasks below. Save your completed data dictionary as a CSV file in your course folder.

**Steps**

1. Identify the 

2. Review the code provided in the section **Subassignment**. 


## Task 1

Fill in the blanks to document the `Age_Code` field in your data dictionary. Use the information from the dataset and the data collection story to complete each section.

🎯 **Goal** "Age_Code"


```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
row <- which(data_dictionary$Field_Name == ________)

data_dictionary[row, "Meaning"] <- "_______________________"
data_dictionary[row, "Units"] <- "_________________________"
data_dictionary[row, "Codes_Values"] <- "__________________"
data_dictionary[row, "Missing_Values"] <- "________________"
data_dictionary[row, "Valid_Values"] <- "__________________"
data_dictionary[row, "Notes_Relationships"] <- "___________"
print(data_dictionary)
```

🗣 Comment about how this code changed `data_dictionary`


## Task 2

🎯 **Goal** "Band_Color"


```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
row <- which(data_dictionary$Field_Name == ________)

data_dictionary[row, "Meaning"] <- "_______________________"
data_dictionary[row, "Units"] <- "_________________________"
data_dictionary[row, "Codes_Values"] <- "__________________"
data_dictionary[row, "Missing_Values"] <- "________________"
data_dictionary[row, "Valid_Values"] <- "__________________"
data_dictionary[row, "Notes_Relationships"] <- "___________"
print(data_dictionary)
```

🗣 Comment about how this code changed `data_dictionary`

## Task 3

🎯 **Goal** "Status"

```{r, eval=FALSE}
# ⚡ Type this in your R Notebook
row <- which(data_dictionary$Field_Name == ________)

data_dictionary[row, "Meaning"] <- "_______________________"
data_dictionary[row, "Units"] <- "_________________________"
data_dictionary[row, "Codes_Values"] <- "__________________"
data_dictionary[row, "Missing_Values"] <- "________________"
data_dictionary[row, "Valid_Values"] <- "__________________"
data_dictionary[row, "Notes_Relationships"] <- "___________"
print(data_dictionary)
```

🗣 Comment about how this code changed `data_dictionary`

## Task 4

🎯 **Goal**

Save your completed data dictionary as a CSV file in your course folder. This will allow you to share it with others and use it for future reference.
