---
title: "Lesson 4"
output: html_notebook
---

# Warm up

I can type freely with out using the "#" here

```{r}
# This is an R script Code Chunk
?gsub
```

# NA vs NaN

I know that I don't need to use the # but when I do, it makes a heading. Check out the outline.

## sub header

```{r}
0/0
```

The NaN is not a number as anything over zero is not a number.

```{r}
NA
```

NA is a missing value in a vector or cell of a dataframe.

# Detecting Missing Values

```{r}
Vector_NA <- c(3, 7, NA, 12)
```
This created the numeric vector Vector_NA

```{r}
anyNA(Vector_NA)

```

This result TRUE means that there is NA in Vector_NA

# Locating Missing Value

```{r}
is.na(Vector_NA)
```

This returned FALSE in the first, second, and fourth position and TRUE in the third position which corresponds with the order the values were typed in when storing Vector_NA.

```{r}
which(is.na(Vector_NA))
```

which() returns the index number for the TRUE value(s) that is.na() tests. The number 3 means that the third position is TRUE.

We need to test this with a larger vector

```{r}
set.seed(100)
LargeVec <- sample(c(1:10, rep(NA, 10)))
LargeVec
```

```{r}
which(is.na(LargeVec))
```

This returns the index positions for NA.

# Summarizing "Missingness"

```{r}
sum(is.na(LargeVec))
```

Memo:

```{r}
mean(is.na(LargeVec))
```

Memo:

# Action Strategies
```{r}
LargeVec
```

```{r}
clean_vec <- LargeVec[!is.na(LargeVec)]
clean_vec
```

All of the NA values have been removed and it is stored in a new vector called clean_vec.

```{r}
impute <- LargeVec
impute
```

```{r}
impute[is.na(impute)] <- mean(impute, na.rm = TRUE)
```

```{r}
impute
```



