---
title: "Lesson 9"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 📦 Packages

Install once (if needed): `install.packages("dplyr")`; `install.packages("dslabs")`

Load the packages at the start of every session:

```{r , eval=FALSE}
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
```

# Warm‑Up

## Logical Test Review

🎯 Goal: To write a simple logical test for numbers one through five showing TRUE for when the numbers are larger than 3.

💬 Prediction: I expect 4 and 5 to be `TRUE`.

```{r , eval=FALSE}
# ⚡Code: TRUE appears where the test is met.
c(1, 2, 3, 4, 5) > 3
```

✅ Check / 🗣 Comment:  Did the output met your expectation?

## `sample()` Review

Recall `sample()` is a `base-R` function that randomly selects elements from a vector.

🎯 Goal: Randomly select 5 numbers from 1 to 10.

```{r , eval=FALSE}
# ⚡Code: 
set.seed(99)  # Ensures reproducibility
sample(1:10, 5)  # Sample 5 numbers from 1 to
```

✅ Check / 🗣 Comment: Confirm the output length is 5 and numbers are between 1 and 10.

# `sample_n()` 

📜 The SYNTAX:`sample_n(data, size, replace = FALSE, weight = NULL)`

🎯 Goal: Randomly select 5 rows from the `mtcars` dataset.

```{r , eval=FALSE}
# ⚡Code: 
sample_n(mtcars, 5)  # Sample 5 rows from mtcars data frame
```

```{r , eval=FALSE}
# ⚡Code: 
mtcars2 <- mtcars %>% sample_n(5)  # Using with pipe
```

✅ Check / 🗣 Comment: Confirm the output has 5 rows and all original columns.

🤔 Reflect: Why are the selected row numbers different each time you run the code?

We'll come back to `sample_n()` later in this lesson. 

# `ifelse()` vs `case_when()`

## `ifelse()` tests 

📜 The SYNTAX: `ifelse(test, yes, no)`

- *test*  – logical vector (`TRUE`/`FALSE`/`NA`)

- *yes*   – value to insert where test is `TRUE`

- *no*    – value to insert where test is `FALSE` or `NA`

🎯 Goal: Label numbers > 3 as "Large" and others as "Small".

💬 Prediction: I expect to see a vector of "Small" and "Large" labels. Specifically, I expect only 4 and 5 to be "Large".

```{r , eval=FALSE}
ifelse(test = c(1, 2, 3, 4, 5) > 3, yes = "Large", no = "Small")
```

✅ Check / 🗣 Comment: Confirm the output positions for 1:3 are "Small"  and 4:5 are "Large".

## `case_when()` tests

dplyr::case_when() can do exactly what `ifelse()` is doing. 

📜 The SYNTAX: `case_when(condition1 ~ value1, condition2 ~ value2, TRUE ~ default_value)`

The `~` symbol means "maps to" or "yields".

🎯 Goal: Label numbers > 3 as "Large" and others as "Small".

💬 Prediction: I expect to see a vector of "Small" and "Large" labels. Specifically, I expect only 4 and 5 to be "Large".

```{r , eval=FALSE}
case_when(c(1, 2, 3, 4, 5) > 3 ~ "Large", TRUE ~ "Small")
```

✅ Check / 🗣 Comment: Confirm the output positions for 1:3 are "Small"  and 4:5 are "Large"

So, what's the point? Why do we need both `ifelse()` and `case_when()`? `case_when()` is usually clearer once you have 3+ categories.


## Multi‑Level Recoding 

Nested `ifelse()`

🎯 Goal: Re-code numbers 1 to 5 as "One", "Two", "Three", "Four", or "Five".

💬 Prediction: I expect to see a vector of the words "One" through "Five" corresponding to the numbers 1 through 5.

```{r , eval=FALSE}
# ⚡Code: 
object2 <- c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5)
length(object2) # Check the length of object2
```


```{r , eval=FALSE}
# ⚡Code
ifelse(object2 == 1, "One",
       ifelse(object2 == 2, "Two",
              ifelse(object2 == 3, "Three",
                     ifelse(object2 == 4, "Four", "Five"))))
```


✅ Check / 🗣 Comment:  Confirm the output length is 10 and the labels match the numbers.

Nested `case_when()`

🎯 Goal: Re-code numbers 1 to 5 as "One", "Two", "Three", "Four", or "Five".

💬 Prediction: The `case_when()` output should be identical in length and order to the nested `ifelse()` version.

```{r , eval=FALSE}
# ⚡Code
case_when(object2 == 1 ~ "One",
          object2 == 2 ~ "Two",
          object2 == 3 ~ "Three",
          object2 == 4 ~ "Four",
          object2 == 5 ~ "Five")
```

✅ Check / 🗣 Comment:  Confirm the output is identical to the `ifelse()` version.

🪞 What I learned: `case_when()` is easier to read because the function name is not repeated at each line.

Apply this nested `case_when()` in Task 3 of the assignment.

# `mutate()`

`mutate()` is a function that adds new columns or modifies existing ones

📜 The SYNTAX: `mutate(data, new_col = value)`

## Overwrite Columns

### Copy A Column

🎯 Goal: Create a new column that is a copy of an existing column.

```{r , eval=FALSE}
# Preview Column Names
names(mtcars) # Check the column names
```


💬 Prediction: In a new data set called `mtcars2`, a **new** column named `new_column_copy` is created that is identical to the `am` column.

```{r , eval=FALSE}
# ⚡ Code: 
mtcars2 <- mtcars %>% 
  mutate(new_column_copy = am)
```

```{r , eval=FALSE}
# ✅ Verify:
names(mtcars2) # Check the column names again
```

🗣 Comment: Did you see the new column in the output?

```{r , eval=FALSE}
# ✅ Verify: is am == to n_column_copy?
mtcars2$am == mtcars2$new_column_copy
```


```{r , eval=FALSE}
# ✅ Verify: is am identical to new_column_copy?
identical(mtcars2$am, mtcars2$new_column_copy)
```

🗣 Comment: `mutate(new_column_copy = am)` created an identical column of `am` called `new_column_copy`

### Overwrite Original Column Values

🎯 Goal: Recode the `am` column in `mtcars` from 0/1 to "Automatic"/"Manual".

```{r , eval=FALSE}
# Check the unique values in the am column BEFORE
unique(mtcars$am) 
```

💬 Prediction: In a new data set called `cars_overwrite`, the `am` column is changed from 0/1 to "Automatic"/"Manual". In a second new data set called `cars_keep`, a **new** column named `transmission` is created that has "Automatic"/"Manual" values, while the original `am` column remains unchanged.

```{r , eval=FALSE}
# ⚡ Code: 
cars_overwrite <- mtcars %>%
  mutate(am = case_when(am == 0 ~ "Automatic", TRUE ~ "Manual"))
```

```{r , eval=FALSE}
# ✅ Verify:
unique(cars_overwrite$am)
```

### New Column & New Values

🎯 Goal: Create a new column `transmission` that recodes `am` from 0/1 to "Automatic"/"Manual", while keeping the original `am` column unchanged.

```{r , eval=FALSE}
# ⚡ Code: 
cars_keep <- mtcars %>%
  mutate(transmission = case_when(am == 0 ~ "Automatic", TRUE ~ "Manual"))
```

```{r , eval=FALSE}
# ✅ Verify:
cars_keep %>% select(am, transmission)
```

🗣 Comment: What is the difference between the code that creates `cars_overwrite` and `cars_keep`?

🪞 Metacognition: Why might you want to create a copy of a column when changing the values?

## Factors and Levels

In Lesson 2, we learned about factors for vectors. A factor is a categorical variable that can take on a limited number of values, called levels. For example, in the vector `tshirt_sizes_data`, the possible values are "S", "M", "L", and "XL".

```{r eval=FALSE}
# ⚡ Creating a character vector
tshirt_sizes_data <- c("M", "L", "S", "M", "L", "XL", "S")
# ⚡ Changing the character vector to a factor class with levels
tshirt_sizes <- factor(tshirt_sizes_data, levels = c("S", "M", "L", "XL"))
```

```{r eval=FALSE}
# ✅ Investigating your factor class with levels
unique(tshirt_sizes)
```
 
In the dataset `iris`, the `Species` column is a factor with three levels: "setosa", "versicolor", and "virginica".
 
```{r eval=FALSE}
data(iris)
?iris
```

```{r eval=FALSE}
# ✅Investigating the Species column
unique(iris$Species)
class(iris$Species)
levels(iris$Species)
```

### Overwrite with Factors

🎯 Goal: Take a 100-row sample from `movielens`, then overwrite the numeric `rating` column with an ordered factor: "Bad" if `< 4`, "Good" if `>= 4`, using `case_when()` inside `mutate()`.

💬 Prediction: Before: `rating` is numeric. After: `rating` is an ordered factor with levels `Bad < Good`. Row count stays 100; no `NA`s created.

📜 SYNTAX reminder:
`mutate(data, new_col = case_when(cond1 ~ value1, cond2 ~ value2, TRUE ~ default))`

`name_data <- data %>% mutate(new_col = case_when(cond1 ~ value1, cond2 ~ value2, TRUE ~ default))`

```{r , eval=FALSE}
data("movielens")
?movielens
```

```{r , eval=FALSE}
set.seed(99)

# Sample 100 movies from the movielens dataset
smaller_movies <- movielens %>% 
  sample_n(100) %>% 
  select(title, rating)

head(smaller_movies)
```

```{r , eval=FALSE}
# ✅ Checks (before)
names(smaller_movies)                # expect "title", "rating"
class(smaller_movies$rating)         # expect "numeric"
summary(smaller_movies$rating)       # quick distribution check
```

🎯 Goal: We'll apply `mutate()` to `smaller_movies` to recode the rating column to "Good" or "Bad".

```{r , eval=FALSE}
# ⚡ Overwrite numeric rating with an ordered factor via case_when()
movies_overwrite <- smaller_movies %>%
  mutate(
    rating = factor(
      case_when(
        rating >= 4 ~ "Good",
        rating < 2 ~ "Bad",
        TRUE        ~ "Ok"
      ),
      levels = c("Bad", "Ok", "Good"),
      ordered = TRUE
    )
  )
```


```{r , eval=FALSE}
# ✅ Checks (after)
names(movies_overwrite)              # still "title", "rating"
```


```{r , eval=FALSE}
# ✅ Checks (after)
is.ordered(movies_overwrite$rating)  # expect TRUE
```


```{r , eval=FALSE}
# ✅ Checks (after)
levels(movies_overwrite$rating)      # expect "Bad", "Ok", "Good"
```


```{r , eval=FALSE}
# ✅ Checks (after)
table(movies_overwrite$rating)       # counts in each level
```

🗣 Comment: Overwriting keeps the column name but replaces its contents and type (numeric → ordered factor). 

🪞 Metacognition: `case_when()` reads cleaner than nested `ifelse()` for multi-branch recodes and makes me specify a default. In the future, I could keep the original numeric ratings (e.g., `rating_num`) and add a new labeled column instead of overwriting, to preserve provenance.


```{r , eval=FALSE}
# ⚡ Overwrite numeric rating with an ordered factor via case_when()
movies_rename <- smaller_movies %>%
  mutate(
    rating_words = factor(
      case_when(
        rating >= 4 ~ "Good",
        rating < 2 ~ "Bad",
        TRUE        ~ "Ok"
      ),
      levels = c("Bad", "Ok", "Good"),
      ordered = TRUE
    )
  )
```

### Bonus Example

🎯 Goal: Create a new column rating_label in smaller_movies using `case_when()` that labels ratings `>= 4` as "Good" and the rest as "Bad". Keep the original numeric rating column.

💬 Prediction: After running, the data will still have 100 rows, keep both `title` and `rating`, and add `rating_label` (character or factor). No `NA`s should appear.

```{r , eval=FALSE}
# ✅ Checks (before)
names(smaller_movies)                # expect "title", "rating"
class(smaller_movies$rating)         # expect "numeric"
```

```{r , eval=FALSE}
# ⚡ Add a NEW column using case_when()
smaller_movies %>%
  mutate(
    rating_label = case_when(
      rating >= 4 ~ "Good",
      TRUE        ~ "Bad"
    )
  ) -> movies_newcol
```

🧐 NOTICE: What is happening here `-> movies_newcol`?

🛠Break Things! How can you reorganize this script? 

```{r , eval=FALSE}
# ✅ Checks (after)
names(movies_newcol)                 # expect "title", "rating", "rating_label"
class(movies_newcol$rating)          # expect "numeric"
class(movies_newcol$rating_label)    # expect "character" or "factor"
```
🗣 The code used -> movies_newcol, which assigns right-to-left.

🛠 Break Things! What happens with `head(movies_newcol) -> movies_newcol`?

🪞 Metacognition: I’ll stick with the usual `<-` to make my code easier to read

## Calculations

🚗 Using `mtcars`, let's create a column `l_per_100km` that converts from US units (miles per gallon) to metric units (liters per 100km). 235.215 is the conversion factor between these units

🎯 Goal: Identify the most fuel-efficient car in L/100 km terms.

💬 Prediction: The new column `l_per_100km` will be added to `mtcars_units`, and the most fuel-efficient car will have the lowest value in this column.


```{r , eval=FALSE}
# ⚡ Code
mtcars_units <- mtcars %>%
  mutate(l_per_100km = round(235.215 / mpg, 1))
```

```{r , eval=FALSE}
# ✅ Check
head(mtcars_units)
```

👀 Check‑in: Which car is most fuel‑efficient in l/100 km terms?

```{r , eval=FALSE}
# Find the most fuel-efficient (smallest L/100km)
best_idx <- which.min(mtcars_units$l_per_100km)
best_car <- rownames(mtcars_units)[best_idx]
best_value <- mtcars_units$l_per_100km[best_idx]
```

```{r , eval=FALSE}
"best car"
best_car
```

```{r , eval=FALSE}
"best value"
best_value
```

🪞 Metacognition: Unit conversions are easy to keep alongside original units with `mutate()` (don’t overwrite).

# 📝 Practice Space

✍️ Practice: Recode the mpg column. Use `mutate()` and `case_when()`

Fill in the Blanks. 

🎯 Goal: Create `mpg_class` with three labels using `case_when()`:
"High MPG" if `mpg > 25`; "Low MPG" if `mpg < 15`; otherwise "Medium MPG".

```{r , eval=FALSE}
# ⚡ Code
mtcars %>%
  ______(mpg_class = _____(
    ___ > 25 ~ "High MPG",
    mpg < 15 ~ "_______",
    TRUE     ~ "Medium MPG"
  )) -> mpg_recode
```

```{r , eval=FALSE}
# ✅ Checks
table(mpg_recode$____)  # expect 3 levels
```

# ➤ Assignment

Replace each ____ placeholder with working code or a short written answer. Run each section; be sure the requested objects appear in the Environment. When finished, submit lesson notes with assignment in one script (.Rmd / .qmd) along with the `.RData` workspace.

Template (use around every code chunk):

🎯/❓ Goal. One sentence stating the outcome.

💬 Prediction. (Optional) What you expect to see before you run code.

```{r}
#⚡Code attempts.

# your code here
```

✅Checks (before/after). Show what you’re verifying (names, head(), table(), class(), etc.).

🗣/🪞What I learned / Next step. 1–2 sentences.

Grader Key — Assignment 9

### Task 1 — Libraries (0.25)

Reference code
```{r}
library(dplyr)
library(dslabs)
```

Checks
```{r}
"package:dplyr"  %in% search()
"package:dslabs" %in% search()
```

Deduct if: packages not loaded; code errors.

### Task 2 — Recode ranges (overwrite) (0.75)

Goal: Overwrite quakes$mag with an ordered factor of exactly 3 levels using mutate() + case_when().

Reference code (thresholds may vary; must yield 3 levels)
```{r}
data(quakes)

quake_recode <- quakes %>%
  dplyr::mutate(
    mag = dplyr::case_when(
      mag < 4.6 ~ "Low",
      mag < 5.0 ~ "Mid",
      TRUE      ~ "High"
    ) %>% factor(levels = c("Low","Mid","High"), ordered = TRUE)
  )
```

Checks
```{r}
exists("quake_recode")
is.ordered(quake_recode$mag)
identical(levels(quake_recode$mag), c("Low","Mid","High"))
length(table(quake_recode$mag)) == 3
# Overwrite confirmed (no extra mag column; class changed from numeric to ordered factor)
is.factor(quake_recode$mag)
```

Acceptable variants: different sensible cutpoints; using round() on means before cut is fine.
Deduct if: created a new column instead of overwriting; not ordered; not 3 levels; used ifelse() instead of case_when().

### Task 3 — Recode categories (new column) (0.5)

Goal: Keep original temp; add temp_group based on temperature ranges.

```{r}
data(stars)  # dslabs

stars_recode <- stars %>%
  dplyr::mutate(
    temp_group = dplyr::case_when(
      temp < 4000 ~ "Cool",
      temp < 7500 ~ "Warm",
      TRUE        ~ "Hot"
    )
  )
```

Checks
```{r}
exists("stars_recode")
all(c("temp","temp_group") %in% names(stars_recode))
head(dplyr::select(stars_recode, temp, temp_group))
```

Acceptable variants: different clear range labels such as "Low", "Medium", and "High".
Deduct if: overwrote temp; did not create a new column; used direct value-to-label matching instead of range-based conditions; omitted the default rule.

### Task 4 — Unit conversion (0.5)

Goal: Add height_cm = height * 2.54; keep original height.

Reference code
```{r}
data(heights)

international_friendly_heights <- heights %>%
  dplyr::mutate(height_cm = height * 2.54)
```

Checks
```{r}
exists("international_friendly_heights")
all(c("height","height_cm") %in% names(international_friendly_heights))
all.equal(
  international_friendly_heights$height_cm,
  international_friendly_heights$height * 2.54
)
```

Reflection (credit in this task): Should explain why add, not overwrite (provenance, preserving original units for reproducibility and future checks).
Deduct if: overwrote height; wrong multiplier; reflection absent or off-topic.

Technical quality & submission (0.5)

Script runs cleanly in a fresh session (no hidden dependencies).

Both files submitted: script (.Rmd/.qmd) + .RData.

Reasonable organization; code chunks runnable top-to-bottom.

Deduct if: missing file(s); script fails on clean run.


# Save and Upload

1. You will be submitting **both** the R Markdown | Quarto Document and the workspace file. The workspace file saves all the objects in your environment that you created in this lesson. You can save the workspace by running the following command in a code chunk of the R Markdown | Quarto Document document:

```{r , eval=FALSE}
save.image("Assignment9_Workspace.RData")
```

Or you can click the "Save Workspace" button in the Environment pane. 

💾 **Always save the R documents before closing.**

2. Find the assignment in this week's module in Canvas and upload **both** the RMD and the workspace file.

# Today you practiced:

- Practiced logical tests and the structure of `case_when()`.

- Created binary and multi‑level recodes.

- Used `mutate()` to overwrite or create columns.

- Performed inline calculations for unit conversion.

🛠 Continue experimenting. Recoding and mutation are the building blocks of clean, analysis‑ready data!
