---
title: "Example Vector Lab R Notebook"
output: html_notebook
---

# Group Members

```{r}
member_name <- c("Name1", "Name2", "Name3", "Name4")
role <- c("Student A", "Student B", "Student C", "Reporter")
```


# Numeric vector (length = 10)

```{r}
numeric <- c(3.14, -2.5, 0, 7.25, 10.01, 4.4, -9.9, 2.0, 6.66, 1.23)

## ✅ Check the length of your vector
length(numeric)
```

🗣 Comment: `numeric` is a numeric (double) vector containing real-valued numbers (decimals allowed).



# Integer vector (length = 4) — must use the L suffix

```{r}
integer <- c(7L, -2L, 0L, 15L)

## ✅ Check the length of your vector
length(integer)
```

🗣 Comment: `integer` is an integer vector; values are whole numbers stored as integer type (with the L suffix).



# Logical vector (length = 4) — built with logical operators
```{r}
logical <- c(5 > 3, 10 <= 4, 7 == 7, 2 != 2)

## ✅ Check the length or class of your vector
class(logical)
```
🗣 Comment: `logical` is a logical vector; it contains TRUE/FALSE values created from comparisons.


# Character vectors
```{r}
name <- c("Ava", "Ava", "Noah", "Mia", "Ava", "Liam", "Mia", "Noah", "Ava", "Mia")
group_name <- c("Alpha", "Beta", "Gamma", "Delta")
at_handle <- c("@data_dan", "@stats_sam", "@viz_vic", "@code_casey")
item <- c("laptop", "mug", "notebook", "pen", "backpack", "charger", "headphones", "stapler", "water bottle", "sticky notes")
activity <- c("study", "study", "run", "cook", "read", "study", "code", "read", "code", "study")
location <- c("library", "coffee shop", "park", "museum")
call_to_action <- c("Keep going.", "Try again.", "Ask questions.", "Stay curious.")
emotion <- c("excited", "nervous", "focused", "happy", "happy", "motivated", "tired", "focused", "happy", "excited")
retro_game <- c("Pac-Man", "Tetris", "Space Invaders", "Donkey Kong")
```


# Factor class

Part 1: convert a character vector to a factor (overwrite original)
```{r}
# Step 1: Check the class
class(emotion)

# Step 2: Change class to factor (overwrite)
emotion <- as.factor(emotion)

# Step 3: Verify the class changed
class(emotion)
```

# Factor class

```{r}
# Part 2: create your own factor vector with ranked levels
social_media <- c(
  "LinkedIn", "Instagram", "X", "TikTok", "LinkedIn",
  "Instagram", "LinkedIn", "TikTok", "X", "LinkedIn"
)

social_media_platform <- factor(
  social_media,
  levels = c("LinkedIn", "Instagram", "X", "TikTok")
)

# ✅ Investigating your factor class with levels
print(social_media_platform)
```


```{r}
# ⚡  Copy and paste this code into your R Notebook
# Function to pick a random element from a vector
pick <- function(x) sample(x, 1)

# ---------- Story 1 ----------
story1 <- paste(
  "After a long week,", pick(name), "unplugged by trying", pick(activity),
  "with", pick(group_name), "at their favorite spot,", paste0(pick(location), "."),
  "Vibe rating:", paste0(pick(numeric), "."),
  "The ", pick(activity), "was going so well, but then a crucial missing", paste0(pick(item), " was discovered."),
  "Going with the flow was", ifelse(pick(logical), "reasonable.", "unreasonable."),
  "Reminding each other", paste0(pick(call_to_action), "."),
  "They made a plan for Episode", paste0(pick(integer), ","),
  "closing with a", paste0(pick(emotion), " mood.")
)

print(story1)
# ---------- Story 2 ----------
# Using Factor Level to pick 
selected_index <- sample(1:length(social_media_platform), 1)
selected_platform <- social_media_platform[selected_index]

story2 <- paste(
  "On a mission for a fashion comeback,", pick(name), "found", pick(item),
  "at", pick(location), "that screamed Y2K.",
  "They spent a total of", paste0(pick(numeric), " dollars"),
  "but they", ifelse(pick(logical), "remembered", "forgot"), "to get a receipt!",
  "After posting a 'before' shot on", paste0(pick(selected_platform), ","),
  "the post went viral", ifelse(pick(logical), "by midnight.", "by the next day."),
  "Their retro level was officially", paste0(pick(retro_game), " status,"),
  "and they immediately upcycled the item online for", paste0(pick(numeric), " dollars"),  
  "and sold in", pick(integer), "minutes.", 
  "The community response was so strong, they had this final message for their followers:", 
  paste0(pick(call_to_action), ".")
)
print(story2)
```



# Dataframe

🎯 Build a tiny table you can reference as “credits” under the stories.

```{r}
# ⚡  Copy and paste this code into your R Notebook
people <- data.frame(
  member_name = member_name,
  role        = role,
  likes_story    = logical,
  dice_roll       = integer, 
  location = location
)

print(people)
```

🗣 Comment: `people` is a data frame combining group member names, roles, their likes_it logical values, and scores as integers.
