---
title: "Chapter 3 Activity (Online)"
author: "by Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactive scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
encoding: UTF-8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Chapter 3 Activity"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
 
Or
title: "Chapter 3 Activity"
author: "by Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "SPring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactive scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```


# Group Activity: Vector Playlab 🎮 (Online)

This is a team-based challenge designed to deepen your understanding of vectors in R and to introduce a challenging research skill: collaborative coding. You will build vectors of different types, convert at least one vector to a factor, generate two MadLib stories, and combine your vectors into a data frame.

## Activity Overview

In the next few minutes you will learn how to:

- 🔨 Build and inspect vectors (numeric, integer, logical, character)

- 🔧 Use logical operators to create logical vectors

- 🪄 Understand factor class + levels

- 💪 Practice reproducible work in an R Notebook with annotations

- 👨🏻👩🏼👧🏽👦🏾 Experience the practical challenges of writing code and debugging as a team

# Activity Prep

Each group member will write their section locally (in R Studio). The "Reporter" assembles one final shared notebook for submission.

## Materials

What you will use

- RStudio

- One final R Notebook (`.Rmd`) submitted by your "Reporter"

- Canvas Group space for collaboration (Group Discussion and/or Group Messages)

Do not post your code in the whole-class discussion. That space is only for forming groups.

# Find Your Group

## Step One: Join your Canvas group

Go to People → Vector Playlab Groups and join one group with fewer than 4 members. Groups should have approximately 4 people 👨🏻👩🏼👧🏽👦🏾. 

## Step Two: Assign Roles

Decide roles in your group. Roles can rotate, but one person must be the Reporter.

- **Student A**: numeric + integer vectors

- **Student B**: logical vector + short memo explaining each comparison

- **Student C**: character vectors + factor conversion

- **Reporter**(required): story + data frame + Run All check + final submission
  
## Step Three: Students A, B, and C

Each person works locally on their section of the code.

1. Complete your section in your own R Studio so the code runs without error. Be creative! You can use any resource you'd like to help you come up with the elements in the vectors (e.g., internet search, brainstorming with AI, etc.).

2. Post one reply in your **Group Discussion** with:

(a) Code (your final working chunk) and 

(b) Memo (2–4 sentences: what you tried, what broke, how you verified/fixed it, something you learned and/or resource used).

## Step Four: Reporter

1. In RStudio: File → New File → R Notebook

Save it in your course folder as: `vector_playlab_groupXX.Rmd` *(Replace XX with your group number/name.)*

2. At the top of the notebook add:

```{r eval=FALSE}
# Vector Playlab Group Members + Roles
member_name <- c("Name1", "Name2", "Name3", "Name4")
role <- c("Student A", "Student B", "Student C", "Reporter")
```

3. Copy and paste code snippets from each group member into the appropriate sections of the R Notebook.

4. Complete the Reporter section: story + data frame + Run All check 

5. Final submission

# Rules

1. Naming rules

- Use the exact object names shown in this handout (for example: `numeric`, `integer`, `logical`, `name`, `group_name`, etc.).

2. Reproducibility rule

Your notebook must run top-to-bottom. That means:

- You must create objects before you use them.

- If something errors, fix it and document the fix with a memo.

3. Memoing rule (transparency)

Add short written notes in your notebook that describe:

- what you tried,

- what worked or didn’t work,

- and what you changed to fix a problem (including any help/resources used).

The length of `member_name` and `role` should match and correspond to each other.

# Vectors

**Rules**: Name your vectors *exactly* as listed. Follow the rules for each type. Element duplication is permissible. 

## Numeric vector

🎯  Create a vector called `numeric` → all real numbers (length = 10)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
numeric <- c( __ , __ , __ , __ , __ , __ , __ , __ , __ , __ )
```
```{r eval=FALSE}
## ✅ Check the length of your vector
_____(numeric)
```

🗣 Comment: Describe `numeric` vector type.

## Integer vector

🎯 `integer` → must use the L suffix (length = 4 or group size)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
integer <- c( __ _ , __ _ , __ _ , __ _ ) # length = 4 or group size
```
```{r eval=FALSE}
## ✅ Check the length of your vector
_____(integer)
```
🗣 Comment: Describe `integer` vector type. 

## Logical vector

🎯  Create a vector called `logical` → build with **logical operators** (length = 4 or group size).

The logical operators are:

- `>` greater than

- `<` less than

- `==` equal to

- `>=` greater than or equal to

- `<=` less than or equal to

- `!=` not equal to

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
logical <- c( __ , __ , __ , __ ) # length = 4 or group size
```
```{r eval=FALSE}
## ✅ Check the length or class of your vector
_____(logical)
```

🗣 Comment: Describe `logical` vector type.

## Character vectors

➤  Make each character vector with words or phrases of your choice:

- 🎯  Create a vector called `name` → *any* name (length = 10); Element duplication is recommended.

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
name <- c( __ , __ , __ , __ , __ , __ , __ , __ , __ , __ )
```

- 🎯  Create a vector called `group_name` → name of your group (length = 4 or group size)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
group_name <- c( __ , __ , __ , __ ) # length = 4 or group size
```

- 🎯  Create a vector called `at_handle` → any social media handle (length = 4 or group size)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
at_handle <- c( __ , __ , __ , __ ) # length = 4 or group size
```

- 🎯  Create a vector called `item` → any kind of thing (length = 10)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
item <- c( __ , __ , __ , __ , __ , __ , __ , __ , __ , __ )
```

- 🎯  Create a vector called `activity` → something you do (length = 10); Element duplication is recommended.

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
activity <- c( __ , __ , __ , __ , __ , __ , __ , __ , __ , __ )
```

- 🎯  Create a vector called `location` → any place to go (real or imagined) (length = 4 or group size)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
location <- c( __ , __ , __ , __ ) # length = 4 or group size
```
- 🎯  Create a vector called `call_to_action` → a catch phrase or statement (length = 4 or group size)

- 🎯  Create a vector called `emotion` → a mood or feeling (length = 10); Element duplication is recommended.

- 🎯  Create a vector called `retro_game` → any old-school game title (length = 4 or group size)

## Factor class

### Part 1

🎯 Change a character vector(s) into factor class. (5 minutes)

➤  Step 1: Check the Class

🎯 Select a character vector (e.g., `name`, `activity`, `emotion`) and use a function `class()` to check the class of the vector.

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
class( __ )
```

➤  Step 2: Change Class to Factor

🎯 Use the `as.factor()` function to change *at least one* character vector into factor class. Overwrite the original character vector with the factor class vector.

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
__ <- _____( __ )
```

➤  Step 3: Check the Class

🎯 Verify that the class of the vector(s) have changes.

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
class( __ )
```

### Part 2

(5 minutes)

➤  Make a Factor Vector

Here is an *example* to follow: 

```{r eval=FALSE}
# ⚡ Creating a character vector
tshirt_sizes_data <- c("M", "L", "S", "M", "L", "XL", "S")
# ⚡ Changing the character vector to a factor class with levels
tshirt_sizes <- factor(tshirt_sizes_data, levels = c("S", "M", "L", "XL"))
# ✅ Investigating your factor class with levels
print(tshirt_sizes)
```

🎯 Create your own factor vector called `social_media_platform`. Pick at least 4 social media / online networking platforms to create your vector. Element duplication is recommended. Rank the platforms using the `levels`. (5 minutes)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
social_media <- c( __ , __ , __ , __ , __ , __ , __ , __ , __ , __ )
social_media_platform <- factor( __ , levels = c( __ , __ , __ , __ ) )
# ✅ Investigating your factor class with levels
print( social_media_platform )
```


# Story

➤  Copy and paste the following code and read your Mad Lib Story!

➤  Before running the story code, confirm these objects exist in your Environment: `numeric`, `integer`, `logical`, `name`, `group_name`, `at_handle`, `item`, `activity`, `location`, `call_to_action`, `emotion`, `retro_game`, `social_media_platform`

```{r , eval=FALSE}
# ⚡  Copy and paste this code into your R Notebook
# Function to pick a random element from a vector
pick <- function(x) sample(x, 1)

# ---------- Story 1 ----------
story1 <- paste(
  "After a long week,", pick(name), "unplugged by trying", pick(activity),
  "with", pick(group_name), "at their favorite spot,", paste0(pick(location), "."),
  "Vibe rating:", paste0(pick(numeric), "."),
  "The ", pick(activity), "was going so well, but then a crucial missing", paste0(pick(item), " was discovered."),
  "Going with the flow was", ifelse(pick(logical), "reasonable.", "unreasonable."),
  "Reminding each other", paste0(pick(call_to_action), "."),
  "They made a plan for Episode", paste0(pick(integer), ","),
  "closing with a", paste0(pick(emotion), " mood.")
)

print(story1)
# ---------- Story 2 ----------
# Using Factor Level to pick 
selected_index <- sample(1:length(social_media_platform), 1)
selected_platform <- social_media_platform[selected_index]

story2 <- paste(
  "On a mission for a fashion comeback,", pick(name), "found", pick(item),
  "at", pick(location), "that screamed Y2K.",
  "They spent a total of", paste0(pick(numeric), " dollars"),
  "but they", ifelse(pick(logical), "remembered", "forgot"), "to get a receipt!",
  "After posting a 'before' shot on", paste0(pick(selected_platform), ","),
  "the post went viral", ifelse(pick(logical), "by midnight.", "by the next day."),
  "Their retro level was officially", paste0(pick(retro_game), " status,"),
  "and they immediately upcycled the item online for", paste0(pick(numeric), " dollars"),  
  "and sold in", pick(integer), "minutes.", 
  "The community response was so strong, they had this final message for their followers:", 
  paste0(pick(call_to_action), ".")
)
print(story2)
```


# Dataframe

🎯 Build a tiny table you can reference as “credits” under the stories.

```{r eval=FALSE}
# ⚡  Copy and paste this code into your R Notebook
people <- data.frame(
  member_name = member_name,
  role        = role,
  likes_story    = logical,
  points       = integer, 
  location = location
)

print(people)
```

🧐 **Notice:** how the vectors line up, associating elements in each vector by row position (1st with 1st, 2nd with 2nd, etc.). The values in this table are made up and only for practicing how data frames are built. *They are not tied to grades, performance, or evaluation in any way.*

## Reporter: Final Steps

👉 The Reporter role submit the R notebook on Canvas.

- Sweep environment

- run all code chunks

- save R Notebook `vector_playlab_groupXX.Rmd` *(Replace XX with your group number/name.)*

- save environment `vector_playlab_groupXX.RData`

- upload both`.Rmd` and `.RData` to the assignment page

# Today you practiced

In this activity, your group practiced:

- Creating vectors of different types (numeric, integer, character, and logical).

- Combining vectors into a data frame.

Your work today previewed how R organizes and manages different data structures.

 
