---
title: "Chapter 3 Activity (In-Class)"
author: "by Lorraine Gaudio"
date:   "`r paste('Version:', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactive scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
encoding: UTF-8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Chapter 3 Activity"
author: "by Lorraine Gaudio"
date:   "`r paste('Version:', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
 
Or
title: "Chapter 3 Activity"
author: "by Lorraine Gaudio"
date:   "`r paste('Version:', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactive scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```


# Group Activity: Vector Playlab 🎮

Today’s activity is a group challenge to deepen your understanding of vectors in R  and to see what happens when you push them into a functions and a data frame. Work together, taking turns typing code on one computer, and discuss your results. By the end of this activity, you will be able to create and inspect vectors of different types, and organize data into a data frame.

Each member of your group will type different vector types on one computer. Together you will create a mini "Mad Lib" story using your vectors.

## Activity Overview

In the next thirty minutes you will:

- 🔍 Differentiate between numeric and integer vector types 

- 📝 Explain how R treats vector types

- 🔨 Construct and manipulate vectors

- 👥 Collaborate as a team to complete a coding challenge


# Activity Prep

To prep for our activity today think on how you would answer the following prompts. Write your answers on a piece of paper, type it in your computer notes, or simply think about how you would answer these prompts. 

1. Your name

2. Another random person's name

3. Dream team name (group name)

4. An `@handle`(any social media handle)

5. An activity (something you do)

6. Location (any place to go real or imagined)

7. Call to action (a catch phrase or statement)

8. An outcome (a phrase describing a result "They lived happily ever after")

9. An emotion

10. Old Game (game title from the past)

# Research Team Creation

Turn to your neighbors and negotiate a collaboration agreement between 3 other people. There should be four people in each group, each with a specific role.

## Form Groups

 Groups should have approximately 4 people 👨🏻👩🏼👧🏽👦🏾. 

## Assign Roles

Decide roles in your group. Roles can rotate, but one person must be the Reporter.

- **Student A**: types **numeric**, **integer**, + **logical** vectors

- **Student B**: types **character** vectors 

- **Student C**: types **factor** conversion

- **Reporter**: types `member_name` and `role`, copy/paste **story** + **data frame** + Run All check + final **submission**

# Rules

1. **Naming rules**

- Use the exact object names shown in this handout (for example: `numeric`, `integer`, `logical`, `name`, `group_name`, etc.).

2. **Reproducibility rule**

Your notebook must run top-to-bottom. That means:

- You must create objects before you use them.

- If something errors, fix it and document the fix with a memo.

3. **Memoing rule** (transparency)

Add short written notes in your notebook that describe:

- what you tried,

- what worked or didn’t work,

- and what you changed to fix a problem (including any help/resources used).

# Setup

## Reporter

➤  Step 1

 In RStudio: File → New File → R Notebook

➤  Step 2

Save the R Notebook in your course folder as: `vector_playlab_groupXX.Rmd` *(Replace XX with your group name.)*

➤  Step 3

At the top of the notebook add a code chunk that creates two vectors: `member_name` and `role`.

```{r eval=FALSE}
# Vector Playlab Group Members + Roles
member_name <- c("Name1", "Name2", "Name3", "Name4")
role <- c("Student A", "Student B", "Student C", "Reporter")
```

✅ The *length* of `member_name` and `role` should match and correspond to each other.

# Vectors
  
## Students A, B, and C

Work together to complete your assigned sections under the Vectors heading. 

- Pass the keyboard around. Each student should type their code into a code chunk the R Notebook. 

- Be creative! Everyone in the group contributes element ideas.

- Document discoveries and transparency (e.g., resources used) outside of code chunks.

**Rules**: Name each vectors *exactly* as listed. Follow the rules for each type. Element duplication is permissible. (15 Minutes Maximum)

## Student A: Numeric, Integer, + logical Vectors

### Numeric vector

➤  Step 1

Make each numeric vector with values of the group's choice.

🎯  Create a vector called `numeric` → all real numbers (length = 10); Element duplication is welcome.

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
numeric <- c( __ , __ , __ , __ , __ , __ , __ , __ , __ , __ )
```
```{r eval=FALSE}
## ✅ Check the length of your vector
_____(numeric)
```

🗣 Comment: Describe `numeric` vector type.

### Integer vector

Make each Integer vector with values of the group's choice.

➤  Step 2

🎯 `integer` → must use the L suffix (length = 4 *or group size*)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
integer <- c( __ _ , __ _ , __ _ , __ _ )
```
```{r eval=FALSE}
## ✅ Check the length of your vector
_____(integer)
```

🗣 Comment: Describe `integer` vector type. 


### Logical vector

➤  Step 3

🎯  Create a vector called `logical` → build with **logical operators** (length = 4 *or group size*).

    The logical operators are:

    - `>` greater than

    - `<` less than

    - `==` equal to

    - `>=` greater than or equal to

    - `<=` less than or equal to

    - `!=` not equal to

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
logical <- c( __ , __ , __ , __ )
```
```{r eval=FALSE}
## ✅ Check the length or class of your vector
_____(logical)
```

🗣 Comment: Describe `logical` vector type.

## Student B: Character Vectors

### Character vectors

➤  Step 1

Make each character vector with words or phrases of the group's choice.

- 🎯  Create a vector called `name` → *any* name (length = 10); Element duplication is recommended.

- 🎯  Create a vector called `group_name` → name of your group (length = 4 *or group size*)

- 🎯  Create a vector called `at_handle` → any social media handle (length = 4 *or group size*)

- 🎯  Create a vector called `item` → *any* kind of thing (length = 10); Element duplication is welcome.

- 🎯  Create a vector called `activity` → something you do (length = 10); Element duplication is welcome.

- 🎯  Create a vector called `location` → any place to go (real or imagined) (length = 4 *or group size*)

- 🎯  Create a vector called `call_to_action` → a catch phrase or statement (length = 4 *or group size*)

- 🎯  Create a vector called `outcomes` → phrase describing a result (length = 4 *or group size*)

- 🎯  Create a vector called `emotion` → a mood, emotion, or feeling (length = 10); Element duplication is welcome.

- 🎯  Create a vector called `retro_game` → any old-school game title (length = 4 *or group size*)

## Student C: Factor class

### Change to Factor Class

Change a character vector(s) into factor class. (5 minutes)

➤  Step 1

Check the Class

🎯 Select a character vector (e.g., `name`, `activity`, `emotion`) 

🎯 Use a function `class()` to check the class of the vector.

```{r eval=FALSE}
## ✅ Check the class of your selected vector
# ⚡ Type this in your R Notebook, Fill in the blanks
class( __ )
```

➤  Step 2

Change Class to Factor

🎯 Use the `as.factor()` function to change *at least one* character vector into factor class. Overwrite the original character vector with the factor class vector.

```{r eval=FALSE}
## ↔️ Change the class of your selected vector
# ⚡ Type this in your R Notebook, Fill in the blanks
__ <- _____( __ )
```

➤  Step 3

Check the Class

🎯 Verify that the class of the vector(s) have changes.

```{r eval=FALSE}
## ✅ Check the class of your selected vector
# ⚡ Type this in your R Notebook, Fill in the blanks
class( __ )
```

### Create as Factor Class

Create a new factor vector from scratch using the `factor()` function. (5 minutes)

➤  Step 4:

Make a Factor Vector

Here is an *example* to follow: 

```{r eval=FALSE}
# ⚡ Creating a character vector
tshirt_sizes_data <- c("M", "L", "S", "M", "L", "XL", "S")
# ⚡ Changing the character vector to a factor class with levels
tshirt_sizes <- factor(tshirt_sizes_data, levels = c("S", "M", "L", "XL"))
# ✅ Investigating your factor class with levels
print(tshirt_sizes)
```

🎯 Create your own factor vector called `social_media_platform`. Pick at least 4 social media / online networking platforms to create your vector. Element duplication is recommended. Rank the platforms using the `levels`. (5 minutes)

```{r eval=FALSE}
# ⚡ Type this in your R Notebook, Fill in the blanks
social_media <- c( __ , __ , __ , __ , __ , __ , __ , __ , __ , __ )
social_media_platform <- factor( __ , levels = c( __ , __ , __ , __ ) )
# ✅ Investigating your factor class with levels
print( social_media_platform )
```

# Story

## Reporter

### Confirm Objects

➤  Step 1

Confirm these objects exist in your Environment: `numeric`, `integer`, `logical`, `name`, `group_name`, `at_handle`, `item`, `activity`, `location`, `call_to_action`, `emotion`, `retro_game`, `social_media_platform`.


```{r eval=FALSE}
# ⚡  Copy and paste this code into your R Notebook
ls()
```

### Mad Lib Story

➤  Step 2

Copy and paste the Mad Lib Story code and read your Mad Lib Story!

```{r , eval=FALSE}
# ⚡  Copy and paste this code into your R Notebook
# Function to pick a random element from a vector
pick <- function(x) sample(x, 1)

# ---------- Story 1 ----------
activity_chosen <- pick(activity)

story1 <- paste(
  "After a long week,", pick(name), "unplugged by trying", activity_chosen,
  "with", pick(group_name), "at their favorite spot,", paste0(pick(location), "."),
  "Vibe rating:", paste0(pick(numeric), "."),
  "The", activity_chosen, "was going so well, but then a crucial missing", paste0(pick(item), " was discovered."),
  "Going with the flow was", ifelse(pick(logical), "reasonable.", "unreasonable."),
  "Reminding each other", paste0(pick(call_to_action), "."),
  "They made a plan for Episode", paste0(pick(integer), ","),
  "closing with a", paste0(pick(emotion), " mood.")
)

print(story1)
# ---------- Story 2 ----------
# Using Factor Level to pick 
selected_index <- sample(1:length(social_media_platform), 1)
selected_platform <- social_media_platform[selected_index]
outcome <- outcomes[selected_index]

story2 <- paste(
  "On a mission for a fashion comeback,", pick(name), "found", pick(item),
  "at", pick(location), "that screamed Y2K.",
  "They spent a total of", paste0(pick(numeric), " dollars"),
  "but they", ifelse(pick(logical), "remembered", "forgot"), "to get a receipt!",
  "After posting a 'before' shot on", paste0(pick(selected_platform), ","),
  "the post went viral", ifelse(pick(logical), "by midnight.", "by the next day."),
  "Their retro level was officially", paste0(pick(retro_game), " status,"),
  "and they immediately upcycled the item online for", paste0(pick(numeric), " dollars"),  
  "and sold in", pick(integer), "minutes.", 
  "The community response was so strong, they had this final message for their followers:", 
  paste0(pick(call_to_action), ".")
)
print(story2)
```


### Dataframe

➤  Step 2

Build a tiny table you can reference as “credits” under the stories.

🎯 Build a tiny table you can reference as “credits” under the stories.

```{r eval=FALSE}

# ⚡  Copy and paste this code into your R Notebook
people <- data.frame(
  member_name = member_name,
  role        = role,
  likes_story    = logical,
  dice_roll       = integer, 
  location = location
)

print(people)
```

🗣 Confirm that all group member's names and roles are listed correctly in the `people` data frame.

### Submission

➤  Step 3

Run All to confirm everything works top-to-bottom 

➤  Step 4

Upload the R Notebook on Canvas.

# Today you practiced

In this activity, your group practiced:

- Creating vectors of different types (numeric, integer, character, and logical).

- Combining vectors into a data frame.

- Collaborating as a team to complete a coding challenge.

Your work today previewed how R organizes and manages different data structures.

👉 Before leaving, make sure the reported submitted the R notebook on Canvas. This notebook should show all of the vectors and learning memos you and your team created during the activity.
