---
title: "Lesson 2"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
team: "Fall 2025"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactive scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Lesson 2"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
 
Or
title: "Lesson 2"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
team: "Fall 2025"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactive scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```


# 👋 Welcome back! 

In lesson 1 you learned how to run code and create simple objects. Lesson 2 moves one step further: you’ll meet *vectors* which are ordered collections of values that form the backbone of nearly everything in R.

To begin Lesson 2, follow these steps:

1. Open your course project for RStudio 
   
2. Create a new R script file (File > New File > R Script).

3. Type in the code provided in this document as you follow along with the video. Pause the video at anytime to answer assignment questions, dig deeper or add memo notes.

## Lesson Overview

In Part One you will learn how to:



- 📍 Run code from a script.

- 💪 Document tenacity in memos.

- 📊 Combine words with the `paste()` function.

In Part Two you will learn how to:

- 🛠 Build numeric, character, and logical vectors.

- 🔍 Check the structure, length, and data type of any object.

- ⚡ Explain how R coerces mixed‑type vectors.

- 📊 Subset vectors by position or logical tests.

- 📍 Summarize vectors with built‑in, *vectorised* functions.

- 📦 Remove objects you no longer need.



# Tenacity

Tenacity is the learning skill of continuing with serious, honest effort when the work gets uncomfortable—without spiraling emotionally or pretending the problem isn’t there. It shows up when you persist through obstacles (confusion, errors, frustration), revise your approach, and know when to seek help after genuine effort rather than quitting too early or grinding aimlessly (King, 2021; Baehr, 2022).

Tenacity is not “never give up.” It’s productive persistence.

## Why R is Challenging

R is not hard because it’s mysterious. It’s hard because it’s precise. Small mistakes (a missing parenthesis, a comma in the wrong place, a misspelled object name) produce errors, and those errors are not personal—they’re information.

Early in Module 1, you may have encounter friction in exactly the places beginners expect to be “simple”:

- Installing or updating R/RStudio

- Saving and submitting scripts and file to Canvas 

- Creating objects and then getting object not found errors

If you treat every error as evidence you “can’t code,” you will stall. Tenacity is what keeps you moving long enough for the patterns to become familiar (King, 2021).

## Two Ways of Getting Stuck

1) Giving up too early (irresolution).
This looks like skipping checks, continuing “halfway,” or abandoning the task the moment you feel lost (King, 2021). In R, that often becomes: running code you don’t understand, hoping it works, and stacking confusion.

2) Pushing too long (intransigence).
This looks like repeating the same approach for too long even though it isn’t working, refusing to change strategy, and refusing help (King, 2021). In R, that often becomes: re-running the same broken line 30 times.

Tenacity lives in the middle: persist, then recalibrate.


# Numeric Vectors

A **vector** is like a row in a spreadsheet: a single dimension of related values.  Use `c()` to *combine*, `:`  to create a simple sequence, or  `seq()`  for custom steps.  Please type the following into your R script and run each line.

## `c()`

```{r , eval=FALSE}
numbers   <- c(4, 8, 15, 16, 23, 42)          # numeric via c()

numbers3  <- c(1.5, 2.5, -3.5)                # numeric with decimals and negative
```

## `seq()`

Generating a sequence of numbers is useful for creating evenly spaced values. Use `seq()` to specify the start, end, and step size. Type the following into your R script and run each line.

```{r , eval=FALSE}
numbers5  <- seq(from = 1, to = 10, by = 2)   

steps     <- seq(from = 0, to = 1, by = 0.2)
```

👀 Check-in: What did `seq` do? Run "numbers3" in the Console to see the result.

## `:`

Please type the following into your R script and run.

```{r , eval=FALSE}
series    <- 1:10                             # shortcut 1,2,…,10
```

🚀 Explore and Play: What are other ways you can create numeric vectors? For example, try `rep()` to repeat values.

```{r}
# Create a memo note in your R script 🚀 Explore and Play. 
# This is where you could earn points for demonstrating learning skills.
## 1. Explain what you did in creating vectors, what your observations are, and 
# what you expect to happen. 

## 2. Demonstrate learning skill(s) in your memo
```

# Vector arithmetic

You can perform arithmetic operations on vectors. R will apply the operation element‑wise, meaning it will add, subtract, multiply, or divide corresponding elements in the vectors.

Recall that we created the objects: 

```{r , eval=FALSE}
numbers   <- c(4, 8, 15, 16, 23, 42)
steps     <- seq(from = 0, to = 1, by = 0.2)
```

🔍 What is the length of `numbers` and `steps`?

Let's do some vector arithmetic with these two vectors. Type the following into your R script and run each line.

```{r , eval=FALSE}
numbers - steps
```

That's numbers minus steps.

# ➤ Part I Assignment

Now it's your turn to practice creating and using vector objects. Follow the tasks below to complete part of the **technical skill practice assignment**.

1. Work through each task in order. Replace the ___ with your code.

2. Run each completed line to be sure no errors appear and objects show in the Environment.

## Task 1

Build two numeric vectors where both vectors are the same length.

1.  Create a vector called `sales_q1` with FOUR numbers of your choice.

2.  Create a vector called `sales_q2` with FOUR numbers of your choice.

## Task 2

Using vector arithmetic, create an object named `growth` that stores the difference between `sales_q2` and `sales_q1`.

🚀 Explore and Play: What happens if you try to subtract a vector of different lengths? 

```{r}
# Create a memo note in your R script if you did the 🚀 Explore and Play. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo

```

_Reminder: Now would be a great time to click save (or press Ctrl + S) to save your script._

✔️ Nice! Now, we'll dig a little deeper into other types of vectors.

# Character Vectors

Frequently used as plot labels, text data, or categories. Use quotes to create character strings. Type the following into your R script and run each line.

```{r , eval=FALSE}
chars     <- c("apple", "banana", "cherry")    # character vector

chars2    <- c("red", "yellow", "green")       # another character vector

chars3    <- c("A", "B", "C")                  # single letters as characters
```

👀 Check-in: What happens if you don't put quotations around character strings? Try it with `c(apple, banana, cherry)` and see what R does. What happens if you put the entire vector in quotes like this `c("apple, banana, cherry")`?  What happens if you quote numbers like `c("1", "2", "3")`? 

```{r}
# Create a memo note in your R script if you completed this 👀 Check-in. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo
```

# Missing Values

Sometimes data is missing or unknown. R uses `NA` to represent these missing values. You can create a vector with `NA` like this:

```{r , eval=FALSE}
z <- c(1, 2, NA, 4, 5)  # vector with a missing value
```

👀 Check-in: What happens if you run `z + 2`?  What happens if you type `0/0` or `Inf - Inf` in the console?

🔍 Look deeper: What is Inf? Type `?Inf` in the console. What happens?

_Note: `NaN` stands for "Not a Number" and is used when a calculation does not produce a valid number._

```{r}
# Create a memo note in your R script if you did the 👀 Check-in and 🔍 Look deeper. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo

```

# Logical Vectors

Logical vectors are used for conditions and comparisons. They contain TRUE or FALSE values.

```{r , eval=FALSE}
logicals  <- c(TRUE, FALSE, TRUE, NA)          # logical (TRUE/FALSE)

logicals2 <- c(1 > 2, 3 < 5, 4 == 4)           # logical comparisons
```

The logical operators are:

- `>` greater than

- `<` less than

- `==` equal to

- `>=` greater than or equal to

- `<=` less than or equal to

- `!=` not equal to

Additionally, if `logical` and `logical2` are logical expressions, then `logical & logical2` is their intersection (*and* operator), and `logical | logical2` is their union (*or* operator). We'll cover logical vectors in more detail in a later lesson.

👀 Check-in: Run each name (numbers, chars, …) in your environment by typing the object name. Note how R prints them. After running, peek at the Environment (Grid view) to see each vector shows its type & size.

```{r}
# Create a memo note in your R script if you did the 👀 Check-in. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo
```

# Indexing

Indexing is how you access specific elements in a vector. R uses 1-based indexing, meaning the first element is at position 1. You can use square brackets `[ ]` to index vectors.

Recall that we created: 

```{r , eval=FALSE}
z <- c(1, 2, NA, 4, 5)

numbers3  <- c(1.5, 2.5, -3.5)
```

## Sub-setting Vectors

🎯 **Goal**  Pull out the pieces you need.  Use `[ ]` with positions or a logical test. 

🧠 Remember: positions start at 1 in R (not 0).

Type the following into your R script and run each line. 

```{r , eval=FALSE}
z[1]       # first element by position

z[3]       # third element by position

z[2:4]     # last two elements by position 

z[-c(1, 5)] # all but first and last elements
```

🛠 Break Things! What happens if you try to index with a number larger than the vector length, like `z[6]`? ⚡ Explain any error message you see.

🧠 Challenge: Extract the last two fruits from *chars*.

## Logical Indexing

Type the following into your R script and run.

```{r , eval=FALSE}
numbers3[numbers3 > 0] # keep only positives
```

🔍 Look deeper: What happens if you run`y <- z[(!is.na(z))]`?  Hint: The function `is.na()` checks for missing values (NA) in the vector. The `!` operator negates the logical vector. Print `y`. 👀 What happened?⚡ Explain

```{r}
# Create a memo note in your R script if you did 🛠 Break Things and 🧠 Challenge 
# Or if you did the 🔍 Look deeper, and 👀 What happened?. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did in Indexing, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo
```

# Inspecting Vectors

Before using data, we know confirm *what* it is.  `str()` gives a compact summary, `length()` counts elements, and `class()` reveals type. Please type the following into your R script and run each line.
 
## Length

Type the following into your R script and run.

```{r , eval=FALSE}
length(chars)
```

🔎 Explore the length of objects in the Environment pane.

## Class

`class(logicals)` 

🔎 Explore the class of objects in the Environment pane.

**Some types of classes in R include:**

- **numeric**: numbers (e.g., `1`, `3.14`)

- **character**: text strings (e.g., `"Hello World", "R is fun"`)

- **logical**: `TRUE` or `FALSE` values

- **integer**: whole numbers (e.g., `1L`, `2L`)

- **factor**: categorical data (e.g., `"red"`, `"blue"`) is
  a special type of character vector used for categorical data.

- **NULL**: represents an empty object (e.g., `NULL`)

- **complex**: complex numbers (e.g., `1 + 2i`)

- **Index**: a special type of vector used for indexing (e.g., `1:10`)

- **list**: a collection of objects (e.g., `list(a = 1, b = "text")`)

- **matrix**: a two-dimensional array (e.g., `matrix(1:6, nrow = 2)`)

- **array**: a multi-dimensional array (e.g., `array(1:12, dim = c(2, 3, 2))`)

- **data.frame**: a table-like structure (e.g., `data.frame(name = c("Alice", "Bob"), age = c(25, 30))`)

## Structure

Type the following into your R script and run.

```{r , eval=FALSE}
str(numbers)    # shows type = num, length = 6
```

🔍 Look deeper: What other functions can you use to inspect objects? 

```{r}
# Create a memo note in your R script if you did 🔍 Look deeper. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did to discover more functions, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo
```

# ➤ Part II Assignment

Now it's your turn to practice creating and inspecting vector objects. Follow the tasks below to complete part of the **technical skill practice assignment**.

1. Work through each task in order.

2. Run each completed line to be sure no errors appear and objects show in the Environment.

## Task 3: Inspect

1. Use `class()` and `str()` to inspect the `object` growth.

2. Write a one-line comment describing what `str()` reveals.

## Task 4: Subset

1. Create a new vector `positive_growth` that contains only values in growth that are greater than zero. (hint: see logical vectors)

2. Create a new vector `last_two` that contains the last TWO elements of `sales_q2` (use numeric positions, not names).

✔️ Nice! Next, lets try to break some rules and learn more functions.

_Reminder: Now would be a great time to click save (or press Ctrl + S) to save your script._

# Data Coercion

⚠️ **Key Idea**  A vector can only hold *one* type. Mixing types triggers *coercion* to the most flexible type (logical → numeric → character).

What happens if you mix numbers and characters?

```{r , eval=FALSE}
mixed <- c("candy", 4, "book", 92)

class(mixed)            # character: all elements coerced to character
```

The `as.numeric()` function tries to convert characters to numbers. If it can't, it returns `NA` (not available).

```{r , eval=FALSE}
as.numeric(mixed)            # character → numeric: text converts to NA

as.numeric(c("1", "2", "3")) # character → numeric: text converts to numbers
```

🔍 Look deeper: What are other ways to convert class types? 

```{r}
# Create a memo note in your R script if you did 🔍 Look deeper. 
# This is where you could earn points for demonstrating learning skills.

## 1. Explain what you did to discover the functions, what your observations are, and what you expect 
# to happen. 

## 2. Demonstrate learning skill(s) in your memo
```

# Vectorized Functions

⚡ **Why it’s cool**  Functions like `sum(),` `mean(),` and `sd()` automatically loop over every element—no for-loop needed. This is called *vectorization*.

🔍 Look deeper: What is a for-loop? 

Recall we created `series <- 1:10`

Type the following R code into your script and run. 🗣 Comment about the output.

```{r, eval=FALSE}
sum(series)

mean(series)

sd(series) 
```

# Cleaning Up

🧹 **When to sweep**  Large projects can crowd the Environment. Use `rm()` to remove objects you no longer need and free memory. Type the following into your R script and run.

```{r , eval=FALSE}
rm(mixed)      # mixed disappears from the Environment pane
```

# Dataframe function

Vectors form the foundation for data frames. 

📊  **What & Why**  A data frame is like a spreadsheet: a table of rows and columns. Use `data.frame()` to create one from vectors of the same length. Each vector becomes a column. 

```{r , eval=FALSE}
column_1 <- c(1, 2, 3)

column_2 <- c("A", "B", "C")

column_3 <- c(TRUE, FALSE, TRUE)

data_frame <- data.frame(column_1, column_2, column_3)
```

# ➤ Part III Assignment

Now it's your turn to practice creating and inspecting vector objects. Follow the tasks below to complete part of the **technical skill practice assignment**.

1. Work through each task in order.

2. Run each completed line to be sure no errors appear and objects show in the Environment.

## Task 5: Vectorised summaries

1. Calculate the average of `sales_q2`; save as `avg_q2`.

2. Calculate the total of `sales_q1`; save as `total_q1`.

## Task 6: Mini data frame 

1. Create a character vector `months` that names four months (e.g., "Jan").

2. Combine months, `sales_q1`, `sales_q2` and `growth` into a data frame called `sales`.

3. Check the structure of sales with `str()`. Explain what it shows.

## Task 7 Reflection

🗣 In one sentence, explain why numeric values may be coerced to character if you mix them inside a single vector.

# Save and Upload

1. You will be submitting both the R script and the workspace file. The workspace file saves all the objects in your environment that you created in this lesson. You can save the workspace by running the following command in your R script:

```{r , eval=FALSE}
save.image("Assignment2_Workspace.RData")
```

Or you can click the "Save Workspace" button in the Environment pane. 

💾 **Always save the R documents before closing. If you don't, you will lose your work.**

2. Find the assignment in this week's module in Canvas and upload **both** the R script and the workspace file.

# Today you practiced

- Building numeric, character, and logical vectors.

- Inspecting structure, length, and data type.

- Understanding and witnessing type coercion.

- Sub-setting vectors with positions and logical tests.

- Applying vectorised summary functions.

- Tidying your workspace with `rm()`.

- Creating a data frame from vectors.

🚀 Great progress! In the next lesson, we'll learn more about functions. 

