---
title: "Lesson 7 grading"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=FALSE, include=FALSE}
title: "Lesson 7 grading"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    citation_package: natbib
    fig_caption: true
    df_print: kable # Data frame printing
    includes:
      in_header: ../assets/header.tex
    latex_engine: xelatex  # Use xelatex to support fontspec
fontsize: 12pt
geometry: margin=1in
mainfont: "Garamond" # Sets the font of the entire document
sansfont: "Gotham-Book.otf" # Set sans-serif font to Gotham Book
monofont: "Courier New" # Set monospace font to Courier New
documentclass: scrreprt
linkcolor: boisestateblue # Customizes the color of hyperlinks
urlcolor: magenta # Customizes the color of URLs
citecolor: black # Customizes the color of citations
bibliography: references.bib # Bibliography file
biblio-style: apalike                 # ⟵ natbib needs a .bst style
natbiboptions: "round,authoryear"     # round brackets, Author (Year)
 
Or
title: "Lesson 7 grading"
author: "by Lorraine Gaudio"
date:   "`r paste('Lesson generated on', format(Sys.Date(), '%B %d, %Y'))`"
team: "Fall 2025"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: true # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    code_folding: show # Enables you to include R code but have it hidden by default. (Show hide button)
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
```




# ➤ Assignment

Grading Notes

* Accept any valid target_year and any two valid region names that appear in unique(gapminder$region).

* For Tasks 4–7, counts and countries will vary by the student’s year/region choices; grade correctness of logic and syntax ([], logical operators, %in%, negative column indices, which()), not the exact numbers.

* Explanations/memos should be short, accurate, and use correct vocabulary (logical vector, rows/cols slots, membership, AND/OR).

**Campus NGO Health Pilot** 🌍💉

Imagine your nonprofit received 100 vaccine coolers and one charter flight. You must assemble a shortlist of countries for a health-equity pilot within the next few weeks. Use dslabs::gapminder to (a) focus on specific regions, (b) filter by need indicators, and (c) apply a simple feasibility window based on population size.

## Task 0

🚀 Setup & Orientation

Load the data, read the help page, and become familiar with the `gapminder` dataset.

```{r}
# install.packages("dslabs")  # uncomment if first time

# TODO 0-A: Load the package and data
library(dslabs)  # use library() to load the package
data("gapminder")   # use data() to load the dataset "gapminder"

?gapminder   # pull up gapminder in the help window
```


```{r}
colnames(gapminder)      # use colnames() to list names of the columns in gapminder
```

```{r}
names(gapminder)      # use names() to list names of the columns in gapminder 
```

✍️ LEARNING SKILL MEMO (1–2 sentences): Compare functions `colnames()` and `names()` output.

`colnames()` and `names()` both return the column names for a data frame. `names()` is more general (works for lists, too); for data frames both produce the same column name vector.

```{r}
# TODO 2-C: How many rows are in gapminder?
nrow(gapminder)
```

## Task 1 

📅 Pick a Year

Choose ONE specific year present in the data and store it:

1. Use unique() to print the available `years` in `gapminder`.

```{r}
# TODO 1‑A: list years
unique(gapminder$year)     # list available years
```

```{r}
# TODO 1‑A: Select ONE year
target_year <- 2010
```

✍️ MEMO (1–2 sentences): “I chose target_year = 2010 because it’s recent enough to reflect current patterns and is present in the dataset.”

## Task 2 

🎯 Focus the Pool (`year` `&` `region` with `%in%`)

First, find the exact region names; then keep only rows from your chosen year and your two target regions.

1. Identify the `unique` `region` options in `gapminder`.

```{r}
# TODO 2-A: Show all region options
unique(gapminder$region)     
```

*Use this to select two regions. Copy exact names.*

2. Create `focus_pool` that keeps rows from your `target_year` AND restricts to two regions you want to prioritize. Use unique(`gapminder$region`) above to copy region names exactly.

```{r}
# TODO 2-B: Create focus_pool using year AND region membership
focus_pool <- gapminder[gapminder$year == target_year & gapminder$region %in% c("Eastern Asia", "Southern Asia"),]
```


```{r}
# TODO 2-C: How many rows are in focus_pool?
nrow(focus_pool)
```

Compare the number of rows in `gapminder` with the number of rows in `focus_pool`

```{r}
# TODO 2-D: Preview a few informative columns
head(focus_pool[, c("country", "region",
                    "life_expectancy", "infant_mortality",
                    "fertility", "population")])
```

✍️ MEMO (1–2 sentences): Describe what `%in%` and `&` are doing here, and why you used `$` with column names.

EXPLANATION: `%in%` keeps rows where region matches any value in the set provided. `&` requires both conditions to be `TRUE` (the row is from `target_year` and the region is one of the two). `$` is used to refer to a specific column inside the data frame.

## Task 3

🗂 Negative Column Indices (Drop Columns by Position)

Create a simpler briefing table `brief_sheet` by dropping exactly two columns by position (not by name). Choose any two you don’t need to discuss need.

```{r}
brief_sheet <- focus_pool[, -c(7, 8)]
head(brief_sheet)
```

✍️ MEMO: Explain how negative indices work for column selection.

EXPLANATION: Negative indices remove columns at those positions. The remaining columns stay in the same order.

## Task 4

🔢 Quick Check with Inequality

How many countries in `focus_pool` have `life_expectancy` >= 70? 60?

```{r}
# Count countries with life expectancy ≥ 70
sum(focus_pool$life_expectancy >= 70)

# (Optional extra from prompt: also ≥ 60)
sum(focus_pool$life_expectancy >= 60)
```

✍️ MEMO (1 sentence): What value is `sum()` counting in this context?

EXPLANATION: sum(logical_vector) counts the number of TRUE values.

## Task 5

🧪 Need Screen (Combine Conditions with `&`)

1. Build `needs_screen` (keep all columns) using these **AND** conditions:

`life_expectancy <= a` where `a` is an appropriate number for your data set. 

`infant_mortality >= b` where `b` is an appropriate number for your data set. 

`fertility >= c` where `c` is an appropriate number for your data set. 

```{r , eval=FALSE}
# TODO 5-A: Create needs_screen using three AND-ed conditions
needs_screen <- focus_pool[
  focus_pool$life_expectancy <= 70 &
    focus_pool$infant_mortality >= 35 &
    focus_pool$fertility >= 2,
]
```

🛠 Break Things! Learning Skills challenge: What would happen if you used `|` instead of `&` here?

We use `&` because we only want rows meeting all three criteria simultaneously. Using `|` would include rows that satisfy only one condition, which is not the goal.

2. Print the number of rows and show a view of only key columns

```{r , eval=FALSE}
# TODO 5-B: How many rows survived? Show a compact view of key columns.
nrow(needs_screen)
```

```{r}
# TODO 5-B: Show a compact view of key columns.
needs_screen[, c("country", "region",
                 "life_expectancy", "infant_mortality",
                 "fertility", "population")]
```

## Task 6

🔁 `|` vs `%in%` (Same Result, Different Syntax)

1. Show that filtering the same two regions with `|` is equivalent to using `%in%`.

```{r}
# TODO 6-A: 
## Recreate the two-region filter using | (OR)
A_or_B <- gapminder[
  gapminder$year == target_year &
    (gapminder$region == "Eastern Asia" |
     gapminder$region == "Southern Asia"),
]

## Recreate the two-region filter using %in%
in_set <- gapminder[
  gapminder$year == target_year &
    gapminder$region %in% c("Eastern Asia", "Southern Asia"),
]
```

2. Do both approaches return the same number of rows?

```{r , eval=FALSE}
# TODO 6-B:
nrow(A_or_B) == nrow(in_set)
```
 
✍️ MEMO (1–2 sentences): Briefly compare `|` and `%in%`. When is `%in%` preferable?

EXPLANATION: `|` chains individual equality tests; `%in%` checks membership in a set and is shorter and less error-prone when matching multiple categories.

## Task 6

📍 Feasibility with `which()` (Population Window)

From `needs_screen`, keep countries with population between 2 million and 50 million (inclusive). Use `which()` to convert the logical test to row positions, then subset.

```{r , eval=FALSE}
# TODO 7-A: Build an index of matching rows with which()
idx <- which(needs_screen$population >= 2e6 &
             needs_screen$population <= 5e7)
```


```{r , eval=FALSE}
# TODO 7-B: Subset rows using idx to create shortlist
shortlist <- needs_screen[idx, ]
```

How many countries are in shortlist? Which regions appear in shortlist?

```{r , eval=FALSE}
# TODO 7-C: Display a compact table of shortlist (choose a few columns)
shortlist[, c("country", "region",
              "life_expectancy", "infant_mortality",
              "fertility", "population")]

```

✍️ MEMO  (1–2 sentences): What does `which()` return, and why might it be useful to store these positions?
 
EXPLANATION: `which()` returns the numeric positions of the `TRUE` rows. Saving indices is helpful when you want to reuse the same row selection across multiple views or operations.

