---
title: "Chapter 11: dplyr subsetting"
author: "Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Spring 2026"
output: 
  html_document: # To create an HTML document from R Markdown
    toc: false # Table of contents (TOC)
    toc_depth: 1 #(meaning that level 1, 2, and 3 headers will be included in the table of contents
    toc_float: # Float the table of contents to the left of the main document
      collapsed: false # Collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level
      smooth_scroll: true # controls whether page scrolls are animated when TOC items are navigated to via mouse clicks.
    number_sections: true # Numbering starts with "#" (H1). Without H1 headers, the H2 headers ("##") will be numbered with 0.1, 0.2, and so on.
    css: ../assets/styles.css # This is the name of the CSS file to style the HTML document with Boise State Brand. The CSS file must be in the same directory as the R Markdown file.
    fig_caption: true #Whether figures are rendered with captions.
    df_print: paged # Printing data frames with interactivne scrolling
    includes:
      in_header: ../assets/header.html
      after_body: ../assets/footer.html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In Chapter 10, you learned about subsetting, extraction, and insertion in data frames using base R functions. This Chapter, you will learn easier subsetting with `dplyr` (`select()` & `filter()`). The `dplyr` package simplifies column and row extraction, making code shorter and easier to read than base-R brackets. Mastering `select()`, `filter()`, and the pipe `%>%` will speed up every day tasks.

To begin Chapter 11, follow these steps:

1. Open your course project for RStudio 
   
2. Create a new file. Today, let's try 📝 "Quarto Document"  (File > New File > Quarto Document).

3. Type in the code provided in this document as you follow along with the video. Pause the video at anytime to answer assignment questions, dig deeper or add memo notes.

**Lesson Overview**

By the end of Chapter 11 you will be able to:

1. ⚡ Remember – List what `select()`, `filter()`, and `%>%` do.

2. 🔍 Understand – Explain how `select()` (columns) differs from `filter()` (rows).

3. 🛠 Apply – Use `select()` and `filter()` to subset a data frame.

4. 📊 Analyze – Chain multiple verbs with `%>%` to answer a real question.

5. 🧠️ Evaluate – Decide when `dplyr` is preferable to `base-R` brackets.

Keep these goals in mind as you move through each section.

# 📦 Packages

🧠 Recall that we need to download packages **once**. If needed, install the required packages we will use in this lesson. After installation, comment `#` out the `install.packages` code. 

```{r , eval=FALSE}
#install.packages("dslabs")
install.packages("dplyr")
```

After installing the package, you need to load it into your R session using the `library()` function. Each time you open R, you will need to "library" the packages. This makes the functions in the package available for use. Type the following code in a new code chunk and run.

💡 Pro tip: You can load multiple packages in one `library()` call by separating package names with commas.

```{r , eval=FALSE}
library(dslabs, dplyr)
```


# The Pipe (%>%)

🎯  The pipe operator passes the result of the left side into the first argument of the right side, reading like "and then". It cuts repetition and keeps code in logical order.

```{r , eval=FALSE}
# ⚡ Test
mtcars %>%        # start with data
  head()          # ▶ shows first six rows
```

🔗 Link Concepts: What other ways have you learned to write code that will view the first six rows of a data frame? Write this code and a memo to demonstrate learning skills.

# Selecting Columns

🎯 The GOAL of `select()` is to keep or drop specific **columns**.

📜 The SYNTAX is `select(data, col1, col2, …)`  

## Positive Selection

🎯 Goal: keep specific columns by name.

```{r , eval=FALSE}
# ⚡ Test
select(mtcars, mpg, cyl, hp)    # keep three columns
```

## Negative Selection

🎯 Goal: remove columns with a leading `-`.

```{r , eval=FALSE}
select(mtcars, -gear)   # drop one column
```

## Piping with Select

🎯 Goal: Let's add a pipe!

```{r , eval=FALSE}
mtcars %>% 
  select(mpg, cyl, hp)                  # easier to read
```

## Subset with Select
                
We'll mix it up by using a different data set, `gapminder` from `dslabs`. This data set contains information about countries over time.

```{r , eval=FALSE}
?gapminder
```

🎯 Goal: Create a copy of the `gapminder` dataset that contains only specific columns!

```{r , eval=FALSE}
# ⚡ Test:
I_Demo <- gapminder %>% 
  select(country, year, infant_mortality, continent)
```

```{r, eval=FALSE}
# ✅  Verify
head(I_Demo)
```

🗣 Explain: A new dataset called `I_Demo` was created from the `gapminder` dataset that only contains the columns `country`, `year`, `infant_mortality`, and `continent`. 

✍️ Practice: fill in the blank prompt

You try to design your own subsetting operation.

🎯 Goal: Write an example that selects only the columns you choose. 

```{r , eval=FALSE}
# ⚡ Test: 
Your_Choice <- gapminder %>%
  select(____, ____, ____)   # choose columns

# ✅  Verify

```

🗣 Explain: What columns did you choose and why? Create a memo note, demonstrate learning skill(s) used.

# Filtering Rows

🎯 The GOAL of `filter()` is to keep or drop specific **rows** based on conditions.

📜 The SYNTAX is `filter(data, condition1, condition2, …)`

## Single Condition

🎯 `filter()` uses logical operators!

```{r , eval=FALSE}
# ⚡ Test:
filter(mtcars, mpg >= 25)               # high‑mileage cars
```

## Piping with Filter

We'll mix it up by using a different data set, `murders` from `dslabs`. This data set contains information about gun murders in the US.

```{r , eval=FALSE}
?murders
```

🎯 Goal: Create a copy of the `murders` dataset that excludes Washington state.

💬 Prediction: 50 states

```{r , eval=FALSE}
# ⚡ Test:
nrow(murders)  
```

🗣 Explain: ... Wait, what? ... ❓ Question: What are these 51 states?

```{r , eval=FALSE}
#⚡ Test:
unique(murders$state)  # 51 states
```

🗣 Explain: Ohhhhh! "District of Columbia" is included in the dataset.

Back to the 🎯 Goal: Create a copy of the `murders` dataset that excludes Washington state.

```{r , eval=FALSE}
# ⚡ Test:
No_Washington <- murders %>% 
  filter(state != "Washington")
```

```{r, eval=FALSE}
# ✅  Verify
nrow(No_Washington)  # 50 states
```
🗣 Explain: A new dataset called `No_Washington` was created from the `murders` dataset that excludes Washington state.

🧐 Review: What is the difference between `==` and `!=`? 

🛠 Break Things! What happens when you use `=` instead of `==`? or `=!` instead of `!=`? Create a memo note, demonstrate learning skill(s) used.

## Multiple Conditions with AND

📌 In `filter()`, commas = AND (`&`).

🎯 Goal: Let's filter with two conditions and while we are at it, let's add a pipe! 

❓ Question: How many cars have *both* mpg ≥ 25 *and* 4 cylinders? 

```{r , eval=FALSE}
# ⚡ Test:
mtcars %>%
  filter(mpg >= 25, cyl == 4)           # two conditions
```

✅  Verify: How many cars have *both* mpg ≥ 25 *and* 4 cylinders? What code would you use to verify the success of this filter? Create a memo note, demonstrate learning skill(s) used by writing and running the code.


## Multiple Conditions with OR

📌 In `dplyr::filter()`, the OR operator is the same as base R `|` or `%in%`.

### |

🎯 Goal: Let's filter with two conditions using OR (`|`).

❓ Question: How many cars have *either* mpg ≥ 25 *or* 4 cylinders?

```{r , eval=FALSE}
# ⚡ Test:
mtcars %>%
  filter(mpg >= 25 | cyl == 4)           # two conditions
```

✅  Verify: How many cars have *either* mpg ≥ 25 *or* 4 cylinders? What code would you use to verify the success of this filter? Create a memo note, demonstrate learning skill(s) used by writing and running the code.

### %in%

🎯 Goal:  The `%in%` operator is used to check if a value is in a set. It’s like "is this value one of these?"

```{r , eval=FALSE}
# ⚡ Test
filter(mtcars, cyl %in% c(4, 6))        # 4 or 6 cylinders
```

# filter() AND select()

🎯 The GOAL of combining `filter()` and `select()` is to **subset** a data frame by both rows and columns.

💪 Best practice: filter rows first, then select columns.

📜 The SYNTAX is `data %>% filter(condition) %>% select(col1, col2, …)`

🎯 We'll mix it up by using a different data set, `us_contagious_diseases` from `dslabs`. This data set contains information about contagious diseases in the US.

```{r , eval=FALSE}
?us_contagious_diseases
```

```{r , eval=FALSE} 
#⚡ Test
us_contagious_diseases %>% 
  filter(state == "Alaska", year >= 1980) %>%   # rows
  select(year, disease, count)                  # columns
```

🚀 Explore and Play: What are the unique diseases in this Alaska subset? Use skills you developed in previous lessons to explore this data set. Compare filter and select with subsetting with brackets `[ ]`. Create a memo note, demonstrate learning skill(s) used.

# One Application

🧐 Let's say you have the research question, ❓"Which automatic cars (am == 0) in mtcars have horsepower > 150? 🎯 Goal: Your report should show only hp and mpg.

```{r , eval=FALSE}
#⚡ Test
mtcars %>%
  filter(am == 0, hp > 150) %>%
  select(hp, mpg) %>%
  head()
```

🗣 Explain:

1. `filter()` keeps matching rows.

2. `select()` trims to needed columns.

3. `head()` previews results. Bonus: Add row count for exploration.

# Duplicates & Sampling

Sometimes, you may need to remove duplicate rows or sample a subset of rows. `unique()` removes duplicate rows; `slice_sample()` picks random rows. 👀 Type `?slice_sample()` to learn more about this `dplyr` function.

🎯 Goal: We'll mix it up by using a different data set, `movielens` from `dslabs`. This data set contains information about movies and ratings.

```{r , eval=FALSE}
?movielens
```


```{r , eval=FALSE}
nrow(movielens)   # how many rows at start?
```

```{r , eval=FALSE}
# ⚡ Test
Movie_Night <- movielens %>%
  filter(rating >= 4) %>%        # keep high ratings
  select(title, rating) %>%      # keep title and rating
  unique()                       # remove duplicates
```

```{r , eval=FALSE}
# ✅  Verify
nrow(Movie_Night)  # how many rows after filter, select, and unique?
```

🗣 Explain: Filter kept only the rows with ratings 4 or higher. Select kept only the title and rating columns. Unique removed duplicate rows, so each movie title appears only once.

❓ Question: What movie should I watch on movie night? 

🎯 Goal: Let's randomly sample 10 movies with ratings four or higher.

```{r , eval=FALSE}
set.seed(420)
# ⚡ Test
Movie_Night %>%
  slice_sample(n=10)  # random sample of 10 rows
```

🗣 Comment: What movie would you choose?

🚀 Explore and Play: What would happen if we didn't use unique()? Why might unique be useful for our question? Create a memo note, demonstrate learning skill(s) used.

# Extra Practice

✍️ Practice: Demonstrate learning skill(s) by completing this bonus practice. Fill the blanks. Use the `gapminder` dataset.

```{r , eval=FALSE}
___ %>%
   filter(life_expectancy >= quantile(life_expectancy, ___)) %>%
   select(country, ___, life_expectancy) -> Top_LifeExp
View(Top_LifeExp)
```

🗣 Explain: What is the `-> Top_LifeExp` do in the code above?

❓ Question: What is the highest life expectancy in this subset?

❓ Question: How many countries are in the `Top_LifeExp` data frame?

❓ Question: What is the average life expectancy in this subset?


# Summary


# Chapter Terms


# 📝 Practice Space


Replace each ____ placeholder (and any TODO comments) with working code or a short written answer. Run each section; be sure the requested objects appear in the Environment. When finished, save **BOTH** your script (which should include your lesson notes and completed assignment) and your `.RData` workspace and upload to Canvas Assignment 8 in Module 8.

## Task 0

📦 Make sure `dplyr` and `dslabs` packages are attached so their functions / datasets load.

## Task 1

Column Selection

📝 Create **`C_Overload`** that keeps only year and `carbon_emissions` from the `temp_carbon` data frame. Use `select()` with or without `%>%`.

```{r, eval=FALSE}
____ <- _____
```

## Task 2

Single-Column Extraction

🌟 Pipe the `stars` dataset in `dslabs` through `select()` to keep only the `star` column. Name the result **`star_names`**.

```{r, eval=FALSE}
____ <- _____
```

## Task 3

Row Filtering

⚠️ From `murders`, remove every row where state equals `"Florida"`. Name the object **`No_Florida`** (any method is fine).

```{r, eval=FALSE}
____ <- _____
```

## Task 4

Multiple Conditions

🚗 From `mtcars`, use `filter()` + `%>%`, pull cars in mtcars that are manual, have 8 cylinders, and have `qsec < 15`. Store as **`Fast_n_Furious`**. 

```{r, eval=FALSE}
summary(mtcars$qsec)  # check qsec range
```

Is `qsec < 15` within range? 

```{r, eval=FALSE}
____ <- _____
```

## Task 6

OR Logic + Column Slice

In 🚗 `mtcars`, show only rows where `disp > 250` OR `disp < 160` and display just the first three columns.

```{r, eval=FALSE}
____ <- _____
```


## Task 7

Reflection

📝 Write a few sentences comparing dplyr verbs over base-R brackets (from lesson 6 & 7) for subsetting.

# Save and Upload

0. Self-check checklist:

For the assignment protion, your environment should contain the following items:

- `C_Overload` exists and has only `year` and `carbon_emissions`.

- `star_names` exists and has one column named `star`.

- `No_Florida` has 0 rows with `state == 'Florida'`.

- `Fast_n_Furious` with rows that only contain manual, 8 cylinders, and where `qsec < 15`.

- **All code in your Quarto document runs top-to-bottom without errors.**

1. You will be submitting **both** the Quarto document and the workspace file. The workspace file saves all the objects in your environment that you created in this lesson. You can save the workspace by running the following command in a code chunk of the Quarto Document document:

```{r , eval=FALSE}
save.image("Assignment8_Workspace.RData")
```

Or you can click the "Save Workspace" button in the Environment pane. 

💾 **Always save the R documents before closing.**


2. Find the assignment in this week's module in Canvas and upload **both** the `.qmd` and the workspace file (`RData`).

# Today you practiced:

- Installed / loaded packages for extra tools.

- Used the pipe `%>%` to chain steps left-to-right.

- Selected columns with `select()` (including negative selection).

- Filtered rows with `filter()` using logical tests and `%in%`.

- Combined verbs to answer data questions concisely.

🎯 Great job! Keep practicing piping and you’ll soon write clear, readable R code!




# Notes

## Learning Skill


LS: 
Core
Review course content
Advanced
Curiosity Skill Highlight



Citing resources

- (Curiosity highlight): “data scavenger hunt” + citation + one curiosity question answered with a quick code experiment (not a paragraph of vibes).


- Courage: Students can share imperfect work in a discussion/lab setting and revise based on feedback.

Maybe Module 6: Autonomy: 
- “trying alternatives” (base R vs dplyr; indexing vs verbs) and documenting what the output shows.
- Students can manage their workflow independently: organized folders, naming, saving, and submitting correct artifacts without resource rescue.
- Students can set boundaries on AI use consistent with instructions and explain their choice in a memo (what they delegated vs what they did themselves).
- Students can select a method and justify it (e.g., base R indexing vs dplyr select/filter) using output evidence. “I chose this approach because…” memo

Maybe Module 7: Open-mindedness
- Students can implement two approaches to the same task (e.g., base R vs dplyr) and compare outputs and readability.
- Students can revise their approach when evidence shows a better method (not because it’s trendy).
- Students can evaluate tool fit (including AI systems) for a specific goal and explain tradeoffs.





