---
title: "Research Journal Report (1)"
output: html_document
date: "2025-09-29"
---

#set up!

##setting directiory
```{r}
setwd("C:/Users/gretc/OneDrive/R155/data")
getwd()
```
##setup chunk
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo= TRUE, message = FALSE)
#include=TRUE shows the code/output from the final document.
#echo=TRUE  show code in the knitted report
```
##library packages
```{r}
library(readxl)
#now i can upload my excel dataset
```

#title & overview

This report is my signature assignment! It will hopefully show my ability to combine R code & explain my process clearly

I chose a dataset with stats pulled from my own Pokemon Go game. It includes names, types, combat power (CP), hit points (HP), attacks, and traits like shiny or lucky. I picked it because it’s personal, fun, and has a good mix of numeric (CP, HP, weight, height), categorical (type), and binary (shiny, lucky) data for practicing R.


#dataset citation & license


Bostain (me), 2025,*Pokémon Go Personal Dataset*, self-collected, unpublished, no external license

This dataset was created by recording attributes of Pokémon from my own Pokémon Go collection, including CP, HP, attacks, type, weight, height, and whether each Pokémon is shiny or lucky. The data was entered manually into Excel and exported into R

```{r}
pogo_data <- read_excel("datasetpogo.xlsx")
```

#data overview

Each *row* represents a single Pokémon from my collection.  
Each *column* represents an attribute, such as name, type, CP, HP, attacks, weight, height, and whether the Pokémon is shiny or lucky
```{r}
head(pogo_data)
#a little preview of the first 6 rows
head(pogo_data, 10)
#a little preview of the first 10 rows
```

##experimenting
```{r}
colnames(pogo_data)
#these are the attributes i chose, or the columns
anyNA(pogo_data)
#interesting, i guess there is some NA present in my dataset
#i didnt intend for that, so i wonder what it is
summary(pogo_data)
#nice, it seemed to summarize each column
#i see some random empty columns though, will need to cut
tail(pogo_data)
#like the head(), but is a preview of the last 6 rows
tail(pogo_data, 10)
na_counts <- colSums(is.na(pogo_data))
na_counts
#i think all my main pokemon attributes are fine, so i want
#to clean this up... will recruit boisestate.ai for this
```
boisestate.ai suggested a package for this! its called janitor
```{r}
#install.packages(janitor)
#Error: object 'janitor' not found
#i forgot the packages have to have "
#install.packages("janitor")
library(janitor)
```
trying the code
```{r}
janitor::remove_empty(pogo_data)
#cool the pop-up looked clean but i want the change to be 
#permanent
summary(pogo_data)
#no idea! looks the same to me
#ah!!! it doesnt remember.. its not like ai!
#i have to make a new, clean dataset
clean_pogo_data <- janitor::remove_empty(pogo_data)
```
time to see if that worked...
```{r}
summary(clean_pogo_data)
anyNA(clean_pogo_data)
#popped FALSE! yippeee!
```
###what does :: do?
it allows you to use a function of a package, without loading the whole thing
###what does $ mean
pulls a column (attribute) by name
####example
```{r}
clean_pogo_data$pokemon
#lists the pokemon column
clean_pogo_data$type
#lists the type column
```
